Inversions Can Accumulate Balanced Sexual Antagonism: Evidence from Simulations and Drosophila Experiments

Version 1

Abstract

Abstract

Chromosomal inversion polymorphisms can be common, but the causes of their persistence are often unclear. We propose a model for the maintenance of inversion polymorphism, which requires that some variants contribute antagonistically to two phenotypes, one of which has negative frequency-dependent fitness. These conditions yield a form of frequency-dependent disruptive selection, favoring two predominant haplotypes segregating alleles that favor opposing antagonistic phenotypes. An inversion associated with one haplotype can reduce the fitness load incurred by generating recombinant offspring, reinforcing its linkage to the haplotype and enabling both haplotypes to accumulate more antagonistic variants than expected otherwise. We develop and apply a forward simulator to examine these dynamics under a tradeoff between survival and male display. These simulations indeed generate inversion-associated haplotypes with opposing sex-specific fitness effects. Antagonism strengthens with time, and can ultimately yield karyotypes at surprisingly predictable frequencies, with striking genotype frequency differences between sexes and between developmental stages. To test whether this model may contribute to well-studied yet enigmatic inversion polymorphisms in Drosophila melanogaster, we track inversion frequencies in laboratory crosses to test whether they influence male reproductive success or survival. We find that two of the four tested inversions show significant evidence for the tradeoff examined, with In(3R)K favoring survival and In(3L)Ok favoring male reproduction. Additionally, all inversions show survival differences between sexes, and paternal success depends on maternal genotype. Based on this work, we expect that balancing selection on antagonistically pleiotropic traits may provide a significant and underappreciated contribution to the maintenance of natural inversion polymorphism.

Introduction

The Enigmatic Prevalence of Inversion Polymorphism

Inversions, DNA segments in reversed sequence order relative to the ancestor, have been a focus of population genetic study for their direct or indirect effects on recombination in natural populations for more than a century. In their most dramatic effect on meiosis, inversions generate aneuploid gametes when a crossover occurs between the interior of an inverted region and its non-inverted homolog (White 1973). As aneuploidy is often fatal in embryos or severely deleterious, this may translate to a fecundity cost for a parent heterozygous for an inversion; however, depending on a species’ reproductive biology, such reproductive fitness effects may be mitigated if the zygotes terminate early or before significant parental resource investment. Alternatively, crossovers may be inhibited directly in regions heterozygous for an inversion, or avoided by achiasmy or segregation of the aneuploid gametes to polar bodies in female meiosis (Sturtevant 1921; Sturtevant and Beadle 1936; Krimbas and Powell 1992; Coyne et al. 1993; Gong et al. 2005). Regardless of the differences between these scenarios, effectively fewer recombinant offspring are produced by inversion heterozygotes.

Inversions may also have direct effects on fitness when disrupting a gene, a regulatory element, or other genomic structure like a topologically associated domain (Lupiáñez et al. 2015; McBroome et al. 2020), but many inversion arrangements are expected to be neutral in and of themselves (Said et al. 2018). So, in the absence of indirect selection on the recombination effects of inversions, inversion variants are expected to evolve similarly to single nucleotide mutations: to be selectively removed when the net direct effects are deleterious, to sweep when the effects are beneficial, to be fixed or lost in proportion to frequency when neutral, or to be selectively maintained when associated with a balanced trait. Given that a common effect of inversion heterozygotes is the deleterious generation of aneuploid gametes, and that studies of extant inversions and engineered inversions suggest that inversions may commonly have minimal effect on regulation or coding regions, many inversion rearrangements may confer either deleterious or neutral fitness consequences in the absence of association with other variants.

In contrast to the above predictions, several well-studied examples are known in which inversion variants are maintained at an intermediate frequency within a population, presumably due to the indirect effects of inversions on recombination. Simulation and analytic theory have demonstrated that inversions associated with a local population can reduce recombination between locally adapted variants and unfit migrants, and thereby increase in frequency by association with the fittest haplotype locally (Kirkpatrick and Barton 2006; Charlesworth and Barton 2018). Such locally adaptive inversions could potentially also link together favorable combinations of variants between epistatically interacting loci (Dobzhansky 1970; Schaeffer et al. 2003; Hoffmann and Rieseberg 2008).

Unless migration rates are high, these models predict inversions to be at high frequency in their favored geographic range but at low frequency elsewhere. Alternately, chance associations of both an inversion and its standard arrangement counterpart with different recessive deleterious alleles may cause a form of overdominant selection, where heterozygotes carrying both main haplotypes experience none of the deleterious effects of a homozygous recessive. This associative overdominance can persist for extended periods, particularly when the rate of recombination between inversion arrangements is low (Zhao and Charlesworth 2016; Gilbert et al. 2020). This model predicts that such haplotypes will most readily accumulate in small populations, or when inversions that are initially rare accumulate recessive deleterious variants before the arrival of a linked positive variant or change in selection pressure. This may extend the polymorphic lifespan of an inversion, and remain stable in small populations, but is not expected to remain indefinitely stable as an intermediate frequency polymorphism in sufficiently large populations due to the greater population recombination rates allowing gene conversion and/or double-crossover events to generate haplotypes lacking deleterious variants. Inversions may also link drivers and enhancers in meiotic drive systems (Babcock and Anderson 1996; Mroczek et al. 2006; Courret et al. 2019), and these drive-associated inversions may persist at equilibrium frequencies in the presence of repressors (e.g. Bastide et al. 2022). In natural populations, it can be difficult to distinguish among these alternate hypotheses, and they are not always mutually exclusive.

Inversions in Drosophila melanogaster

Drosophila melanogaster is a widely studied organism in many fields of biology, including evolutionary and population genetics. Multiple polymorphic inversions segregate at meaningful frequencies in natural populations of D. melanogaster (Aulard et al. 2002; Kapun et al. 2016; Kapun and Flatt 2019), but even in this model system, it is not entirely clear which population genetic processes are most responsible for shaping their frequencies. In this species, achiasmy in males (Morgan 1910; John et al. 2016) and a lack of evidence for female fecundity costs for most inversions (Sturtevant 1921; Sturtevant and Beadle 1936; Coyne et al. 1993; Gong et al. 2005) could allow inversion rearrangements to have limited fitness consequences. Further, genome editing tools have allowed for the creation of synthetic inversions, and the finding that these structural rearrangements themselves are not responsible for the gene regulatory changes that are associated with natural inversions in D. melanogaster (Said et al. 2018). Instead, given that breakpoints of common D. melanogaster inversions occur disproportionately in regions that confer greater recombination suppression (Corbett-Detig 2016), the primary evolutionary relevance of these rearrangements appears to be the resulting reduction of recombinant offspring between otherwise highly recombining loci. Hence, D. melanogaster inversions would appear to be maintained for the sake of maintaining linkage associations between distant variants.

D. melanogaster originated in southern-central Africa, in seasonally dry Miombo and Mopane woodlands, and (potentially after adapting to a human commensal niche) expanded out across much of Africa and into the Middle East around 13,000 years ago, and into Europe around 1,800 years ago (Sprengelmeyer et al. 2020). With an ancestral Ne of around 2 million (Sprengelmeyer et al. 2020), D. melanogaster populations have high levels of standing diversity and high population recombination rates. Many D. melanogaster inversions have geographic clines repeated between several colonized regions (Mettler et al. 1977; Reinhardt et al. 2014; Kapun et al. 2016), and inversions often show geographic differences in frequency that exceed genome-wide differentiation between the same populations (e.g. Pool et al. 2017). It has therefore been suggested that local ecological adaptation may shape the distribution of D. melanogaster inversions (Kapun et al. 2016; Kapun and Flatt 2019). Indeed, it is difficult to explain their geographic differentiation without some type of spatially-varying selective pressures.

Yet, it is unclear whether ecological adaptation is a sufficient explanation for D. melanogaster inversion frequency patterns, for two reasons. First, most of the common inversions become less frequent as one moves away from the sub-Saharan ancestral range (or similar warm environments) into potentially more challenging, colder high latitude and altitude environments (Aulard et al. 2002; Kapun et al. 2016; Lack et al. 2016a; Pool et al. 2017; Kapun and Flatt 2019). As these inversions are all derived (being absent from closely related species), local adaptation theory would predict that they arose to protect local adaptation against ongoing migration, more likely in the novel environment if it is experiencing high levels of immigration (Kirkpatrick and Barton 2006). It is unclear why inversions would be needed to adapt to the species’ ancestral environments, or why ancestral “standard” arrangements would have an ecological advantage in derived environments that are very different from the ancestral range.

Second, even in the locations where inversions would hypothetically be ecologically advantageous, they rarely surpass intermediate frequencies (Pool et al. 2017; Sprengelmeyer et al. 2020). Under a local adaptation model, that observation would require a secondary explanation, such as very high levels of long distance gene flow or the fixation of recessive deleterious variants on inverted haplotypes, generating an associative overdominance at linked neutral loci (Frydenberg 1963; Sved 1968; Ohta and Kimura 1970; Zhao and Charlesworth 2016; Gilbert et al. 2020) that resists fixation. While migration may well limit inversion frequencies in some populations, we note that some putatively adaptive Single Nucleotide Polymorphism (SNP) variants do achieve much greater frequency differentials between populations (da Silva Ribeiro et al. 2022), and thus we would have to suppose that inversions are under relatively weaker selection to explain their lesser differentiation. Regarding the hypothesis of associative overdominance, we note that D. melanogaster populations do contain substantial recessive genetic load (Greenberg and Crow 1960), and it has been speculated that associative overdominance might boost inversion frequencies in D. melanogaster populations during founder events (Pool et al. 2012) or in isolated wilderness environments (Sprengelmeyer et al. 2020). However, it seems likely that in large human-commensal populations, D. melanogaster inversions would escape from linked deleterious variants through gene conversion or double crossover events with standard chromosomes, in light of both the antiquity of most of these inversions (Corbett- Detig and Hartl 2012) and the very high population recombination rates of D. melanogaster. Homozygous lines have also been generated for each common inversion, demonstrating a lack of perfect linkage of the inversions with recessive lethal or infertile variants (e.g. Lack et al. 2016a).

While local adaptation, gene flow, and associative overdominance may all contribute to geographic variation in inversion frequency among D. melanogaster populations, it is not clear that these processes are collectively sufficient to explain observed patterns. Alternatively, the allele frequencies and diversity patterns of at least some of these inversions might be primarily shaped by some form of balancing selection acting on inversion-linked variation, in which the balanced frequency is dependent on the environment (Kapun et al. 2023). Balancing selection on inversions has received increasing attention in recent literature (Wellenreuther and Bernatchez 2018; Faria et al. 2019). However, there has not been a proposed mechanism whereby alleles at multiple linked loci undergoing balancing selection would directly benefit from linkage and thereby maintain an associated inversion polymorphism under indirect selection. Here we propose one such mechanism, in which alleles that contribute to naturally frequency-dependent traits like competitive mating display simultaneously contribute antagonistically to another trait, such as the survival to reproductive maturity.

Sexually Antagonistic Polymorphism in D. melanogaster

Among mechanisms of selection, sexual selection is both prevalent, and provides a simple mechanism for frequency dependence (O’ Donald et al. 1997). Alleles that contribute to mate choice systems may often follow a negative frequency dependence within the population, as display success depends on the relative quality of a male among local competitors, and the more common a display-enhancing allele is, the fewer competitors it succeeds over. Combinations of display-enhancing alleles may similarly gain a disproportionate benefit together, as male mating success in competitive display systems can be skewed towards a few successful individuals (Bateman 1948; Jones et al. 2002; Tatarenkov et al. 2008, though see Gowaty et al. 2012; Hoquet et al. 2020) such that the last few alleles to push display into the top quantile contribute a significantly greater fitness increase to the genotype than the first few display alleles. This interaction of alleles is a form of epistasis - a non-independent fitness interaction between loci - that is emergent and non-specific to the genetic identity of the variants. In other words, the alleles contribute independently to the generation of a trait, and the fitness of the individual is a nonlinear function of that trait (Blows and Brooks 2003; Rest et al. 2013; Otwinowski et al. 2018). This epistatic interaction suggests a potential for indirect selection on the recombination-suppressing effects of inversions, in order to maintain combinations of sexually antagonistic variants that have the greatest fitness when carried together. The tendency of selection to favor the reduction of recombination when the haplotypes present are at a fitness optimum is well studied, and has been termed the ‘reduction principle’ (Altenberg et al. 2017).

While the emergent epistasis generated by a mate choice system may favor linkage of variants, alleles that contribute only to a sexually selected display are not expected to remain as balanced polymorphisms, rather they are expected to fix, albeit perhaps slowly as their fitness benefits diminish at high frequencies. However, if these alleles simultaneously contribute deleteriously to other components of fitness, they are pleiotropic (Stearns 2010), and given the sex-limited benefits of the display they would likely also be sexually antagonistic (Rowe et al. 2018), whereby traits increase the fitness of one sex but decrease fitness if found in the other. Pleiotropy can result from traits that are inherently physiologically coupled, from resource allocation restraints, or from the specific role or behavior of genes. Pleiotropy can be sexually antagonistic when the traits in the tradeoff are differentially important to different sexes, which may be common given the differing fitness requirements of the sexes in a population.

D. melanogaster mating involves a male display performance and female acceptance or rejection. A number of experimental evolution studies (Rice 1992; Rice 1996; Audet et al. 2023) and genetic studies (Innocenti and Morrow 2010; Cheng and Kirkpatrick 2016; Glaser-Schmitt et al. 2023) have confirmed the prevalence of sexually antagonistic loci in D. melanogaster. Pleiotropic alleles that trade off between survival and male display quality may be the simplest case of antagonism to consider for a model of balancing selection, and one highly relevant to the biology of D. melanogaster.

Given the relevance of sexual selection and sexual antagonism to Drosophila evolution, we propose that polymorphic D. melanogaster inversion and standard arrangements may help maintain haplotypes of alleles that specify opposite ends of a sexually antagonistic tradeoff. We first explore this model through forward simulation, to establish that it may in theory arise in a population and maintain inversion polymorphism over a significant number of generations. Then we examine the results of an experiment with an outbred laboratory population of D. melanogaster, to test whether common inversions show changes in frequency over the course of a generation in a manner consistent with a trade-off between survival and male reproductive success. Many such antagonistically pleiotropic scenarios are possible under different selective tradeoffs, but we focus on a specific scenario of sexual antagonism here.

Results

A Model of Sexually Antagonistic Inversions

Here, we consider a population featuring a competitive mate choice system that results in a skewed distribution of reproductive success among males. Within this population, we model a genomic region containing a number of linked loci, each with alternate antagonistic, pleiotropic alleles that trade off between male display success and survival likelihood (of females or both sexes). If mate choice is highly competitive, then superior display quality can enable males to mate successfully in excess of other display phenotypes they compete against. This gives a disproportionate benefit to additional mate competition quality if it allows a male to out-compete others, while additional survival costs stay proportional. Yet, fitness conveyed by display quality is also frequency-dependent; since mating success depends on the display qualities of other males, the relative advantage of a display trait will be diminished as more males carry it. And so, for an antagonistic variant that benefits display and begins under net positive selection, the benefit will decrease as the variant rises in frequency, until eventually it is equal in magnitude (across both sexes) to its pleiotropic survival cost – at which point an equilibrium frequency is reached. At this stage, the highest fitness among haplotypes may favor either of two extreme strategies - either producing among the most reproductively competitive males (by carrying a maximal number of antagonistic variants and paying the necessary cost of survival likelihood) or else paying none of the survival cost (at the expense of male reproductive competitiveness, carrying none of the antagonistic variants). In these conditions, we propose that selection will generate linkage disequilibrium between sets of associated alleles that together strongly favor either display or survival, and selection will favor the association of these haplotypes with alternate inversion karyotypes to reduce the recombination load experienced in heterozygous parents carrying one haplotype each from the extreme survival- and display-focused classes. We further predict that in a population that is close to the balanced equilibrium state, the population of adults will contain a higher frequency of survival-focused haplotypes than the population of zygotes, but the display-focused haplotypes will increase in frequency in zygotes relative to adults.

To build an intuition for the model, let us consider a simplistic model population with mate choice and phenotypic variation in male display. Particularly if each female can choose amongst several or many males, the variance of offspring number among males (Fig 1A) may be much higher than for females - yielding a relatively convex distribution in which the least and most successful males produce no or many offspring respectively. A genetic change that increases reproductive quality in a male can increase competitiveness and offspring count, but an increase in the display of a male with low quality (change i in Figure 1A) results in a smaller change in number of offspring than an increase in display of a male with high quality (change ii in Figure 1A), due to the reproductive skew towards highly competitive individuals. Therefore, multiple variants increasing the reproductive rank of an individual together have a larger fitness benefit than the combination of each substitution separately - a positive, synergistic epistatic effect.

Conceptual representations of the proposed model of inversion-associated balanced sexual antagonism.

(A) Hypothetical skewed distribution of male mating success, yielding a greater variance in reproductive success for males than for females in our model. (B) Fitnesses of display-favoring alleles under sexually antagonistic balancing selection, illustrating a pleiotropic variant that should rise in frequency when rare, but decline if frequency exceeds the balanced equilibrium value. (C) Fitnesses of display-favoring alleles at two haploid loci under the same model, illustrating synergistic epistasis between displaying-favoring alleles. (D) A hypothetical trajectory of four such mutations, in which B outcompetes A, an inversion links B and C to create a more strongly display-favoring haplotype, and then the addition of D to that haplotype furthers the accumulation of antagonistic variants, reaching an equilibrium frequency while displacing less extreme display-favoring haplotypes.]

And yet, as pointed out above, this positive epistasis is accompanied by a negative frequency dependence due to an antagonistic variant’s decreasing advantage over other males as its frequency rises and more competing males carry it. Display-enhancing variants with little or no pleiotropic cost may ultimately fix in spite of their diminishing advantage as a sweep progresses, whereas those with relatively high pleiotropic costs (modeled here as reduced survival probability) may be unconditionally deleterious and lost quickly. However, pleiotropic variants with more comparable fitness effects on display enhancement and survival reduction may end up at a balanced equilibrium frequency: when rare the allele experiences a large benefit from its display effects and is expected to increase in frequency, but above a certain frequency the benefit becomes less than the antagonistic cost (Figure 1B) and the allele is expected to decrease in frequency. In the absence of other forces, this dynamic is expected to maintain the allele under balancing selection indefinitely.

Consider a haploid population with two linked loci that carry such antagonistically pleiotropic alleles. Further consider the unrealistic circumstance that all alleles are at a frequency of 0.5 with perfect linkage equilibrium, and so all possible haplotypes are present at a frequency of 0.25. The haplotypes with all display-favoring alleles will then tend to be disproportionately chosen over other haplotype classes during male reproductive competition, yielding a positive epistatic interaction as seen in Figure 1A. If selection proceeds without recombination, the most display-heavy haplotype increases in frequency until reaching an equilibrium at which its relative advantage in mate choice is reduced enough to be balanced by the pleiotropic survival cost, which is assumed to be frequency- independent and non-epistatic among loci (Figure 1B). In this new state, the fitness of the most display-heavy haplotype and its opposite survival-focused haplotype are equally fit. However, haplotypes of intermediate composition all have lower fitness: they experience some pleiotropic cost, but due to the reproductive skew of mate choice they are selected disproportionately less often than the display-heavy haplotypes, so the net fitness contribution is negative (Figure 1C). Therefore, recombination between the two extreme haplotypes will generate less fit haplotypes in the gametes, and a local modifier of recombination that reduces this cost of recombination, like an inversion, will be favored by indirect selection (Feldman et al. 1980; Otto and Lenormand 2002). As inversions reduce recombinant gametes between haplotypes heterozygous for the inversion, an inversion introduced to this population should benefit from associating with one or the other extreme haplotype and should then increase to the balanced frequency of that haplotype.

Simulation Results

We investigated the predictions of the above model of inversion-associated balanced sexual antagonism using a purpose-built forward “Sexually Antagonistic Inversion simulator” (SAIsim; see Materials and Methods). This simulation program incorporates mutation of new nucleotide and inversion polymorphisms, crossing over and gene conversion (including between chromosomes heterozygous for inversions), drift, and sexually antagonistic selection (Materials and Methods). SAIsim allowed us to model the influence of polymorphic inversions on recombination, as well as the changes in linkage between sites entailed by the mutation and fixation of an inversion. It explicitly models mate selection, in that each female randomly selected to reproduce is assumed to encounter a number of randomly selected males, and the display values of each male are assigned based on the sum of genetic and environmental effects (with genetic effects based on genotypes at antagonistic loci), and the male with the highest display value then succeeds in this mating (Figure 2). We found it easiest to write a new simulator as no extant simulator accurately managed mutation and recombination of multiple inversion polymorphisms, as well as an explicit mate competition in generating offspring.

The layout of a single simulated generation. Simulated individuals each reach the reproductive stage with probability in proportion to the product of the survival effects of their alleles. To generate each offspring in the next generation, a surviving female is first randomly sampled (with replacement) to be the mother. Then a specified number of surviving males are randomly sampled to generate the pool of males encountered by the sampled female - without replacement for each pool and with replacement between pools. For each encounter pool, each male is assigned an observed display quality as the sum of the display effects of its alleles, plus a normally distributed noise effect for each encounter. The highest quality male in the encounter pool is selected as the father, and the offspring genome is generated with crossover and gene conversion events, where odd numbers of crossovers in the interior of an inversion are thrown out and resampled from the same parent. In the simulations presented, males do not recombine when generating gametes, in light of the biology of D. melanogaster.

With this simulator, we first confirmed that individual variants with an antagonistic pleiotropy between survival (initially of both sexes) and male reproduction can generate stable selectively balanced polymorphisms in silico. We simulated populations of N = 1000 carrying a single variant of a specified survival and display quality at an initial frequency of 0.5. After 20N generations, we calculated the average final frequency of the variant. From this analysis we can see that there is a range of paired trait effects for which the mutation remained at intermediate frequency after 20N generations (Figure 3). In a scenario where only females pay the survival cost, the balanced parameter space is shifted somewhat, but qualitatively the results are similar (Figure S1). Alleles with small effect sizes had only a narrow band of balanced selection, outside of which they were either fixed or lost, while large- effect alleles remained balanced across a larger area of simulated parameter space. While this balanced range is narrow, such variants may persist under balancing selection indefinitely in the absence of other forces.

Simulations show that a balanced equilibrium frequency exists for certain sexually antagonistic variants. The average final frequency of a single mutation with given effects on survival proportion (of both sexes) and male reproductive quality score is shown, with green indicating intermediate balanced frequencies. These simulations involve a single antagonistic variant with initial frequency 0.5, in a population of 1,000 individuals simulated for 20,000 generations, with further details as indicated in the Materials and Methods. The left panel depicts trait values at 0.05 increments, whereas the insert presents a finer-scale set of simulations, every 0.005 trait units.

To test the effects of association with an inversion, we initially ran two sets of simulations, considering two antagonistic variants found to undergo balancing selection separately. First, we spaced the two antagonistic loci at a varying genomic distances, in the absence of an inversion (Figure 4). In each case, the stronger antagonistic variant (modelled at fixed position 0.225M) persists, and so we focus on whether the weaker variant (at varying positions) also persists with it, or else is outcompeted by the more antagonistic variant. As seen in the upper panel of Figure 4, the weaker variant (pale red line) is retained at close linkage, as the haplotype containing both antagonistic variants acts as a single larger effect supergene. However, as the weaker variant moves farther from the stronger variant, recombination increasingly generates unfit intermediate genotypes compared to the extremes of the reproduction-survival tradeoff, and the stronger variant outcompetes the weaker variant among these haplotypes, leading to the loss of the weaker variant.

Simulated inversions persist as polymorphisms when linked to sexually antagonistic, pleiotropic variants. In turn, the presence of inversions facilitates the accumulation of antagonistic variation. Results of simulations with two defined antagonistic variants alone (upper panel) and in the presence of a defined inversion (lower panel). The 1 Morgan chromosomal segment is diagrammed above the plots, with the inversion breakpoints marked. The proportion of simulations in which the second, variably-positioned weaker antagonistic variant (survival and reproductive values 0.86, 0.12) maintains polymorphism is shown (light red line), dependent on the recombination distance from a stronger allele (0.75,0.3) at fixed position 0.225M, which always persists (dark red line). In the lower panel, the proportion of replicates in which the inversion is retained is also plotted (blue line). Populations start with an equal frequency of all haplotypes in a population of 1000 individuals. 1000 replicate simulations were run for 20N generations in each parameter combination.

We then examined the same spacings between antagonistic variants across a region containing a polymorphic inversion, with its left breakpoint close to the stronger variant, and the right breakpoint within the range of the positions in which the weaker variant was simulated (Figure 4, lower panel). When the inversion is added, the weaker variant is now retained when the two antagonistic variants are relatively more distant, due to association of the weaker variant with the right-hand inversion breakpoint (pale red line), at which point this association also preserves the inversion itself in a polymorphic state (light blue line). While more stochastic, a weaker antagonistic variant near the middle of the inversion also appears to maintain polymorphism more frequently in systems starting with an inversion than without, i.e. across approximate positions 0.30-0.45M in Figure 4, in spite of its weaker linkage with the inversion itself due to double cross-over events. The inversion is not maintained when the weaker antagonistic variant is within the inversion but tightly linked to the stronger variant (see positions 0.22- 0.25M); here there is little crossing-over for an inversion to suppress, and so the effect of the inversion approaches neutrality. Interestingly, there is a small window of weaker variant position in which the inversion maintains polymorphism even in the face of relatively tight linkage to the stronger variant (see positions 0.2-0.22M). Here, the weaker antagonistic variant is more tightly linked to the breakpoint than to the stronger variant in this range, and presumably benefits more consistently from the crossover reduction. The maintenance of the weaker variant drops more rapidly outside of the inversion than over similar distances in the simulation without inversions (positions 0-0.2M, bottom versus top panel). This seemingly counter-intuitive result may stem from crossover events that result in aneuploid gametes being resampled in our simulations, which effectively increases the crossover rate outside the inversion breakpoints, a behavior which should be biologically reasonable in many clades. Hence, for a weaker variant outside the left inversion breakpoint, the presence of the inversion can increase the crossover rate between the two antagonistic variants, and thus reduce the probability of of the weaker variant surviving by retaining linkage to the stronger one.

In further simulations, we began with a population initially free of variation, in which sexually antagonistic mutations arise at a specified rate, randomly located across a 1M simulated region. These mutations had pleiotropic effects increasing reproductive quality score and decreasing survival proportion, each drawn separately from distributions (Materials and Methods). Depending on the effect scores drawn, a mutation could be broadly advantageous (potentially leading to a selective sweep), broadly deleterious, or potentially subject to balancing selection (Figure 3). Inversions (of random lengths and positions) also arose at a specified rate. The frequency trajectories of antagonistic variants and inversions are tracked, and the simulation was run for 20N generations to allow near-equilibrium patterns to be established. These simulations established the potential of a population to mutate and evolve an inversion that links sexually antagonistic, pleiotropic alleles to maintain polymorphism in a balanced manner (Figure 5B), although the degree of antagonism accumulated depends strongly on the rate of crossing over (relative to the occurrence of antagonistic mutations; Figure 5C).

Simulated randomly mutating inversions persist as polymorphisms and accumulate sexually antagonistic variation when linked to sexually antagonistic, pleiotropic variants. (A) In simulated populations with sexually antagonistic mutations, simulations which also allow inversion mutations accumulated more average display benefit and survival cost across generations (right panel) than those without inversions (left panel). (B) When inversion mutation is enabled, a low crossover rate (left panel) allows populations to accumulate larger differences between the most common karyotype and all others than when the rate of crossover is particularly high (right panel). 1000 replicate simulations were run for 20N generations in each parameter combination.

The simulations that successfully maintained balanced antagonistic chromosomes also had an interesting distribution of allele and genotype frequencies of the balanced haplotypes at equilibrium. Without any selection, the inversion frequency follows a pattern characteristic of the expected neutral site frequency spectrum (Figure S2). However, in parameter spaces where simulations with shared survival costs between sexes maintained polymorphism under balancing selection, inversion frequency distributions were tightly clustered around an intermediate frequency (Figure S2, 6A). Strikingly, the frequencies of alternate haplotypes consistently clustered around 0.25 and 0.75 among the zygote population. We found that this outcome was related to the genotype frequencies of reproducing adults. These simulations evolve toward strong antagonism, and it is predominantly males homozygous for the display-favoring haplotype (whether that is inverted or standard) who succeed in reproducing, and these males therefore contribute one copy of the display-favoring haplotype to each of their male and female offspring. However, this haplotype’s significant cost to viability and longevity would select for the heterozygous females over the display-favoring homozygotes. Given that successful parents would predominantly be heterozygous females and display-homozygous males, three quarters of transmitted chromosomes have the display-favoring haplotype and one quarter does not, yielding the 0.25 and 0.75 inversion frequencies seen in simulations.

The sex-specific fitness dynamics present at equilibrium under this model also result in population-wide genotype frequencies that depart from standard Hardy-Weinberg expectations. If the preponderance of matings are between heterozygous females and display-favoring males, then homozygotes for the survival-focused haplotype will rarely be produced (as confirmed in Figure 6B). We note that in cases where the standard arrangement favors male display, then especially if antagonism is strong, this model could explain an observed deficiency or absence of inversion homozygotes, even if the inversion was not associated with any recessive deleterious variation. In contrast, heterozygotes may become disproportionately frequent after viability selection (Figure 6B). And yet, this heterozygote abundance reflects neither the existence of recessive deleterious variants on each haplotype, nor overdominance with respect to viability. Hence, our model may offer an alternative explanation for deviations from Hardy-Weinberg genotype frequencies observed for polymorphic inversions in natural populations.

Antagonism-associated karyotypes reach predictable equilibrium frequencies in simulations with antagonistic and inversion mutations, when survival costs are shared between sexes. (A) Histograms of the number of arrangements at a specified frequency across all simulation replicates in females (left) and males (right), colored by how the average survival effect of that arrangement ranks compared to the others in its population. Ranks are normalized to be relative to the median arrangement value, to better account for rare arrangements and ties. These simulations reach an equilibrium with two predominant arrangements, with the more survival-focused arrangement around an overall frequency of 0.25 and more common in females, and the more male-competition- focused arrangement around a frequency of 0.75 and more common in males. (B) Because successful males tend to be homozygotes and successful females heterozygotes, the less frequent of the two major arrangements (which tends to favor survival) has homozygote frequencies far below the Hardy-Weinberg expectation, and there is an excess of heterozygotes. These genotype frequencies shift somewhat between zygote (left panel) and adult (right panel) populations as the genotype affects survival.

We further ran simulations in which the survival costs were not shared, but instead specific to females, representing a distinct, stronger form of sexual antagonism. In these simulations, we surprisingly recovered three different equilibrium arrangement states (Figure 7). 20.0% of simulations exhibited a state similar to the shared costs simulations presented in Figure 6, where we infer that fathers were homozygous for a Z-like display-favoring arrangement, and mothers were heterozygous and carried a W-like survival-focused arrangement that is observed in surviving but not reproducing males. 37.0% of simulations also developed two predominant arrangements, but with Y-like instead of W-like dynamics. In these cases, we infer that the mothers were homozygous for an X-like survival-focused arrangement, but the fathers were heterozygous and carried a Y-like display-focused arrangement that was apparently lethal in females. 40.6% of simulations instead featured three intermediate frequency haplotypes: with both Y-like and W-like arrangements restricted to the fathers and mothers respectively, and a third shared arrangement with intermediate phenotypic effects.

In simulations with female-limited survival costs, three distinct outcomes are observed involving either two or three balanced haplotypes. From simulations with randomly occurring antagonistic and inversion mutations, histograms of inversion frequencies across simulations for surviving females (left) and males (right) are partitioned into three categories based on sex-specific haplotype frequency outcomes. The “Favors Female” category (top) includes simulations in which there are exactly two arrangements with frequencies between 0.1 and 0.9 in both sexes (20.0% of replicates). The “Male-Specific” category (middle) includes simulations in which there are two arrangements with whole-population frequencies between 0.1 and 0.9, but one arrangement is absent in adult females (37.0% of replicates). The “Three Arrangement” category (bottom) includes simulations in which there are exactly three arrangements with whole-population frequencies between 0.1 and 0.9 frequency (40.6% of replicates). 2.3% of simulations did not fit in these categories and had only one predominant arrangement. As in Figure 6, inversions are colored by how the average survival effect of that arrangement ranks compared to the others in its population, and ranks are normalized to be relative to the median arrangement value, to better account for rare arrangements and ties.

Specific properties of our simulations may have influenced the occurrence of these three stable states under the female-limited cost model. To reduce the computational complexity of both crossovers and arrangement-specific trait values, the simulations do not allow two inversions on the same individual chromosome, whether overlapping or not, though two chromosomes sampled from the population may have distinct inversions with overlapping breakpoints. This means that if two arrangements, each with an inversion, become the prevalent haplotypes and thereby remove all standard arrangement chromosomes from the population, there is no mutational path to generate a third arrangement. However, a portion of the two-state simulations still had the ancestral, non-inverted arrangement as one of the two haplotypes present, which could have been mutated to a third arrangement to establish the three-arrangement pattern. We are not sure what selective or mutational barriers may prevent transitioning between each state once established. One clear barrier is that it is difficult to generate a less extreme arrangement from a Y-like arrangement, as these can accumulate an indefinite number of mutations since males experience no costs.

Empirical tests for inversion-associated tradeoffs

A specific motivation for developing and evaluating the model of inversion-associated, balanced sexual antagonism was to potentially better explain the pattern of inversion frequencies observed in D. melanogaster, in which multiple inversions reach some of their highest frequencies in populations from the ancestral range or with similarly tropical/subtropical climates. We therefore set out to test experimentally whether any of four inversions common in a Zambia population of this species show evidence for tradeoffs between male reproductive success and survival. We assembled an outbred population with moderate frequencies of each inversion to obtain karyotypically diverse males. We set up four crosses using these males along with females from one of four different inversion-free inbred Zambia lines. This crossing scheme is represented in Figure 8. These parental crosses occurred in a single large population cage, with populations of at least 600 of each sex. We developed a PCR and next-generation sequencing-based assay to estimate inversion-associated SNP frequencies in each pool of males and in each corresponding pool of embryos. These SNPs (given in Table S1) were based on our own analysis of variants that showed perfect association with inversions among 197 Zambia genomes (Table S2; see Materials and Methods). We then implemented a resampling method to test for significant inversion frequency changes between parents and offspring (while accounting for the variance observed in technical replicates; see Materials and Methods), in order to determine whether males carrying each inversion were more or less successful in reproducing than the average male. We also compared frequencies between embryos and aged adults that eclosed either in the first or second half of the collection period, to investigate the relationship between inversions and viability/longevity. We calculated p-values from the resampling tests to assess the significance of each individual change in inversion frequency between relevant samples (Table S3). However, in accord with our primary question regarding fitness tradeoffs, we primarily focus here on joint tests of two opposing frequency changes, as further described below. We could therefore test the hypothesis that a given inversion had opposing influences on survival and male reproduction, which would provide evidence consistent with our model of balanced sexual antagonism, while also examining whether any such tradeoffs depend on maternal genetic background.

The layout and potential expectations of the laboratory evolution experiment. We cross outbred males from a high-inversion population to inbred, non-inverted females from a specific inbred line, and collect DNA samples from the fathers, embryo offspring, and aged adult offspring (left). We hypothesize from our antagonistic pleiotropy model that the inversions will experience opposing selection between fathers and embryos versus between embryos and aged adults (middle). To test the significance of observed frequency changes, we designed a resampling model to represent the sources of neutral variation expected in the experiment (right).

When we initially examined the joint effects of male reproduction and survival, we found that in a majority of cases, a given inversion increased in frequency in adult offspring compared to the null expectation based on the parental male pool (in 13 out of 16 inversion-line trials; binomial P = 0.02; Figure 9). Given that inversion homozygotes are not generated by our crossing design, these results are likely to indicate that inversion heterozygotes often had higher fitness than standard homozygotes (across male reproduction and survival jointly). We note that our standard homozygotes are not identical-by-descent but are instead heterozygous for unique standard-arrangement chromosomes from two independent Zambia strains, and that long genomic tracts of identity-by-descent between Zambia strains are quite rare in this large, ancestral-range population (Lack et al. 2016a). Hence, we do not expect that inversion heterozygosity is buffering against deleterious effects of inbreeding. Instead, cases where inversions are associated with fitness advantages suggest either a simple directional advantage of inversion-associated alleles for a given fitness trait, or else an advantage for heterozygous inversion/standard genotypes in particular. In the former case, the inversions may still be under balancing selection in the wild, if the equilibrium frequency of the inversion was shifted due to lab conditions such that one karyotype was always favored in the lab environment tested. This scenario would mirror previous findings in which inversions found at intermediate frequencies in natural D. melanogaster populations (Inoue 1979), as well as inversions from seaweed flies known to be under balancing selection in the wild (Mérot et al. 2020), were both inferred to be under purely directional selection in the lab. We might still expect some difference in benefit across the life history of a single generation if there is a balanced tradeoff in some alternate environment or at a different frequency, consistent with the differences observed in our populations.

Some Drosophila inversions show evidence for antagonistic tradeoffs and sex-specific survival. (A) Testing for antagonistic fitness effects on reproduction and survival. Each inversion’s frequency in parents, embryo offspring, and adult offspring are shown for each maternal inbred line. Parental frequency is taken as the paternal frequency divided by two to account for the inversion-free inbred mothers. Adult offspring frequencies are taken from the sum of estimated allele counts across all adult offspring cohorts. * indicates p-value < 0.05 for evidence of significant reversed frequency change from parental to embryo and from embryo to offspring, combined across all four maternal line crosses and corrected for multiple tests, when compared to neutral simulated experiments (details in Materials and Methods). (B) Testing for evidence of sex-specific survival effects. The changes in inversion frequencies between embryos and both female and male adult offspring are illustrated. * indicates p-value < 0.05 for significant differences in female versus male inversion frequency among adult offspring, combined across all four maternal crosses and corrected for multiple tests.

When we separated the effects of male reproduction (based on father to embryo changes) and survival (embryos versus adults), we found that the effects of inversions on survival were generally consistent across maternal genotypes. Three inversions were associated with increased survival: In(3R)K was in all four crosses, while In(2L)t and In(2R)NS each showed elevated survival with all maternal strains except ZI418N (Figure 9A). In contrast, In(3L)Ok was associated with lower survival in each cross (Figure 9A). With regard to male reproduction, the trajectory of inversion frequencies from fathers to embryos to offspring was less consistent between different crosses (Figure 9A), implying that differences in male reproductive success were dependent on the maternal genotype, which is consistent with past findings in D. melanogaster (Reinhart et al. 2015). Despite such variability, each of the four inversions had multiple frequency changes that departed strongly from null expectations (Figure 9A; Table S3), as might be expected if selection was present, but varied in strength and direction between crosses and life stages.

To test our central hypothesis of a tradeoff between male reproductive success and cumulative survival during development and adult aging, we evaluated the significance of each change in inversion frequency between the paternal and embryo cohorts, alongside the change in inversion frequency between the embryo and combined adult offspring cohorts. Among these results, In(3R)K had particularly consistent changes in frequency across replicate lines, with a decrease between parents and embryos, and then an increase between embryos and aged offspring (Figure 9A). When we assessed the likelihood of observing contrasting frequency shifts between life stages of this magnitude, integrating across the four maternal crosses and accounting for multiple test correction, we found that the In(3R)K pattern was significantly non-random (p = 2.70 × 10-3; Table S4; see Materials and Methods). The other inversion that gave reversed frequency shifts from cross-averaged data was In(3L)Ok, which unlike In(3R)K, was associated with (variably) higher male reproductive success and (consistently) lower survival. The frequency reversal observed for In(3L)Ok was also found to be statistically significant (p = 2.97 × 10-7).

We also tested for frequency differences between the different sexes of adult offspring and between the early-eclosing and late-eclosing cohorts of adult offspring. All four inversions displayed fairly consistent differences in the production of female versus male offspring (Figure 9B), suggesting that each of them have differential effects on sex-specific survival. Strikingly, In(3L)Ok and In(3R)K each displayed sex survival differences in alignment with their antagonistic effects implied above: In(3L)Ok favored male survival more-so than female survival (p = 0.0450), whereas In(3R)K favored the survival of females over males (p = 4.61 × 10-3; Table S4; see Materials and Methods). The other two inversions, In(2L)t and In(2R)NS, each consistently favored female over male survival in all four maternal crosses, although their initial unidirectional cross-combined p values (0.0260 and 0.0126) were not significant after multiple test correction (p = 0.282 and 0.137). In contrast, we found no consistent effects of inversions on eclosion time (Figure S3). Overall, we conclude that D. melanogaster inversions have distinct and often context-dependent influences on male reproduction and on survival, with some evidence for antagonistic tradeoffs of the sort proposed in our model.

Discussion

Here, we have introduced a new conceptual model in which (1) sexually antagonistic variation is present and subject to balancing selection that maintains extreme genotypes favoring either male reproduction or survival, and (2) inversions are predicted to facilitate the aggregation of progressively more antagonistically differentiated haplotypes. We have tested this model using simulations, confirming the predicted efficacy of balancing selection and demonstrating that inversions indeed facilitate the construction of haplotypes containing multiple antagonistic variants. Finally, we have deployed Drosophila lab experiments to uncover some evidence for inversion-associated tradeoffs between reproduction and survival, depending on the specific inversion and strains tested.

Inversions in D. melanogaster

We have presented results from a single generation laboratory evolution experiment, in which we crossed outbred male D. melanogaster carrying inversions In(2L)t, In(2R)NS, In(3L)Ok, and In(3R)K to four separate inbred female lines and tracked the frequency of the inversions in the initial males, embryonic offspring, and aged adults. For most inversions, the strength and direction of the tradeoff varied notably between crosses to different inbred maternal lines, mostly due to maternally-dependent variation in male reproductive-success. However, greater consistency was observed for In(3R)K, which was associated with lesser male reproductive success but greater survival. A significant tradeoff was also inferred for In(3L)Ok, which favored male reproduction (albeit variably) over survival. These results provide significant but measured evidence for the presence of sexually antagonistic autosomal inversions in this model species, while emphasizing the importance of genetic background.

As indicated in the Introduction, models like ours in which inversion frequencies are influenced by balancing selection on pleiotropic tradeoffs may help explain multiple aspects of observed D. melanogaster inversion frequencies. This class of models inherently accounts for the intermediate maximal frequency of most D. melanogaster inversions in natural populations. To the extent that inversions favor male reproduction, this could also help account for the predominantly autosomal locations of common inversions in this species, given that male-specific fitness effects have a weaker influence on the frequencies of X-linked variants since the X chromosome spends only one third of its time in males.

This model could also help explain the observed geographic variability of inversion frequencies, as a change in environment may change the fitness value of either pleiotropic effect of the alleles. If a population experiences a novel harsh environment with low survivability, then the population density may decrease, and the number of encountered competing males may decrease as well, reducing the relative fitness advantage of enhanced male display. If there is a genetic tradeoff between mating display and viability, the equilibrium frequency of an antagonistic inversion that favors the genetic extreme of display should be lower. Additionally, in hostile environments, the survival cost of display-favoring variants could be increased, which could further shift their equilibrium frequencies downward or render them not worth carrying at any frequency. These predictions could contribute to the observed negative relationships between some inversion frequencies and latitude or altitude (Kapun et al. 2016; Pool et al. 2017). As a specific example, it is notable that in a high altitude Ethiopian environment with year-round cool weather, which hosts some of this species’ most pronounced phenotypic evolution (Bastide et al. 2014; Lack et al. 2016b), inversions are virtually absent - occurring at the lowest frequencies of any analyzed population (Sprengelmeyer et al. 2020). Consistent with such reasoning, a disadvantage of some inversions in challenging thermal environments (both cold and hot) was supported by a previous experimental evolution analysis (Kapun et al. 2014). However, our experimental results include evidence that some inversions, such as In(3R)K, may have the opposite tradeoff - that the inverted arrangement is associated with survival success instead of reproductive success. Notably, we have only tested survival and reproduction in a benign laboratory environment, and not under challenging environments such as cold temperatures. Other studies in D. melanogaster, including those on focused seasonal evolution, have found evidence for important tradeoffs between reproduction and robustness to environmental challenges (e.g. Behrman et al. 2015). Hence, it would clearly be desirable to study the effects of inversions on a wider range of potential tradeoffs. For example, Said et al. (2018) suggested that inversions In(2L)t and In(3R)K might be maintained by immunity-related tradeoffs, in light of an over-representation of immune-related genes among differentially expressed transcripts between inverted and standard karyotypes.

In addition to a broader array of balanced tradeoffs that could modulate D. melanogaster inversion frequencies, drift and other forms of selection may contribute as well. Although we emphasize above the potential interplay between ecological adaptation and balanced tradeoffs, some inversions could be impacted by simpler forms of directional selection. However, we must then invoke secondary explanations to account for their limited maximal frequencies (see Introduction). Other forms of balancing selection may also come into play as well. It is clear that some variation in this species is subject to temporally-varying selection, and some inversions have displayed seasonal frequency differences (Kapun et al. 2016; Lange et al. 2022). As with spatial clines, temporal shifts could reflect either simple directional selection or else environmentally-modulated tradeoffs in a model such as ours. Spatial selective pressures may vary both between populations and within population scales.

As indicated in the Introduction, balancing selection on inversions could also be achieved by ‘associative overdominance’ (Zhao and Charlesworth 2016; Gilbert et al. 2020), where two haplotypes are balanced by recessive cost to both homozygotes. Inversions may be particularly susceptible to carrying recessive deleterious alleles if population size and/or inversion frequency are low, since under these circumstances inversion homozygotes are rarely generated and hence the crossover rate among inverted chromosomes is low, plus there are fewer inverted chromosomes to facilitate rare double recombination events with standard chromosomes. However, this dynamic is expected to decay with increased age and population size of the inversion, as mutation, gene conversion, or double crossover resolve the negative linkage between recessive deleterious alleles. Most common inversions in D. melanogaster are both relatively old and have large effective population sizes. Inversions including In(2L)t, In(2R)NS, and In(3R)K have breakpoints estimated to be several times older than the expansion of the species out of southern-central Africa (Corbett-Detig and Hartl 2012; Sprengelmeyer et al. 2020). Those three inversions, plus In(3L)Ok, for which inversion age has not been estimated, all maintain frequencies greater than 0.15 within very large populations (Sprengelmeyer et al. 2020), and so it is not clear that the persistence of recessive deleterious or lethal alleles in the inverted regions is expected to be so much higher than for standard arrangements that inversions should be maintained by associative overdominance. In agreement with these predictions, many wild-derived inbred strains of D. melanogaster are homozygous for inverted haplotypes (Lack et al. 2016a). Furthermore, as noted in the Results, even an observed lack of inversion homozygotes could reflect strong sexual antagonism as opposed to recessive deleterious variation. Nevertheless, further simulation and experimental testing regarding the potential influence of associative overdominance and other processes on inversion frequencies in D. melanogaster and other species would be desirable.

Generality of Balanced Pleiotropic Inversions

We have proposed that some polymorphic inversions in D. melanogaster, and in other species with analogous mating dynamics, may be maintained in part through balancing selection acting on a frequency-dependent tradeoff. While we explored the tradeoff between male reproduction and viability, we emphasize that any antagonistic pleiotropy involving at least one frequency-dependent trait has the potential to generate similar inversion polymorphism. Such antagonistic pleiotropy may be common in evolution, and might be useful in explaining the many examples of inversions associated with distinct phenotypes.

For the specific model explored here, of antagonistic pleiotropy between survival and display in a mate choice system, the prevalence of the dynamic across species and populations is unclear. It seems plausible that a similar model of antagonistic pleiotropy and balancing selection under mate choice underlies the balanced inversion system in the European ruff Philomachus pugnax, where three inverted haplotypes on an autosome determine aggressive, cooperative, and female-mimic male lekking morphs, albeit with a dominance relationship and homozygous inviability for an inverted haplotype (Küpper et al. 2016). More generally, a number of factors contribute to the potential prevalence of this dynamic. Mate competition and choice during reproduction is rather common among animals, and this may mean sexually antagonistic inversions are plausibly also common. The model depends on the distribution of fitness effects of new mutations and the availability of sexually antagonistic variation, but such variation does seem prevalent (Zajitschek and Connallon 2018; Ruzicka et al. 2019). More likely to limit the prevalence of such inversions are the meiotic costs experienced by inversion heterozygotes in some systems (White 1973), whether sexually antagonistic variants are clustered in a region that would allow inversions to increase linkage, and whether other mechanisms like sex-biased expression (Ingleby et al. 2015), gene duplication (Connallon and Clark 2011), and linkage to the sex chromosome or sex-determining locus would resolve the antagonism more frequently (though see following subsection on sex chromosome evolution).

Perhaps more importantly, there are a number of well-studied cases of inversion polymorphism that may follow this dynamic of balancing selection between haplotypes demonstrating antagonistic pleiotropy, but which do not involve sexual antagonism. Papilio butterfly populations harbor well-studied mimicry polymorphisms that segregate between inverted segments and may often experience frequency-dependent selection (Clarke and Sheppard 1960; Kunte et al. 2014; Poul et al. 2014). The snail Cepaea nemoralis harbors diverse and distinct shell patterning haplotypes associated with inversions that remain balanced within single sampling locations (Cook 2013; Surmacki et al. 2013). Several inversions in different ant species segregate a single-queen colony strategy from a cooperative multi-queen strategy within the same range (Wang et al. 2013; Libbrecht and Kronauer 2014).

Essentially, any population with antagonistic pleiotropy involving some form of negative frequency-dependent balancing selection on one of the phenotypes might exhibit the same dynamic. The population would experience similar divergent selection towards alternate phenotypic extremes, defining an emergent level of epistasis between alleles that trade off between the pleiotropically determined traits. This idea integrates with existing thought on the evolution of recombination and divergent selection on balanced systems. For example, Kopp and Hermisson (2006) have proposed a specific form of frequency dependent divergent selection (FDDS), involving competitive exclusion between similar genotypes for a trait otherwise under stabilizing selection, and this model was expected to result in single or few loci of large effect. Their model did not consider pleiotropy or linkage modifiers, but still had comparable dynamics in that selection in the FDDS regime favored small numbers of large effect loci, while the inversion haplotypes in the model presented here simplify the recombination architecture to approximate a single locus. This simplification of the genetic architecture can be further seen as related to the reduction principle in recombination modifier theory (Feldman et al. 1980; Altenberg et al. 2017). When a population experiences selection towards two (or more) divergent optima, under a selection regime that provides a balancing mechanism to maintain both, indirect selection will favor a reduction of the genetic architectures to prevent the generation of intermediate phenotypes. The likelihood of either model likely rests on the inherent polygenicity of the trait under selection, with simpler traits being more likely to resolve by single locus FDDS dynamics than by inversion association (e.g. Yassin et al. 2016).

Autosomal inversions and sex chromosome evolution

If balanced, sexually antagonistic, autosomal inversions occur frequently enough, they may also contribute to the evolution of sex chromosomes. Given an established inversion with significant sexual antagonism, the inverted haplotype that favors one sex still experiences an antagonistic cost of transmission to offspring of the opposite sex. Mutations that link the haplotype with the sex it benefits are therefore favored under selection, and this can occur either by fusing to an existing sex chromosome (Charlesworth and Charlesworth 1980), though this only fully resolves the antagonism for linkage to Y and W chromosomes, or by linkage to a new sex determining mutation (van Doorn and Kirkpatrick 2007). In particular, we discovered that strong inversion-linked antagonism could yield a scenario where essentially all highly-successful males were homozygous for a display-favoring karyotype, whereas reproducing females were overwhelmingly heterozygotes (Figure 5). This situation is similar to the ZW sex chromosome system, suggesting that this particular pleiotropic trade-off may favor resolution by linkage to a female sex determination factor. Further, simulations in which only females experienced antagonistic survival cost usually generated populations with a chromosome essentially lethal in females, thus exclusively transmitted in males similar to a Y chromosome (Figure 7). These simulations sometimes had three states that appeared stable at 20N generations, including Y-like, W-like, and shared intermediate arrangements. This three-arrangement stable polymorphism seems unusual for a natural population, but might be interesting to explore in populations with polymorphic sex determination.

There are often deleterious effects of chromosome fusion or novel sex determination (Pennell et al. 2015; Saunders et al. 2019), which may represent a general obstacle to sex chromosome evolution. However, balanced sexually antagonistic inversions may concentrate enough antagonism that the built-in benefit of resolving it may sometimes overcome the fitness costs of evolving new sex chromosomes (Blackmon and Brandvain 2017). Hence, antagonistic inversions might be more likely than other loci to be involved in the evolution of new sex chromosomes.

A basic prediction of the above hypothesis is that at least one inversion difference between sex chromosomes should already exist at the time of new sex chromosome formation, and genomic differentiation therefore begins between inversion karyotypes on an autosome even before sex-specific transmission of the region. This contrasts with current models of sex chromosome evolution in which inversion accumulation and genomic differentiation accrue only after sex chromosome formation, due to advantages in suppressing recombination involving sexual antagonistic variants already associated with a sex-determining locus (Wright et al. 2016). Additional genomic studies of new sex chromosomes are needed to assess the potential role of sexually antagonistic inversions in sex chromosome evolution.

Summary and Future Prospects

We have advanced a model in which inversion polymorphism is maintained by the frequency-dependent fitness effects of a pleiotropic tradeoff associated with inversion-linked genetic variation. We have used sexual antagonism associated with increased male reproduction versus survival as a specific motivating case, and we explore the predictions of this model using a novel simulation algorithm. We confirmed model predictions that some antagonistic variants can be maintained at an equilibrium frequency, and that inversions can allow multiple antagonistic variants to persist and form more strongly antagonistic haplotypes. We complemented our conceptual and computational modeling with laboratory experiments designed to detect fitness effects of four inversions common in an ancestral range population of D. melanogaster. We indeed found some evidence of tradeoffs between male reproduction and survival for these inversions, with two out of four inversions showing significant evidence for such a tradeoff, although we also observed a strong dependence of male reproductive success on female genetic background.

In light of the persistent mystery of how D. melanogaster inversion frequencies are determined, and how readily we detected tradeoffs involving male reproduction and survival, we suggest that further studies investigating the potential fitness tradeoffs of inversions in this model system are strongly warranted. While some such studies have previously been conducted, the applicability of our amplicon sequencing approach to estimate inversion frequencies in large pools of flies may improve the sensitivity of such experiments. Our understanding of the phenotypic impacts of D. melanogaster inversions would benefit from investigations of a wider range of fitness- related traits, including female fecundity, immunity, and performance in challenging environments such as cool temperatures and high population densities.

Clearly, it will also be important to test the generality of our model by estimating the fitness associations of inversions in a wider range of species. And particularly in the case of species with very recent changes sex chromosomes, it will be of great interest to test whether one or more inversions are already present between these new sex chromosomes, as predicted a model in which sex chromosome evolution involves the resolution of inversion-associated sexual antagonism. Such studies will advance our broader understanding of the roles of balancing selection, pleiotropic tradeoffs, and genome structure in shaping genetic variation and genomic evolution.

Materials and Methods

Simulations

We developed a new individual-level forward simulator (‘SAIsim’ for Sexually Antagonistic Inversion simulator) written in python 3.7 for the simulation of inversions and sexually antagonistic alleles at infinite loci, with uniform rates of mutation, crossing over, and gene conversion (per bp, per generation) over a number of independent chromosome arms. The dynamic we were interested in modeling requires direct accounting of pleiotropy between different components of fitness, crossing over and gene conversion within inversions (as well as the possibility of inversion fixation), and directly modeling male display and female choice, and no existing simulator provided this functionality. These simulations were used to establish the potential for the model to occur in a population generally, and not to compare against specific empirical observations. We have used literature estimates of mutation rate and conversion rate of Zambian D. melanogaster for these simulations (Comeron et al. 2012; Huang et al. 2016), but have no information on the joint distribution of effects of mutations on our model of survival and display in D. melanogaster. So, we sample from relatively broad but arbitrary combinations of mutation effects, only assuming that there is likely some degree of antagonistic pleiotropy naturally present in genomic variation.

In SAIsim, a population is instantiated as a python object, and populated with individuals (also python objects) based on provided (or initially empty) genomes. Each generation, as diagrammed in Figure 5, the population calculates a survival probability for each individual as the product of the survival values of each variant they carry, which range [0,1]. Male and female survival costs can each be included in a sex-specific manner, but only globally in the simulation, not per mutation. For each offspring in the next generation, a mother is selected randomly, weighted to survival values, with replacement. Then a set number of potential fathers (i.e. the uniformly specified encounter number) is randomly sampled for each offspring from the full population, weighted by their survival values. The males each have a display quality given as the sum of the reproductive values of their alleles, plus a normally distributed additional noise with standard deviation equal to one display quality unit. SAIsim allows the specification of mutation rates and the distribution of effect sizes of new mutations; our simulations use mutations with uniformly distributed quality values in the range [0,1]. A noise parameter, normally distributed with a standard deviation of 1, is also applied to each male display value, with the noise effect generated separately for each interaction. Once the male display qualities are calculated based on genotype and noise, the mother of each offspring chooses a father deterministically by the highest quality. As offspring are generated, crossover locations are sampled, and resampled in cases where the resultant gamete would be aneuploid. Gene conversion is modeled as a flat probability per locus, with each event independently affecting just one locus, rather than modeling a conversion tract length. Inversions and new mutations are added to the resultant gamete, with inversions resampled when overlapping an inversion present on the chromosome. The simulation is intended to represent D. melanogaster life history and allows removal of recombination in males, though this also removes gene conversion by default.

SAIsim’s object-oriented design allows easy specification of a population model by a new user familiar with the basics of python scripting. A population object can be parameterized and populated in a few lines of python script, and then parameters can be changed between periods of simulation, allowing bottlenecks and demographic changes as well as changes in mutational parameters between generations. Replicate SAIsim simulations were performed via the UW-Madison Center For High Throughput Computing (CHTC), which manages cycle servers using the HTCondor distributed computing project (Thain et al. 2005; Erlandson and Theisen 2018).

Experimental Populations and Sequencing

To act as a high-inversion source population for the potential fathers in our experiment, we combined 15 males and 15 females from each of 10 inbred strains derived from wild flies collected in Zambia. These 10 strains were chosen to collectively carry the following common inversions: In(2L)t, In(2R)NS, In(3R)K, In(3R)P. We then collected at least 500 F2 males from this high-inversion population to use in each of our independent maternal line experiments, and these males were expected to carry a wide range of inversion genotypes. Each batch of high- inversion males was crossed to an equal number of virgin females from a single inbred Zambian strain (Figure 6). Four such experiments were performed, each using one inversion-free maternal strain. Since no inversions could be inherited through the mothers, inversion frequencies of successfully reproducing F2 males could be inferred from their pooled offspring. Unless otherwise noted, all flies were kept in a lab space of 23°C with around a degree of temperature fluctuation and without a controlled day/night cycle. All fly food used was prepared in batches of 4.5 L water, 500 mL cornmeal, 500 mL molasses, 200 mL yeast, 54 g agar, 20 mL propionic acid, and 45 mL tegosept 10% (in 95% ethanol). The four parental populations were each allowed to mate freely in clear plexiglass cages of dimension 19.3 by 19.3 by 30.0 cm, with a loose mesh covering on one end for access by hand, twisted and clipped shut with a binder clip. The parents were given either fly food bottles or grape agar in 7.9 cm petri dishes as food or laying substrate on alternating days over two weeks, after which all surviving male parents were counted and frozen. The grape agar was prepared in batches of 500 mL from FlyStuff.com grape agar powder, and wet Red Star yeast was brushed over the agar plates before use to encourage laying. Embryos were collected from the agar, counted, and frozen for sequencing. Embryos laid in the bottles were allowed to hatch and develop into adults. These adult offspring were separated by sex and by the time after laying at which they eclosed. In our husbandry conditions, the earliest eclosion was 11 days after laying, but this was rare and a first eclosion at 12 days was more common among the bottles. Most flies took several days longer to develop, and we considered flies collected in the first 6 days of a 10 day collection window to be the early-eclosing pool, in order to split the pool sample sizes closer to evenly. The collected adult flies were then aged by sequentially transferring them between bottles weekly for approximately 2.5 months, and then the adult offspring were sexed, counted, and collected. A 10-week aging window was chosen based on the time taken for a control set of four bottles of flies from the paternal outbred population to reach 80% mortality. All adult and embryo samples were frozen at -80°C before DNA extraction. This sampling scheme was chosen to identify inversion frequencies in the set of potential parents, in embryonic offspring, and in older adults, and so to assess potentially selected changes in inversion frequency across male reproduction or in egg to adult viability plus longevity.

DNA for each replicate age group was extracted by a phenol-chloroform extraction protocol used previously in preparing DNA for the Drosophila Genome Nexus and Drosophila Population Genomics Project. The protocol is similar to protocol ‘BPC’ used in DPGP phase 1 (Langley et al. 2012), except with the addition of proteinase-K during homogenization of adult flies, and generally based on the common quick preparation protocol presented in Drosophila Protocols (Huang et al. 2009). Inversion genotyping was performed by amplicon sequencing, where amplicons associated with each inversion were designed to target at least one single nucleotide polymorphism in the inversion interior that is fixed for alternate alleles between the inversion karyotypes, while still sharing conserved primer regions across karyotypes to prevent primer bias. Genetic variation in these inversions was assessed using previously sequenced and inversion-called haploid genomes from the Zambian D. melanogaster population, available from the Drosophila Genome Nexus (Lack et al. 2016a). The amplicon region and primer selection was performed with a custom script. The script calls Primer3 version 2.4.0 for primer candidate generation (Koressaar and Remm 2007; Untergasser et al. 2012; Koressaar et al. 2018). Primers thus selected were ordered from IDT as oligos ligated to Illumina stubby adaptors [Table S1]. Illumina sequencing libraries were then prepared for each sample following the Illumina metagenomic amplicon sequencing protocol (Illumina 2013), with 50 ng/uL diluted DNA samples. and sequenced 2x150 on an Illumina NovaSeq to at least a depth of 20000 reads. Depth was selected to ensure high likelihood of detecting a 0.025 change in inversion frequency from an initial frequency of 0.10, based on simulated resampling of the collected flies, DNA extraction, and sequencing.

Sequence Analysis and Statistics

Sequence preparation and analysis were performed using custom pipeline scripts. Read trimming, alignment, and QC were performed using bwa version 0.7.17-r1188 (Li 2013), GATK version 3.4-46 (Van der Auwera and O’Connor 2020), and Picard version 1.79 (Broad Institute 2019). Reads were trimmed at the ends to remove segments with base quality less than 20, and read pairs were filtered to retain only those pairs where both reads aligned with mapping quality >20. SNP identity at the sites of interest was called from those remaining reads with base quality >20 at the SNP of interest. Due to the presence of chimeric reads in the amplicon library sequencing, we chose to use a single discriminant SNP to determine each inversion’s frequency in the read pool, instead of incorporating haplotype or additional SNP information.

To assess the significance of the difference in inversion frequency between two sampled life stages in the same experiment, we first calculated a chi-square test statistic from the two-by-two table of inverted and non-inverted allele counts at two life stages. Allele counts were taken by multiplying the inversion-specific SNP frequency in the sequencing reads by the allele sample size. We then generated an approximate distribution of the chi-square statistic when selection is absent (our null hypothesis), using a resampling approach (Figure 6). Comparing the observed statistic to our modeled neutral distribution of test-statistic values allowed us to identify significance thresholds.

To generate the neutral distribution of chi-square test statistics, we took the observed paternal inversion frequencies and then sampled new paternal and offspring read sets to represent a single neutral sample. Ideally, we could sequence from the pool of all fathers in the experiment, but up to 15% of the paternal population died before collection. So, in generating a new paternal read set, we first modeled down-sampling of the paternal allele sample from the initial census size to the collected allele size as a hypergeometric sampling. Then we used a binomial sampling to account for the variation introduced by the extraction and library preparation. The extraction and library preparation appeared to introduce variation independent of allele sample size, possibly due to variation introduced during PCR or sample homogenization. This inference was based on analysis of data from control library preparations (Table S5) and sequencing done in parallel with our experiment from sample fly pools with known inversion frequencies. These control samples included an average of 6 technical replicates (separate library prep and sequencing replicates from the same flies) from four different expected frequencies for In(2L)t, as well as one extraction each for In(2R)NS and In(3R)K. Due to genotyping and husbandry difficulties we did not add a control set for In(3L)Ok. Based on these technical replicates, we estimated that taking a binomial sampling of size 316 from the sample allele frequency reflected a maximum likelihood model for the sample-size-independent variation introduced by the library prep and sequencing process. From the preliminary resampled paternal inversion frequency that now accounted for the library variation, we then took a binomial sample of size equal to the read count to account for sequencing variation, thus yielding resampled paternal read counts that account for individual sample size, library preparation, and depth of coverage. For the new offspring read sets, we instead used a binomial sampling of half the offspring alleles from the fully resampled paternal allele frequency for each separate offspring cohort, since the homozygous mothers all contributed non-inverted alleles. Finally, each offspring allele sample was resampled using the same size 316 binomial sampling to account for the variation introduced during this library’s preparation, and binomial sampled again with the read count size to account for sampling reads from the library, thus yielding resampled offspring read counts that accounted for the same three factors as for paternal counts.

For a pairwise comparison of inversion frequency between two samples, we calculated the same allele count chi- squared test statistic as described above for the empirical data for each resampled set of allele counts, generating the full distribution of neutral chi-squared test statistics from which we could identify extreme value cutoffs. We then obtained a p-value from the proportion of resampled replicates in which the chi-square statistic was greater than or equal to the empirical value. If the test was directional, we took the proportion of resampled replicates with both a chi-square statistic greater than or equal to the empirical value and a frequency difference in the expected direction.

In order to identify a frequency change reversal between life history stages that could reflect a fitness tradeoff, we tested for the joint significance of two opposing frequency changes using three cohorts. A tradeoff would require a frequency change in one direction between paternal and embryo samples, followed by a frequency change in the other direction between embryo and adult offspring samples. These changes both depend on the shared embryonic cohort in the test, and so are not independent. Therefore, we evaluated the two inversion frequency changes jointly from a set of resampled neutral data sets in which the three cohorts arose from the same resampling process. Here we considered how likely it was that a sample of the three cohorts from the associated estimated neutral model had chi-square values for both the comparison of the fathers to embryos and the embryos to combined adults that were equivalent or more extreme in the test direction than the observed data.

For example, if we were testing the significance of an observed paternal-embryo increase and then embryo-adult decrease in frequency, our p-value would be defined by the proportion of resampled replicates in which the frequency changes were both in these same directions, and their chi-square values were both more extreme than the empirical values.

To assess the significance of inversion frequency changes considered across all four maternal line crosses at once, we used Fisher’s combined probability test. We applied this to both particular three-cohort frequency change reversals and to the set of two-cohort frequency changes. The inversion trajectory is independent between maternal crosses, but the differences between each cross in maternal line and initial paternal genetic variation make the replicates difficult to compare in a more structured way, so we felt this was an appropriate test for combined significance. In assessing our tradeoff hypothesis, we applied this test only for inversions where it was well-motivated; i.e. where the averaged frequency change across maternal crosses was in opposite directions between paternal/embryo and embryo/adult comparisons.

To account for multiple hypothesis testing in our cohort comparisons, we used a Benjamini-Yekutieli false discovery rate correction for significance. We applied this correction across the set of Fisher’s method combined p- values we generated to assess the relevant frequency change reversal hypotheses. So, this corrected for the multiple hypotheses of different tradeoff directions and different inversions involved, from p-values that already combined evidence from different maternal line crosses. Benjamini-Yekutieli was chosen due to its robustness to the unclear dependence relationships between some of the tests involved. For example, multiple tests of frequency changes from a single cohort (e.g. an embryo sample compared against both paternal and aged adult samples) are likely to be positively correlated.

Supporting information

eLife Assessment

This valuable study introduces a new simulation model to explain the wide-spread occurrence of genetic inversions in fruit-fly genomes, based on sexually antagonistic alleles and a trade-off between male reproduction and survival. The evidence supporting the conclusions is currently incomplete, but it might be possible to address this with additional simulations and experiments as well as more rigorous analysis of the model and the data. The work will be of interest to population geneticists beyond the fruit-fly system.

Reviewer #1 (Public Review):

The idea is that inversions capture genetic variants that have antagonistic effects on male sexual success (via some display traits) and survival of females (or both sexes) until reproduction. A series of simulations are presented and show that the scenario works at least under some conditions. While a polymorphism at a single locus with large antagonistic effects can be maintained for a certain range of parameters, a second such variant with somewhat smaller effects tends to be lost unless closely linked. It becomes much more likely for genomically distant variants that add to the antagonism to spread if they get trapped in an inversion; the model predicts this should drive the accumulation of sexually antagonistic variants on the inversion versus standard haplotype, leading to the evolution of haplotypes with very strong cumulative antagonistic pleiotropic effects. This idea has some analogies with one of the predominant hypotheses for the evolution of sex chromosomes, and the authors discuss these similarities. To provide empirical support for this idea, the authors study the dynamics of inversions in population cages over one generation, tracking their frequencies through amplicon sequencing, from the parental generation through embryos to aged adults of either sex. Out of four inversions included in the experiment, two show patterns consistent with antagonistic effects on male sexual success (competitive paternity) and the survival of offspring, especially females, until old age, which the authors interpret as consistent with their theory.

This is an interesting idea, and the authors should be praised for combining a model with experimental data. However, in addition to the potential for improvement of presentation (details below), the study has some substantial weaknesses that could be addressed with additional simulations and additional experiments.

(1) The authors claim that the negative frequency dependence that maintains polymorphism in their model results from a non-linear relationship between the display trait and sexual success. I am not convinced about that. It seems to me that the "best of n" female choice implemented in the model (l. 741ff and Figure 2) does not lead to negative frequency dependence. Let p be the frequency of the competitively inferior male genotype. Assuming no noise in the male display, a female will mate with an inferior male only if all males among the n males sampled by the female are of the inferior genotype, which will be the fraction p^n, the remaining 1-p^n matings will go to the superior males. Thus, per capita, the inferior males will achieve (p^n)/p or p^(n-1) matings while the per-capita matings per superior male will be (1-p^n)/(1-p). Thus, the ratio of the mating success of the inferior to the superior males will be (1-p) p^(n-1) / (1- p^n). For the range of p from 0 to 1, this is an increasing function of p. E.g., with n = 2, the sexual fitness of the inferior genotype relative to that of the superior phenotype is p/(1+p). Thus, at least in the absence of noise in the mate choice, this generates positive rather than negative frequency dependence. Maybe I missed something, but the authors do not provide support for their claim about the negative frequency-dependence of sexual selection in their simulations. To do so they could (1) extract the relationship between the relative mating success of the two male types from the simulations and (2) demonstrate that polymorphism is not maintained if the relationship between male display trait and mating success is linear.

(2) The authors only explore versions of the model where the survival costs are paid by females or by both sexes. We do not know if polymorphism would be maintained or not if the survival cost only affected males, and thus if sexual antagonism is crucial.

(3) The authors assume no cost to aneuploidy, with no justification. Biologically, investment in aneuploid eggs would not be recoverable by Drosophila females and thus would potentially act against inversions when they are rare.

(4) The authors appear to define balanced polymorphism as a situation in which the average allele frequency from multiple simulation runs is intermediate between zero and one (e.g., Figure 3). However, a situation where 50% of simulation runs end up with the fixation of allele A and the rest with the fixation of allele B (average frequency of 0.5) is not a balanced polymorphism. The conditions for balanced polymorphism require that selection favors either variant when it is rare.

(5) Possibly the most striking result of the experiment is the fact that for 14 out of 16 combinations of inversion x maternal background, the changes in allele frequencies between embryo and adult appear greater in magnitude in females than in males irrespective of the direction of change, being the same in the remaining two combinations. The authors interpret this as consistent with sexually antagonistic pleiotropy in the case of In(3L)Ok and In(3R)K. The frequencies of adult inversion frequencies were, however, measured at the age of 2 months, at which point 80% of flies had died. For all we know, this may have been 90% of females and 70% of males that died at this point. If so, it might well be that the effects of inversion on longevity do not systematically differ between the ages and the difference in Figure 9B results from the fact that the sample includes 30% longest-lived males and 10% longest-lived females.

(6) Irrespective of the above problem, survival until the age of 2 months is arguably irrelevant from the viewpoint of fitness consequences and thus maintenance of inversion polymorphism in nature. It would seem that trade-offs in egg-to-adult survival (as assumed in the model), female fecundity, and possibly traits such as females resistance to male harm would be much more relevant to the maintenance of inversion polymorphisms.

(7) The experiment is rather minimalistic in size, with four cages in total; given that each cage contains a different female strain, it essentially means N=1. The lack of replication makes statements like " In(2L)t and In(2R)NS each showed elevated survival with all maternal strains except ZI418N" (l. 493) unsubstantiated because the claimed special effect of ZI418N is based on a single cage subject to genetic drift and sampling error. The same applies to statements on inversion x female background interaction (e.g., l. 550), as this is inseparable from residual variation. It is fortunate that the most interesting effects appear largely consistent across the cages/female backgrounds. Still, I am wondering why more replicates had not been included.

Reviewer #2 (Public Review):

Summary:

In their manuscript, the authors address the question of whether the inversion polymorphism in D. melanogaster can be explained by sexually antagonistic selection. They designed a new simulation tool to perform computer simulations, which confirmed their hypothesis. They also show a tradeoff between male reproduction and survival. Furthermore, some inversions display sex-specific survival.

Strengths:

It is an interesting idea on how chromosomal inversions may be maintained

Weaknesses:

General points:

The manuscript lacks clarity of writing. It is impossible to fully grasp what the authors did in this study and how they reached their conclusions. Therefore, I will highlight some cases that I found problematic.

Although this is an interesting idea, it clearly cannot explain the apparent influence of seasonal and clinal variation on inversion frequencies.

Specific points:

The simulations are highly specific and make very strong assumptions, which are not well-justified.

Reviewer #3 (Public Review):

Summary:

In this study, McAllester and Pool develop a new simulation model to explain the maintenance of balanced inversion polymorphism, based on (sexually) antagonistic alleles and a trade-off between male reproduction and survival (in females or both sexes). In support of the plausibility of this model, the authors use laboratory experiments on four naturally occurring inversion polymorphisms in Drosophila melanogaster, finding evidence for the existence of the above-mentioned trade-off in two out of the four cases.

Strengths:

1. The study develops and analyzes a new (Drosophila melanogaster-inspired) model for the maintenance of balanced inversion polymorphism, combining elements of (sexually) antagonistically (pleiotropic) alleles, negative frequency-dependent selection, and synergistic epistasis. To this end, the authors developed and used a new simulator (although it was not 100% clear as to why SLiM could not have been used as SLiM has been used to study inversions).

2. The above-mentioned model assumes, as a specific example, a trade-off between male reproductive display and survival; in the second part of their study, the authors perform laboratory experiments on four common D. melanogaster inversions to study whether these polymorphisms may be subject to such a trade-off. The authors find that two of the four inversions show suggestive evidence that is consistent with a trade-off between male reproduction and survival. The new amplicon sequencing approach to track inversion frequencies used by the authors seems promising in terms of studying fitness effects/trade-offs associated with polymorphic inversions and how such effects play out dynamically.

Weaknesses:

1. Mechanisms of balancing selection maintaining balanced inversion polymorphism. In Section 1.1 a better and more accurate overview of the different selective mechanisms that might contribute to the maintenance of balanced inversion polymorphisms should be given (for a recent review see Berdan et al. 2023). For example, negative frequency-dependent selection (NFDS), spatially and temporally varying selection are not mentioned here and are brought up only later, which is not really ideal. While I agree that in most cases our understanding of balanced inversion polymorphism is very limited, there are many empirical examples of these and other mechanisms of balancing selection being at play (e.g., NFDS: Wright & Dobzhansky 1946; Nassar et al. 1973; Álvarez-Castro and Álvarez 2005; Jay et al. 2021; and many examples of evidence for other mechanisms as well). Thus, while the prevalence of inversion polymorphisms is indeed in many cases "enigmatic", the reader should not be given the impression that we do not have yet any empirical evidence for specific mechanisms in particular cases. Similarly, the authors mention the classical (essentially Dobzhansky's) scenario of epistatically interacting loci only in passing, even though this coadaptation scenario may be simpler than the local adaptation mechanism of Kirkpatrick & Barton (2016): in its simplest form, epistatic coadaptation does not require any migration load or locally adaptive alleles à la Kirkpatrick & Barton, but just the capture of 2 overdominant loci, with inversion protecting this fittest double heterozygote from recombination load (Charlesworth 1974; also see Charlesworth & Charlesworth 1974; Charlesworth & Flatt 2021; also see discussion in Charlesworth and Barton 2018). On the other hand, the plausibility of the mutational load/associative overdominance (AOD) mechanism (Sturtevant & Mather 1938; Nei et al. 1967; Ohta 1971) seems to be given too much weight: new work by Charlesworth (2023; https://www.biorxiv.org/content/10.1101/2023.10.16.562579v1) suggests that load likely contributes only very modestly to heterokaryotypic advantage of inversions at intermediate frequencies, and that is very unlikely to provide a sufficient selective (heterotic) advantage to new autosomal inversions in order to explain their establishment (also see Nei et al. 1967; Connallon and Olito 2021; Jay et al. 2022).

2. The general reduction principle and inversion polymorphism. In Section 1.2., the authors state that "there has not been a proposed mechanism whereby alleles at multiple linked loci would directly benefit from linkage and thereby maintain an associated inversion polymorphism under indirect selection." Perhaps I am misunderstanding something, but in my reading, this statement is factually incorrect. In fact, the simplest version of Dobzhansky's epistatic coadaptation model (see Charlesworth 1974; also see Charlesworth and Charlesworth 1973 and discussion in Charlesworth & Flatt 2021; Berdan et al. 2023) seems to be an example of exactly what the authors seem to have in mind here: two loci experiencing overdominance, with the double heterozygote possessing the highest fitness (i.,e., 2 loci under epistatic selection, inducing some degree of LD between these loci), with subsequent capture by an inversion; in such a situation, a new inversion might capture a haplotype that is present in excess of random expectation (and which is thus fitter than average). The selective benefits of recombination suppression in the inversion heterokaryotype will then confer a heterozygote advantage to the inversion and prevent it from going to fixation (see Charlesworth 1974). This is probably the simplest (or one of the simplest) models of multilocus balancing selection that can act on inversions. Incidentally, this model represents a prime example of the "reduction principle", which the authors mention on two occasions in their paper: generally, any multi-locus polymorphism held at equilibrium by any type of balancing selection involving fitness epistasis will cause selection for reduced recombination (e.g., Feldman & Liberman, 1986; Zhivotovsky et al., 1994); notably, the example of inversion polymorphism is explicitly discussed in Altenberg's and Feldman's (1987) paper on the reduction principle. It is also noteworthy in this context that the 2-locus epistatic model of Charlesworth (1974) assumes constant fitness values/selection coefficients but actually leads to what one could call "apparent" frequency-dependent selection with different equilibria.

3. Trade-offs and antagonistic pleiotropy involved in maintaining inversions. Throughout the manuscript, previous work implicating trade-offs and/or antagonistic pleiotropy (AP) in the maintenance of inversion polymorphisms should be more adequately acknowledged and discussed - from the text as it currently stands one gains the impression that barely anything is known about the connection between inversions and pleiotropy/AP/trade-offs (e.g., Betrán et al. 1998; Mérot et al. 2020, which is cited but not really in the context of AP/trade-offs; Pei et al. 2023, etc.). The paper by Pei and colleagues is particularly relevant in the context of the present study: the authors find that the inverted allele has beneficial effects on male siring success and female fecundity but negative effects on survival. Generally, numerous studies have found that inversion polymorphisms have "pleiotropic" (albeit not always antagonistically pleiotropic) effects upon multiple fitness components (e.g., Etges 1989; Betrán et al. 1998; Küpper et al. 2016; Durmaz et al. 2018; Mérot et al. 2020). More broadly, the general role of AP in maintaining (life-history) polymorphisms should be mentioned by referring to previous theories (e.g., Rose 1982, 1985; Curtsinger et al. 1994; Charlesworth & Hughes 2000 chapter in Lewontin Festschrift; Conallon & Chenoweth 2019 - this latter paper is particularly relevant in terms of AP effects in the context of sexual antagonism).

4. Sexually antagonistic selection and inversion polymorphism. The authors' model of sexual antagonism being involved in maintaining an inversion polymorphism is novel and interesting, but again I felt that the authors' ideas could be better connected to what has been done before. First, several papers have made connections between sexual antagonism and inversion polymorphisms: for example, an important study that deserves discussion in this context is the paper by Natri and colleagues (2019) where the authors study sexual antagonism as a source of balancing selection that maintains an inversion polymorphism in the ruff. Similarly, another relevant study in this context is Hearn et al. 2022 on Littorina saxatilis snails. Also see Giraldo-Deck et al. (2022). A very interesting paper that may be worth discussing is Connallon & Chenoweth (2019) about dominance reversals of antagonistically selected alleles (even though C&C do not discuss inversions): AP alleles (with dominance reversals) affecting two or more life-history traits provide one example of such antagonistically selected alleles (also see Rose 1982, 1985; Curtsinger et al. 1994) and sexually antagonistically selected alleles provide another. The two are of course not necessarily mutually exclusive, thus making a conceptual connection to what the authors model here.

5. The model. In general, the description of the model and of the simulation results was somewhat hard to follow and vague. There are several aspects that could be improved: (1) it would help the reader if the terminology and distinction of inverted vs. standard arrangements and of the three karyotypes would be used throughout, wherever appropriate. (2) The mention of haploid populations/situations and haploid loci (e.g., legend to Figure 1) is somewhat confusing: the mechanism modelled here, of course, requires suppressed recombination in the inversion/standard heterokaryotype; and thus, while it may make sense to speak of haplotypes, we're dealing with an inherently diploid situation. (3) The authors have a situation in mind where the 2 karyotypes (INV vs. STD) in the heterokaryotype carry distinct sets of loci in LD with each other, with one karyotype/haplotype carrying antagonistic variants favoring high male display success and with the other karyotype/haplotype carrying non-antagonistic alternative alleles at these loci and which favor survival. Thus, at each of the linked loci, we have antagonistic alleles and non-antagonistic alleles - however, the authors don't mention or discuss the degree of dominance of these alleles. The degree of dominance of the alleles could be an important consideration, and I found it curious that this was not mentioned (or, for that matter, examined). (4) In many cases, the authors do not provide sufficient detail (in the main text and the main figures) about which parameter values they used for simulations; the same is true for the Materials & Methods section that describes the simulations. Conversely, when the text does mention specific values (e.g., 20N generations, 0.22-0.25M, etc.), little or no clear context or justification is being provided. (5) The authors sometimes refer to "inversion mutation(s)" - the meaning of this terminology is rather ambiguous.

6. Throughout the manuscript, especially in the description and the discussion of the model and simulations, a clearer conceptual distinction between initial "capture" and subsequent accumulation / "gain" of variants by an inversion should be made. This distinction is important in terms of understanding the initial establishment of an inversion polymorphism and its subsequent short- as well as long-term fate. For example, it is clear from the model/simulations that an inversion accumulates (sexually) antagonistic variants over time - but barely anything is said about the initial capture of such loci by a new inversion.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

[…]

(1) The authors claim that the negative frequency dependence that maintains polymorphism in their model results from a non-linear relationship between the display trait and sexual success [...] Maybe I missed something, but the authors do not provide support for their claim about the negative frequency-dependence of sexual selection in their simulations. To do so they could (1) extract the relationship between the relative mating success of the two male types from the simulations and (2) demonstrate that polymorphism is not maintained if the relationship between male display trait and mating success is linear.

We believe that there is a confusion of terminology here. We agree that for the two alleles at a locus impacting male display in our model, the allele conferring inferior display quality will have a fitness that increases as its frequency increases, so this allele displays positive frequency dependent fitness. And, the alternate, display-favoring allele at the locus does display negative frequency dependence. Our use of the terminology ‘negative frequency dependence’ was meant to refer to the negative dependence of the fitness of the display-favoring allele with respect to its own frequency. However, a significant body of literature instead discusses models in which both an allele and its alternate(s) are beneficial when at low frequency and deleterious when at high frequency under the same selective challenge, entailing negative frequency dependence of fitness for all alleles involved. This benefit-when-rare model of a single trait is often described simply as negative frequency dependence, and generates balancing selection at the locus, but is not the model we are presenting here, and does not encompass all models involving negative frequency dependent fitness. This lexical expectation may make the interpretation of our work more difficult, and we have amended the manuscript to make our model clearer (lines 227-231). In this model, we have a negative frequency dependence for the fitness of the display-favoring allele in mate competition, but the net selective disadvantage of this allele at high frequency is due to a cost in another, pleiotropic, fitness challenge: the constant survival effect. So, the alleles are under balancing selection where alternate alleles are favored by selection when rare, but not due solely to selection during mate competition. Instead, our model relies on pleiotropy for an emergent form of frequency-dependent balancing selection (in the sense that each allele is predicted to be beneficial on balance when rare).

In the reviewer’s model of the success of two alleles at one locus, the ratio of success is vaguely linear with allele frequency for n=3, though it starts quite convex and has an inflection point between convex and concave segments (for the disfavored allele) at p≈0.532. This is visualized easily by plotting the function and its derivatives in Wolfram-Alpha. For n>=4, the fitness function with respect to the display-favoring/disfavoring allele becomes increasingly concave/convex respectively, and this specific nonlinearity is needed to act along with the antagonistic pleiotropy to maintain balancing selection, rather than being maintained by a model that favors any rare allele on the basis of its rarity in some manner. In an attempt to make the importance of the encounter number parameter clearer, we’ve generated new panels for Figure S1 which simulate encounter numbers 2, 3, and 4, and we have updated corresponding text and figure references in lines 335-338.

For (1-2), it is not clear how to modify the simulation such that the relationship between the trait value and mating success can be perfectly linear - either linear with respect to allele frequency in a one locus model or linear with respect to trait value at a specific population composition, without removing the simulation of mate competition altogether. While it may be of interest to explore a more comprehensive range of biological trade-offs in future studies, we are not able to meaningfully do so within the context of the present manuscript.

(2) The authors only explore versions of the model where the survival costs are paid by females or by both sexes. We do not know if polymorphism would be maintained or not if the survival cost only affected males, and thus if sexual antagonism is crucial.

We now present simulations with male costs only as added panels to Figure S1 and mention these results in the main text (lines 334-335). Maintenance of the polymorphism is significantly reduced or completely absent in such simulations.

(3) The authors assume no cost to aneuploidy, with no justification. Biologically, investment in aneuploid eggs would not be recoverable by Drosophila females and thus would potentially act against inversions when they are rare.

We did offer some discussion and justification of our decision to model no inherent fitness of the inversion mutation itself, specifically aneuploidy, in lines 36-39 and 78-80 of the original reviewed preprint. Previous research suggests that D. melanogaster females may not actually invest in aneuploid eggs generated from crossover within paracentric inversions. While surprising, and potentially limited to a subset of clades, many ‘r-selected’ taxa or those in which maternal investment is spread out over time may have some degree of reproductive compensation for non-viable offspring, which can reduce the costs of generating aneuploids significantly (for example, t-haplotypes in mice). We have added this example and citation to lines 34ff in the current draft.

(4) The authors appear to define balanced polymorphism as a situation in which the average allele frequency from multiple simulation runs is intermediate between zero and one (e.g., Figure 3). However, a situation where 50% of simulation runs end up with the fixation of allele A and the rest with the fixation of allele B (average frequency of 0.5) is not a balanced polymorphism. The conditions for balanced polymorphism require that selection favors either variant when it is rare.

We originally chose mean final frequency for presenting the single locus simulations based on the ease of generating a visual plot that included information on fixation vs loss and equilibrium frequency. Figure 3 and related supplemental images have been changed to now also represent the proportion of simulations retaining polymorphism at the locus in the final generation.

(5) Possibly the most striking result of the experiment is the fact that for 14 out of 16 combinations of inversion x maternal background, the changes in allele frequencies between embryo and adult appear greater in magnitude in females than in males irrespective of the direction of change, being the same in the remaining two combinations. The authors interpret this as consistent with sexually antagonistic pleiotropy in the case of In(3L)Ok and In(3R)K. The frequencies of adult inversion frequencies were, however, measured at the age of 2 months, at which point 80% of flies had died. For all we know, this may have been 90% of females and 70% of males that died at this point. If so, it might well be that the effects of inversion on longevity do not systematically differ between the ages and the difference in Figure 9B results from the fact that the sample includes 30% longest-lived males and 10% longest-lived females.

This critique deserves some consideration. The aging adults were separated by sex during aging, but while we recorded the number of survivors, we did not record the numbers of eclosed adults and their sexes initially collected out of an interest in maintaining high throughput collection. We therefore cannot directly calculate the associated survival proportions, but we can estimate them. We collected 1960 females and 3156 males, and we can very roughly estimate survival if we assume that equal numbers of each sex eclosed, and that the survivors represent 20% of the original population. That gives 12790 individuals per sex, or 84.7% female mortality and 75.3% male mortality.

So, we have added a qualification discussing the possibility of stronger selection on females and its influence on observed sex-specific frequency changes, on lines 602-605.

(6) Irrespective of the above problem, survival until the age of 2 months is arguably irrelevant from the viewpoint of fitness consequences and thus maintenance of inversion polymorphism in nature. It would seem that trade-offs in egg-to-adult survival (as assumed in the model), female fecundity, and possibly traits such as females resistance to male harm would be much more relevant to the maintenance of inversion polymorphisms.

Adult Drosophila will continue to reproduce in good conditions until mortality, and the estimated age of a mean reproductive event for a Drosophila melanogaster individual is 24 days (Pool 2015), and likewise for D. simulans (Turelli and Hoffman 1995). Given that reproduction is centered around 24 days, we expect sampling at 2 months of age to still be relevant to fitness. In seasonally varying climates, either temperate or with long dry season, survival through challenging conditions is expected to require several months. In many such cases, females are in reproductive diapause, and so longevity is the main selective pressure. See lines 931-936 in the revised manuscript.

As we agreed above, it would of interest to investigate a wider range of trade-offs in future studies. We focused here on the balanced between survival and male reproductive success because the latter trait generates negative frequency dependence for display-favoring alleles and a disproportionate skew towards higher quality competitors, whereas many other fitness-relevant traits lack that property.

(7) The experiment is rather minimalistic in size, with four cages in total; given that each cage contains a different female strain, it essentially means N=1. The lack of replication makes statements like " In(2L)t and In(2R)NS each showed elevated survival with all maternal strains except ZI418N" (l. 493) unsubstantiated because the claimed special effect of ZI418N is based on a single cage subject to genetic drift and sampling error. The same applies to statements on inversion x female background interac7on (e.g., l. 550), as this is inseparable from residual variation. It is fortunate that the most interesting effects appear largely consistent across the cages/female backgrounds. Still, I am wondering why more replicates had not been included.

Our experimental approach might be described as “diversity replication”. Essentially, the four maternal genetic backgrounds are serving dual purposes – both to assess experimental consistency and to ensure that our conclusions are not solely driven by a single non-representative genotype (which in so many published studies, can not be ruled out). It would indeed be interesting if we could have quadrupled the size of our experiment by having four replicates per maternal background. However, we suspect the reviewer may not recognize the substantial effort involved in our four existing experiments. Each of these involved collecting 500+ virgin females, hand-picking thousands of embryos during the duration of egg-laying, and repeatedly transferring offspring to maintain conditions during aging, such that cages had to be staggered by more than a month. These four cages took a year of benchwork just to collect frozen samples, before any preparation and quality control of the associated amplicon libraries for sequencing. Adding a further multiplier would take it well beyond the scope of a single PhD thesis. Fortunately, we were able to obtain the key results of interest without that additional effort, even if clearer insights into the role of maternal background would also be of strong interest.

We do agree that no firm conclusions about maternal background can be reached without further replication, and so we have qualified or removed relevant statements accordingly (lines 568ff, 620-622).

Reviewer #1 (Recommendations For The Authors):

The description of the model is confusing and incomplete, e.g., the values of several parameters used to obtain the numerical results are not given. It is first stated (l. 223) that the model is haploid, but text elsewhere talks about homozygotes and heterozygotes. If the model is diploid (this in itself is not clear), what is assumed about dominance?

We are not presenting results for a mathematical model estimated numerically. We have now clarified our transition from a conceptual depiction of our model, in which we use haploid representations for simplified presentation, to our forward population genetic simulations, which are entirely diploid. More broadly, we have improved our communication of the assumptions and parameters used in our simulations. The scenarios we investigate involve purely additive trait effects within and between loci (except that survival probabilities are multiplicative to avoid negative values). We think that considering other dominance scenarios would be a worthy subject for a follow-up study, whereas the present manuscript is already covering a great deal of ground.

Similarly, it is hard to understand the design (l.442ff). I was confused as to whether a population was set up for each inversion or for all of them and what the unit or replication was. I found the description in Methods (l. 763-771) much clearer and only slightly longer; I suggest the authors transfer it to the Results. Also, Figure 8 should contain the entire crossing scheme; the current version is misleading in that it implies males with only two genotypes.

All four tested inversions were segregating within the same karyotypically diverse population of males, and were assayed from the same experiments. We have attempted to improve the relevant description. For Figure 8, we had trouble conceiving a graphic update that contained a more complete cross scheme without seeming much more confused and cluttered. We have tried to clarify in the relevant text and the figure caption instead.

There are a number of small issues that should be addressed:

- No epistasis for viability assumed - what would be the consequence?

We explored a model in which we intentionally included no terms for epistatic effects on phenotype. All epistasis with regard to fitness is emergent from competition between individuals with phenotypes composed of non-epistatic, non-dominant genetic effects. So, the simplest model of antagonism would have no epistasis for viability whatsoever. One could explore a model that has emergent viability epistasis in a similar way, by implementing stabilizing selection on a quantitative trait with a gaussian or similar non-linear phenotype-to-fitness map, but that might be better served as a topic for a future study. We have, however, tried to make this intent clearer in the text.

l. 750 implies that aneuploidy generated by the inversion has no cost (aneuploid games are resampled)

Yes, as addressed in public review item (3). Alternately see lines 34ff, 293, 369, 392 for in-text edits.

l. 24-25: unclear; is this to mean that there is haplotype x sex interaction for survival?

l. 25: success in what? (I assume this will be explained in the paper, but the abstract should stand on its own).

l. 193-4: "producing among most competitive males": something missing or a word too much?? Figure 1B,C: a tiny detail, but the plots would be more intuitive if the blue (average) bars were ager (i.e., to the right) of the male and female ones, given that the average is derived from the two sex-specific values.

Each of the above have been edited or implemented as suggested

l. 205. It is convex function, but I do not understand what the authors mean by "convex distribution".

Hopefully the updated text is clearer: “yielding a distribution of male reproductive output that follows a relatively convex trend”.

l. 223ff: some references to Fig 1 panels in this paragraph seem off by one letter (i.e., A should be B, etc.).

l. 231 "fitness...are equally fit": rephrase

l. 260: maybe "thrown out" is not the most fortunate term, maybe "eliminated" would be better?

Each of the above have been edited or implemented as suggested

Figure 3: I do not understand the meaning of "additive" and "multiplicative" in the case of a single locus haploid model

All presented simulations are diploid, and these refer to the interactions between the two alleles at the locus. Hopefully the language is overall clearer in this draft.

l. 274: "Mutation of new nucleotide" meaning what? Or is it mutation _to_ a new nucleotide?

Hopefully the revised text is clearer.

Figure 5. The right panel of figure 5A implies that, with the inversion, the population evolves to an extreme display trait that is so costly that it fills 95% of all individuals (or of all females?

What is assumed about this here?). Apart from the biological realism of this result, what does it say about the accumulation of polymorphism and maintenance of the inversion? The graphs in fig 5B do plot a divergence between haplotypes, but it is not clear how they relate to those in panel A - the parameter values used to generate these plots are again not listed. Furthermore, from the viewpoint of the polymorphism, it would be good to report the frequencies at the steady-state.

We have now clarified the figure description, including the parameter values used. The distribution of frequencies at the end of the simulation is represented in figure 6. Given that we set up the simulation with assumptions that are otherwise common to population models, what biological process would prevent this extreme? Why isn’t this extreme observed in natural populations? One possible explanation is that they become sex chromosomes, with increasing likelihood as the cost increases. Or other compensatory changes may occur that we don’t simulate, like regulatory evolution giving a complementary phenotype. Maybe genetic constraints in natural populations prevent the mutation of the kind of pleiotropic mutations that drive this dynamic. The populations still survive, though they are parameterized by relative fitness. What would an absolute fitness population function be? Would it go extinct or not? It would be of interest to explore a wider range of models, but it is the purpose of this paper to establish that this is a viable model for the maintenance of sexually antagonistic polymorphism and association with inversions. We have added a paragraph motivated by this comment to the Discussion starting on line 765.

l. 401-2: Z-like, W-like : please specify you are talking about patterns resembling sex chromosomes.

l. 738: "population calculates"?

l. 743-4 and 746-7: is this the same thing said twice, or are there two components of noise? l. 357: there is no figure 5C.

Each of the above have been addressed with text edits.

L. 473-5: Yes, the offspring did not contain inversion homozygotes, but the sire pool did, didn't it? So homozygous inversions may have affected male reproductive success. Anyway, most of this paragraph (from line 473) seems to belong in Discussion rather than Results.

We have revised this sentence to focus on offspring survival.

We can understand the reviewer’s suggestion about Results vs. Discussion text. While this can often be a challenging balance, we find that papers are often clearer if some initial interpretation is offered within the Results text. However, we moved the portion of this paragraph relating our findings to the published literature to the Discussion.

l. 516: " In(3L)Ok favored male survival": this is misleading/confusing given the data, " In(3L)Ok reduced female survival more strongly than male survival..."

Hopefully the phrasing is clearer now.

l. 663ff: I did not have an impression that this section added anything new and could safely be cut.

We have done some editing to make this more concise and emphasize what we think is essential, but we believe that the model of an autosomal, sexually antagonistic inversion differentiating before contributing to the origin of a sex chromosome is novel and interesting. And, that this additional emphasis is worthwhile to encourage thought and consideration of this idea in future research and among interested researchers.

l. 751: "flat probability per locus": do the authors mean a constant probability?

Edited.

Reviewer #2 (Public Review):

The manuscript lacks clarity of writing. It is impossible to fully grasp what the authors did in this study and how they reached their conclusions. Therefore, I will highlight some cases that I found problematic.

Hopefully the revised manuscript improves writing clarity.

Although this is an interesting idea, it clearly cannot explain the apparent influence of seasonal and clinal variation on inversion frequencies.

We do not believe that our model predicts a non-existence of temporal and spatial dependence of the fitness of inverted haplotypes, nor do we seek to identify the manner in which seasonal and clinal differences affect fitness of inverted haplotypes. Rather, we argued that the influence of seasonal and clinal selection on inversions does not on its own predict the observed maintenance of inversions at low to intermediate frequencies across such a diverse geographic range, along with the higher frequencies of many derived inversions in more ancestral environments.

We might imagine that trade-offs between life history traits such as mate competition and survival should be universal across the range of an organism. But in practice, the fitness benefits and costs of a pleiotropic variant (or haplotype) may be heavily dependent on the environment. A harsh environment such as a temperate winter may both reduce the number of females that a male encounters (decreasing the benefit of display-enhancing variants) and also increase the likelihood that survival-costly variants lead to mortality (thus increasing their survival penalty). In light of such dynamics, our model would predict that equilibrium inversion frequencies should be spatially and temporally variable, in agreement with a number of empirical observations regarding D. melanogaster inversions.

We have edited the introduction to emphasize that inversion frequencies vary temporally as well as seasonally, on lines 144ff. We also note relevant discussion of the potential interplay between the environment and trade-offs such as those we investigate, on lines 153-155.

The simulations are highly specific and make very strong assumptions, which are not well-justified.

We respond to all specific concerns expressed in the Recommendations For The Authors section below. We also note that we have made further clarifications throughout the text regarding the assumptions made in our analysis and their justification.

Reviewer #2 (Recommendations For The Authors):

I think that the manuscript would greatly benefit from a major rewrite and probably also a reanalysis of the empirical data.

In particular, a genome-wide analysis of differences in SNP frequencies between sexes and developmental stages would help the reader to appreciate that inversions are special.

[moved up within this section for clarity] We are lacking a genomic null model-how often do the authors see similar allele frequency differences when looking at the entire genome? This could be easily done with whole genome Pool-Seq and would tell us whether inversions are really different from the genomic background. I think that this information would be essential given the many uncertainties about the statistical tests performed.

We expect that autosome-wide SNP frequencies will be heavily influenced by the frequencies of inversions, which occur on all four major autosomal chromosome arms. These inversions often show moderate disequilibrium with distant variants (e.g. Corbett-Detig & Hartl 2012).

Furthermore, the limited number of haplotypes present, given that the paternal population was founded from 10 inbred lines, would further enhance associations between inversions and distant variants. Therefore, we do not expect that whole-genome Pool-Seq data would provide an appropriate empirical null distribution for frequency changes. Instead, we have generated appropriate null predictions by accounting for both sampling effects and experimental variance, and we have aimed to make this methodology clearer in the current draft.

Some basic questions:

why start at a frequency of 50% (line 287)?

Isn't it obvious that in this scenario strong alleles with sexually antagonistic effects can survive?

The initial goal of the associated Figure 4 was not to show that a strongly antagonistic variant could persist. Instead, we wanted to test the linkage conditions in which a second, relatively weaker antagonistic variant survived – which did not occur in the absence of strong linkage.

We have now added simulations with relatively lower initial frequencies, in which the weaker variant and the inversion both start at 0.05 frequency, while the stronger variant is still initialized at 0.5 to reflect the initial presence of one balanced locus with a strongly antagonistic variant. Here, the weaker antagonistic variant is still usually maintained when it is close to the stronger variant, and while the inversion-mediated maintenance of the weaker variant at greater distance from the stronger variant because less frequent than the original investigated case, it still happens often enough to hypothetically allow for such outcomes over evolutionary time-scales.

Still, we should also emphasize that the goals of this proof-of-concept analysis are to establish and convey some basic elements of our model. Subsequently, analyses such as those presented in Figures 5 and 6 provide clearer evidence that the hypothesized dynamics of inversions facilitating the accumulation of sexual antagonism actually occur in our simulations.

The experiments seem to be conducted in replicate (which is of course essential), but I could not find a clear statement of how many replicates were done for each maternal line cross.

How did the authors arrive at 16 binomial trials (line 473)? 4 inversions, 4 maternal genotypes?

How were replicates dealt with?

In Figure 9, it would be important to visualize the variation among replicates.

Unfortunately, we did not have the bandwidth to perform replicates of each maternal line. Instead, we use four maternal backgrounds to simultaneously establish consistency across independent experiments and genetic backgrounds (see our response to Reviewer 1, point 7). We’ve edited the draft to make this clearer and more clearly delineate what is supported and not supported by our data. Replicate variation for the control replicates of the extraction and sequencing process, and the exact read counts of the experiment, are available in Supplemental Tables S5, S6, and S7.

The statistical analysis of trade-off is not clear: which null model was tested? No frequency change? In my opinion, two significances are needed: a significant difference between parental and embryo and then embryo and adult offspring. The issue with this is, however, that the embryo data are used twice and an error in estimating the frequency of the embryos could be easily mistaken as antagonistic selection.

Hopefully the description of our null model is clearer in the text, now starting around line 967 in the Methods. We are aware of the positive dependence when performing tests comparing the paternal to embryo and then embryo to offspring frequencies, and this is accounted for by our analysis strategy - see lines 1009-1012.

It was not clear how the authors adjusted their chi-squared test expectations. Were they reinventing the wheel? There is an improved version of the chi-squared test, which accounts for sampling variation.

We did not actually perform chi-square tests. Instead, we used the chi statistic from the chi-squared test as a quantitative summary of the differences in read counts between samples. We compared an observed value of chi to values for this statistic obtained from simulated replicates of the experiment. Sampling from this simulation generated our ‘expected’ distribution of read counts, sampled to match sources of variance introduced in the experimental procedure, but without any effect of natural selection, per lines 825ff in the original submission. Hence, we are approximating the likelihood of observing an empirical chi statistic by generating random draws from a model of the experiment and comparing values calculated from each draw to the experimental value: a Monte Carlo method of approximating a p-value for our data. We have attempted to make the structure of these simulations and their use as a null-model clearer in this draft.

It is not sufficiently motivated why the authors model differences in the extraction procedure with a binomial distribution.

Adding a source of variance here seemed necessary as running control sequencing replicates revealed that there was residual variance not fully recapitulated by sample-size-dependent resampling. Given that we were still sampling a number of draws from a binomial outcome (the read being from the inverted or standard arrangement), a binomial distribution seemed a reasonable model, and we fit the level of this additional noise source to an experiment-wide constant, read-count or genome-count independent parameter that best fit the variance observed in the controls (lines 830ff in the original draft). Clarification is made in this manuscript draft, lines 979-989.

How many reads were obtained from each amplicon? It looks like the authors tried to mimic differences between technical replicates by a binomial distribution, which matches the noise for a given sample size, but this depends on the sequence coverage of the technical replicates.

We provide read counts in Supplemental Tables S6 and S7. The relevant paragraph in the methods has been edited for clarity, lines 972ff. Accounting for sampling differences between replicates used a hypergeometric distribution for paternal samples to account for paternal mortality before collection, and the rest were resampled with a binomial distribution. There were two additional binomial samplings, to account for resampling the read counts and to capture further residual variance in the library prep that did not seem to depend on either allele or read counts.

It would be good to see an estimate for the strength of selection: 10% difference in a single generation appears rather high to me.

Estimates of selection strength based on solving for a Wright-Fisher selection coefficient for each tested comparison can now be found in Table S8, mentioned in text on lines 589-590. The mean magnitude of selection coefficients for all paternal to embryo comparisons was 0.322, and for embryo to all adult offspring it was 0.648. For In(3L)Ok the mean selection coefficients were 0.479 and -0.53, and for In(3R)K they were -0.189 and 1.28, respectively. Some are of quite large magnitude, but we emphasize that the coefficients for embryo to adult are based on survival to old age, rather than developmental viability. That factor, in addition to the laboratory environment, makes these estimates distinct from selection coefficients that might be experienced in natural populations.

Reviewer #3 (Public Review):

Strengths:

(1) …the authors developed and used a new simulator (although it was not 100% clear as to why SLiM could not have been used as SLiM has been used to study inversions).

Before SLiM 3.7 or so (and including when we did the bulk of our simulation work), we do not think it would have been feasible to use SLiM to model the mutation of inversions with random breakpoints and recombination between without altering the SLiM internals. Separately, needing to script custom selection, mutation, and recombination functions in Eidos would have slowed SLiM down significantly. Given our greater familiarity with python and numpy, and the ability to implement a similar efficiency simulator more quickly than through learning C++ and Eidos, we chose to write our own.

It should be a fair bit easier to implement comparable simulations in SLiM now, but it will still require scripting custom mutation, selection, and recombination functions and would still result in a similarly slow runtime. The current script recipe recommended by SLiM for simulating inversions uses constants to specify the breakpoints of a single inversion, without the ability to draw multiple inversions from a mutational distribution, or model recombination between more complicated karyotypes. Hence, our simulator still seems to be a more versatile and functional option for the purposes of this study.

Weaknesses:

[Comments 1 through 4 on Weaknesses included numerous citation suggestions, and some discussion recommendations as well. In our revised manuscript, we have substantially implemented these suggestions. In particular, we have deepened our introduction of mechanisms of balancing selection and prior work on inversion polymorphism, integrating many

suggested references. While especially helpful, these suggestions are too extensive to completely quote and respond to in this already-copious document. Therefore, we focus our response on two select topics from these comments, and then proceed to comment 5 thereafter.]

(2) The general reduction principle and inversion polymorphism. In Section 1.2., the authors state that "there has not been a proposed mechanism whereby alleles at multiple linked loci would directly benefit from linkage and thereby maintain an associated inversion polymorphism under indirect selection." Perhaps I am misunderstanding something, but in my reading, this statement is factually incorrect. In fact, the simplest version of Dobzhansky's epistatic coadaptation model

(see Charlesworth 1974; also see Charlesworth and Charlesworth 1973 and discussion in Charlesworth & Flatt 2021; Berdan et al. 2023) seems to be an example of exactly what the authors seem to have in mind here: two loci experiencing overdominance, with the double heterozygote possessing the highest fitness (i.,e., 2 loci under epistatic selection, inducing some degree of LD between these loci), with subsequent capture by an inversion; in such a situation, a new inversion might capture a haplotype that is present in excess of random expectation (and which is thus filer than average)…

We agree that the quoted statement could be misleading and have rewritten it. We intended to point out that we are presenting a model in which all loci contribute additively (with respect to display) or multiplicatively (with respect to survival probability), without any dominance relationships or genetic interaction terms. And yet, the model generates epistatic balancing selection in a panmictic population under a constant environment. This represents a novel mechanism by which (the life-history characteristics of) a population would generate epistatic balancing selection as an emergent property, instead of assuming a priori that there is some balancing mechanism and representing frequency dependence, dominance effects, or epistatic interactions directly using model parameters. We have therefore refined the scope of the statement in question (lines 155-158).

(4) Hearn et al. 2022 on Littorina saxatilis snails.

A good reference. There is considerable work on ecotype-associated inversions in L. saxatalis, but we previously cut some discussion of this and of other populations with high gene flow but identifiable spatial structure for inversion-associated phenotypes (e.g. butterfly mimicry polymorphisms, Mimulus, etc.). Due to the spatially discrete environmental preferences and sampled ranges of the inversions in these populations, we considered these examples to be somewhat distinct from explaining inversion polymorphism in a potentially homogenous and panmictic environment.

(4) cont. A very interesting paper that may be worth discussing is Connallon & Chenoweth (2019) about dominance reversals of antagonistically selected alleles (even though C&C do not discuss inversions): AP alleles (with dominance reversals) affecting two or more life-history traits provide one example of such antagonistically selected alleles (also see Rose 1982, 1985; Curtsinger et al. 1994) and sexually antagonistically selected alleles provide another. The two are of course not necessarily mutually exclusive, thus making a conceptual connection to what the authors model here.

We had removed a previously drafted discussion of dominance reversal for brevity’s sake, but this topic is once again represented in the updated draft of the manuscript with a short reference in the introduction, lines 76-80. We also mention ‘segregation lift’ (Wittmann et al. 2017) involving a similar reversal of dominance for fitness between temporally fluctuating conditions, as opposed to between sexes or life history stages.

(5) The model. In general, the description of the model and of the simulation results was somewhat hard to follow and vague. There are several aspects that could be improved: [5](1) it would help the reader if the terminology and distinction of inverted vs. standard arrangements and of the three karyotypes would be used throughout, wherever appropriate.

We have attempted to do so, using the suggested heterokaryotypic/homokaryotypic terminology.

[5](2) The mention of haploid populations/situations and haploid loci (e.g., legend to Figure 1) is somewhat confusing: the mechanism modelled here, of course, requires suppressed recombination in the inversion/standard heterokaryotype; and thus, while it may make sense to speak of haplotypes, we're dealing with an inherently diploid situation.

While eukaryotes with haploid-dominant life history may still experience similar dynamics, we do expect that most male display competition is in diploid animals, and we are only simulating diploid fitnesses and experimenting with diploid Drosophila. We have tried to minimize the discussion of haploids in this draft.

[5](3) The authors have a situation in mind where the 2 karyotypes (INV vs. STD) in the heterokaryotype carry distinct sets of loci in LD with each other, with one karyotype/haplotype carrying antagonistic variants favoring high male display success and with the other karyotype/haplotype carrying non-antagonistic alternative alleles at these loci and which favor survival. Thus, at each of the linked loci, we have antagonistic alleles and non-antagonistic alleles - however, the authors don't mention or discuss the degree of dominance of these alleles. The degree of dominance of the alleles could be an important consideration, and I found it curious that this was not mentioned (or, for that matter, examined).

In this study, our goal was to show that the investigated model could produce balanced and increasing antagonism without the need to invoke dominance. We think there would be a strong case for a follow-up study that more investigates how dominance and other variables impact the parameter space of balanced antagonism, but this goal is beyond our capacity to pursue in this initial study. We’ve added several lines clarifying the absence of dominance from our investigated models, and pointing out that dominance could modulate the predictions of these models (lines 211-213, 278-282).

[5](4) In many cases, the authors do not provide sufficient detail (in the main text and the main figures) about which parameter values they used for simulations; the same is true for the Materials & Methods section that describes the simulations. Conversely, when the text does mention specific values (e.g., 20N generations, 0.22-0.25M, etc.), little or no clear context or justification is being provided.

We have sought to clarify in this draft that 20N was chosen as an ample time frame to establish equilibrium levels and frequencies of genetic variation under neutrality. We present a time sequence in Figure 5, and these results indicate that that antagonism has stabilized in models without inversions or with higher recombination rates, whereas its rate of increase has slowed in a model with inversions and lower levels of crossing over.

The inversion breakpoints and the position of the locus with stronger antagonistic effects in Figure 4 were chosen arbitrarily for this simple proof of concept demonstration, with the intent that this locus was close to one breakpoint. Hopefully these and other parameters are clearer in the revised manuscript.

[5](5) The authors sometimes refer to "inversion mutation(s)" - the meaning of this terminology is rather ambiguous.

Edited, hopefully the wording is clearer now. The quoted phrase had uniformly referred to the origin of new inversions by a mutagenic process.

(6) Throughout the manuscript, especially in the description and the discussion of the model and simulations, a clearer conceptual distinction between initial "capture" and subsequent accumulation / "gain" of variants by an inversion should be made. This distinction is important in terms of understanding the initial establishment of an inversion polymorphism and its subsequent short- as well as long-term fate. For example, it is clear from the model/simulations that an inversion accumulates (sexually) antagonistic variants over time - but barely anything is said about the initial capture of such loci by a new inversion.

We do not have a good method of assessing a transition between these two phases for the simulations in which both antagonistic alleles and inversions arise stochastically by a mutagenic process. However, we have tried to be clearer on the distinction in this draft: we have included simulations in Figure 4 with variants starting at lower frequencies, and we have tried to better contextualize the temporal trajectories in Figure 5 as (in part) modeling the accumulation of variants after such an origin.

Reviewer #3 (Recommendations For The Authors):

- In general: the whole paper is quite long, and I felt that many parts could be written more clearly and succinctly - the whole manuscript would benefit from shortening, polishing, and making the wording maximally precise. Especially the Introduction (> 8 pages) and Discussion (7.5 pages) sections are quite long, and the description of the model and model results was quite hard to follow.

We have attempted to condense some portions of the manuscript, but inevitably added to others based on important reviewer suggestions. Regarding the length Introduction and Discussion, we are covering a lot of intellectual territory in this study, and we aim to make it accessible to readers with less prior familiarity. At this point, we have well over 100 citations – far more than a typical primary research paper – in part thanks to the relevant sources provided by this reviewer. We are therefore optimistic that our text will provide a valuable reference point for future studies. We have also made significant efforts to clarify the Results and Methods text in this draft without notably expanding these sections.

- In general: the conceptual parts of the paper (introduction, discussion) could be better connected to previous work - this concerns e.g. the theoretical mechanisms of balancing selection that might be involved in maintaining inversions; the general, theoretical role of antagonistic pleiotropy (AP) and trade-offs in maintaining polymorphisms; previously made empirical connections between inversions and AP/trade-offs; previously made empirical connections between inversions and sexual antagonism.

In the revised manuscript, we have improved the connection of these topics to prior work.

- L3: "accumulate". A clearer distinction could be made, throughout, between initial capture of alleles/haplotypes by an inversion vs. subsequent gain.

Please see point 6 in the response to the Public Review, above.

- L29: I basically agree about the enigma, however, there are quite many empirical examples in D. melanogaster / D. pseudoobscura and other species where we do know something about the nature of selection involved, e.g., cases of NFDS, spatially and temporally varying selection, fitness trade-offs, etc.

At least for our focal species, we have emphasized that geographic (and now temporal) associations have been found for some inversions. For the sake of length and focus, we probably should not go down the road of documenting each phenotypic association that has been reported for these inversions, or say too much about specific inversions found in other species. As indicated in our response to reviewer 2, some previously documented inversion-associated trade-offs may be compatible with the model presented here. However, we did locate and add to our Discussion one report of frequency-dependent selection on a D. melanogaster inversion (Nassar et al. 1973).

- L43: it is actually rather unlikely, though not impossible, that new inversions are ever completely neutral (see the review by Berdan et al. 2023).

This line was intended to convey that, in line with Said et al. 2018’s results, the structural alterations involved in common segregating inversions are not expected to contribute significantly to the phenotype and fitness (as indicated by lack of strong regulatory effects), and that their phenotypic consequences are instead due to linked variation. We have rewritten this passage to better communicate this point, now lines 44-52. Interpreting Section 2 and Figure 1 of Berdan et al. 2023, the linked variation may be what is in mind when saying that inversions are almost never neutral. We have also added a line referencing the expected linked variation of a new inversion (lines 49-52).

- L51-73: I felt this overview should be more comprehensive. The model by Kirkpatrick & Barton (2016 ) is in many ways less generic than the one of Charlesworth (1974) which essentially represents one way of modeling Dobzhansky's epistatic coadaptation. Also, the AOD mechanism is perhaps given too much weight here as this mechanism is very unlikely to be able to explain the establishment of a balanced inversion polymorphism (see Charlesworth 2023 preprint on bioRxiv). NFDS, spatially varying selection and temporally varying selection (for all of which there is quite good empirical evidence) should all be mentioned here, including the classical study of Wright and Dobzhansky (1946) which found evidence for NFDS (also see Chevin et al. 2021 in Evol. Lett.)

On reflection, we agree that we put too much emphasis on AOD and have edited the section to be more representative.

- L57. Two earlier Dobzhansky references, about epistatic coadaptation, would be: Dobzhansky, T. (1949). Observations and experiments on natural selection in Drosophila. Hereditas, 35(S1), 210-224. hlps://doi.org/10.1111/j.1601-5223.1949.tb033 34.xM; Dobzhansky, T. (1950). Genetics of natural populations. XIX. Origin of heterosis through natural selection in populations of Drosophila pseudoobscura. Genetics, 35, 288-302.hlps://doi.org/10.1093/gene7cs/35.3.288 - In general, in the introduction, the classical chapter by Lemeunier and Aulard (1992) should be cited as the primary reference and most comprehensive review of D. melanogaster inversion polymorphisms.

- L101: this is of course true, though there are some exceptions, such as In(3R)Mo.

- L110: the papers by Knibb, the chapter by Lemeunier and Aulard (1992), and the meta-analysis of INV frequencies by Kapun & Flatt (2019) could be cited here as well.

Citation suggestions integrated.

- L123 and elsewhere: the common D. melanogaster inversions are old but perhaps not THAT old - if we take the Corbett-Detig & Hartl (2012) es7mates, then most of them do not really exceed an age of Ne generations, or at least not by much. I mean: yes, they are somewhat old but not super-old (cf. discussion in Andolfatto et al. 2001).

Edited to curb any hyperbole. We agree that there are much more ancient polymorphisms in populations.

- L133-135. This needs to be rewritten: this claim is incorrect, to my mind (Charlesworth 1974; also see Charlesworth and Charlesworth 1973; discussion in Charlesworth & Flatt 2021).

Edited. See public review response (2).

- L154: the example of inversion polymorphism is actually explicitly discussed in Altenberg's and Feldman's (1987) paper on the reduction principle.

Edited to mention this. Inversions are also mentioned in Feldman et al. 1980, Feldman and Balkau 1973, Feldman 1972, and have been in discussion since the origins of the idea.

- L162ff: see Connallon & Chenoweth (2019).

Citation suggestion integrated, along with Cox & Calsbeek 2009 which seems more directly applicable, now line 185ff.

- L169: why? There is much evidence for other important trade-offs in this system.

Reworded.

- L178-179: other studies have found that trade-offs/AP contribute to the maintenance of inversion polymorphisms, e.g. Mérot et al. 2020 and Betrán et al. 1998, etc.

Added Betrán et al. 1998 - a good reference. Moved up mention of Mérot et al. 2020 from later in the text and directed readers to the Discussion, lines 202-205.

- L198. "alternate inversion karyotypes" - you mean INV vs. STD? It would be good to adopt a maximally clear, uniform terminology throughout.

Edited to communicate this better.

- L215-217: this is a theoretically well-known result due to Hazel (1943); Dickerson (1955); Robertson (1955); e.g., see the discussion in the quantative genetics book by Roff (1997) or in the review of Flatt (2020).

Citations integrated, now lines 232ff.

- L223 and L245: "haploid" - somewhat confusing (see public review).

- L259-260: This may need some explanation.

- L261-262: simply state that there is no recombination in D. melanogaster males.

Edited for increased clarity.

- L274 (and elsewhere): the meaning of "mutation...of new..inversion polymorphisms" is ambiguous - do you mean a polymorphic inversion and hence a new inversion polymorphism or do you mean polymorphisms/variants accumulating in an inversion?

- L275: maybe better heterokaryotypic instead of heterozygous? (note that INV homokaryotypes or STD homokaryotypes can be homo- or heterozygous, so when referring to chromosomal heterozygotes instead of heterozygous chromosomes it may be best to refer to heterokaryotypes).

Per [5](1) and [5](5) in the public review, we have edited our terminology.

- L276: referral to M&M - I found the description of the model/simulation details there to be somewhat vague, e.g. in terms of parameter settings, etc.

Further described.

- L281-282: would SLiM not have worked?

See public review response.

- L286-287: why these parameters?

Further described.

- L296ff: it is not immediately clear that the loci under consideration are polymorphic for antagonistic alleles vs. non-antagonistic alternative alleles - maybe this could be made clear very explicitly.

Edited to be explicit as suggested.

- L341, 343: "inversion mutation" - meaning ambiguous.

- L348, 352: "specified rate" - vague.

- L354-357: initial capture and/or accumulation/gain?

- L401, 402, 404: Z-, W- and Y- are brought up here without sufficient context/explanation.

The above have been addressed by edits in the text.

- L523, 557, 639, 646, and elsewhere: not the first evidence - see the paper by Mérot et al. (2020) (and e.g. also by Yifan Pei et al. (2023)).

Citations integrated in the introduction and discussion. Mérot et al. (2020) was cited (L486 in original) but discussion was curtailed in the previous draft.

- L558-559. I agree but it is clear that there are many mechanisms of balancing selection that can achieve this, at least in principle; for some of them (NFDS, etc.) we have pretty good evidence.

- L576-577. This is correct but for In(3R)C that study did find a differential hot vs. cold selection response.

Addressed with text edit.

- L584-L586: cf. Betrán et al. (1998), Mérot et al. (2020), Pei et al. (2023), etc.

- L591. "other forms of balancing selection": yes! This should be stressed throughout. Multiple forms of balancing selection exist and they are not mutually exclusive.

- L593: consider adding Dobzhansky (1943), Machado et al. (2021)

- L596-597: this is rather unlikely, at least in terms of inversion establishment (see Charlesworth 2023; hlps://www.biorxiv.org/content/10.1101/2023.10.16.562579v1).

- L608: consider adding Kapun & Flal (2019).

- L611-612: see studies by Mukai & Yamaguchi, 1974; and Watanabe et al., 1976.

- L639, 646: AP - see general literature on AP as a factor in maintaining polymorphism (Rose

1982, 1985; Curtsinger et al. 1994; Charlesworth & Hughes 2000 chapter in Lewontin Festschrift; Conallon & Chenoweth 2019 - this latter paper is par7cularly relevant in terms of AP effects in the context of sexual antagonism)

Citation suggestions integrated.

- L657: inversion polymorphism is explicitly discussed in Altenberg's and Feldman's (1987) paper on the reduction principle.

Hopefully this is better communicated.

- L724-755: I felt that this section generally lacks sufficient details, especially in terms of parameter choices and settings for the simula7ons.

- L732L: why not state these rates?

Parameter values are now given a fuller description in figure legends and in the methods.

- L746: but we know that mutational effect sizes are not uniformly distributed (?).

We made this choice for simplicity and to avoid invoking seemingly arbitrary distribution, but one could instead simulate trait effects with some gamma distribution. Display values would still have variable fitness effects that fluctuate with population composition, but we agree that distribution shifted toward small effects would be more realistic.

- L765: In(3R)P is not mentioned elsewhere - is this really correct?

That was incorrect, fixed.

Version 2

Abstract

Abstract

Chromosomal inversion polymorphisms can be common, but the causes of their persistence are often unclear. We propose a model for the maintenance of inversion polymorphism, which requires that some variants contribute antagonistically to two phenotypes, one of which has negative frequency-dependent fitness. These conditions yield a form of frequency-dependent disruptive selection, favoring two predominant haplotypes segregating alleles that favor opposing antagonistic phenotypes. An inversion associated with one haplotype can reduce the fitness load incurred by generating recombinant offspring, reinforcing its linkage to the haplotype and enabling both haplotypes to accumulate more antagonistic variants than expected otherwise. We develop and apply a forward simulator to examine these dynamics under a tradeoff between survival and male display. These simulations indeed generate inversion-associated haplotypes with opposing sex-specific fitness effects. Antagonism strengthens with time, and can ultimately yield karyotypes at surprisingly predictable frequencies, with striking genotype frequency differences between sexes and between developmental stages. To test whether this model may contribute to well-studied yet enigmatic inversion polymorphisms in Drosophila melanogaster, we track inversion frequencies in laboratory crosses to test whether they influence male reproductive success or survival. We find that two of the four tested inversions show significant evidence for the tradeoff examined, with In(3R)K favoring survival and In(3L)Ok favoring male reproduction. In line with the apparent sex-specific fitness effects implied for both of those inversions, In(3L)Ok was also found to be less costly to the viability and/or longevity of males than females, whereas In(3R)K was more beneficial to female survival. Based on this work, we expect that balancing selection on antagonistically pleiotropic traits may provide a significant and underappreciated contribution to the maintenance of natural inversion polymorphism.

Introduction

The Enigmatic Prevalence of Inversion Polymorphism

Inversions, DNA segments in reversed sequence order relative to the ancestor, have been a focus of population genetic study for their direct or indirect effects on recombination in natural populations for more than a century. In their most dramatic effect on meiosis, inversions generate aneuploid gametes when a crossover occurs between the interior of an inverted region and its non-inverted homolog (White 1973). As aneuploidy is often fatal in embryos or severely deleterious, this may translate to a fecundity cost for a heterokaryotypic parent: a parent heterozygous for an inversion arrangement. However, depending on a species’ reproductive biology, such reproductive fitness effects may be mitigated if the zygotes terminate early or before significant parental resource investment (Charlesworth 1994, Munasinghe and Brandvain 2023). Alternatively, crossovers may be inhibited directly in heterokaryotypes around the inversion by mechanisms similar to crossover interference (Koury 2023), avoided by achiasmy in one sex or all, or aneuploid gametes may be segregated to polar bodies in female meiosis (Sturtevant 1921; Sturtevant and Beadle 1936; Krimbas and Powell 1992; Coyne et al. 1993; Gong et al. 2005). Regardless of the differences between these scenarios, effectively fewer recombinant offspring are produced in heterokaryotypes.

Inversions may also have direct effects on fitness when disrupting a gene, a regulatory element, or other genomic structure like a topologically associated domain (Lupiáñez et al. 2015; McBroome et al. 2020). Given also that a potential effect of inversions in heterokaryotypic individuals is the deleterious generation of aneuploid gametes, it seems reasonable to expect inversion rearrangements to confer either deleterious or neutral fitness consequences in the absence of association with other variants. However, when a new inversion mutates, it samples a single haplotype from a population, which is likely to have some non-neutral fitness effect, even if mild. Greater natural variation in fitness within a population increases the likelihood of sampling haplotypes with exceptionally deleterious or beneficial associations of alleles, even when there are no other effects on fitness (Berdan et al. 2023).

In contrast to the above predictions, a considerable number of well-studied inversion variants are maintained at an intermediate frequency within a population, presumably due to the indirect effect inversions have on reducing recombination and maintaining fit haplotypes that are themselves under selection (reviewed in Berdan et al. 2023). The evolutionary forces maintaining both haplotypes are thought to be quite varied between populations and arrangements, but a common first hypothesis for a frequently sampled inversion is that local adaptation in one region or environment is balanced with migration from another (Kirkpatrick and Barton 2006; Charlesworth and Barton 2018). This association may be maintained more strongly when the alleles have beneficial epistatic interactions, and assumed beneficial epistasis between loci is often paired with another mechanism of balancing selection operating at a locus to explain inversion (Dobzhansky 1949, 1950, 1970; Schaeffer et al. 2003; Hoffmann and Rieseberg 2008). However, unless migration rates are very high, local adaptation models predict inversions to be at high frequency in their favored geographic range but at low frequency elsewhere, and many inversions do not have sampled locations at which they reach very high frequency.

Alternately, chance associations of both an inversion and its standard arrangement counterpart with different recessive deleterious alleles may mimic the fitness profile of a single overdominant locus (sometimes termed associative overdominance), where heterozygotes carrying both main haplotypes experience none of the deleterious effects of a homozygous recessive across the region (Sturtevant and Mather 1938; Frydenberg 1963; Zhao and Charlesworth 2016; Gilbert et al. 2020). However, recent theory and simulation suggests that this dynamic is uncommon unless an arrangement polymorphism is maintained at intermediate frequency under some other selective force, and the population scaled recombination rate between arrangements is low (Charlesworth 2024).

Balanced polymorphic loci may be maintained in other ways, under various models and selective pressures. For example, multiple overdominant loci that are each most beneficial as heterozygotes may benefit from linkage maintaining complementary haplotypes (Sved 1968; Charlesworth and Charlesworth 1973; Charlesworth 1974). Several recent studies have proposed that functionally overdominant loci may be more common than previously considered. Specifically, the dominance relation between alleles at a locus may change between different sexes, seasons, life history stages, or other trait or fitness challenge contexts, generally termed ‘reversal of dominance’, which in some cases can lead heterozygotes to have greatest average fitness (Rose 1982; Wittman et al. 2017; Grieshop and Arnqvist 2018; Connallon and Chenoweth 2019). Other specific biological dynamics may predict the maintenance of inversion polymorphism in particular ways. For example, inversions may also link drivers and enhancers in meiotic drive systems (Babcock and Anderson 1996; Mroczek et al. 2006; Courret et al. 2019), and these drive-associated inversions may persist at equilibrium frequencies in the presence of repressors (e.g. Bastide et al. 2022).

In natural populations, it can be difficult to distinguish among these alternate hypotheses. They are not always mutually exclusive, and the selective pressures acting upon an inversion may vary across space and time. The frequency patterns of many inversions, including a number in Drosophila melanogaster, remain poorly explained.

Inversions in Drosophila melanogaster

D. melanogaster is a widely studied organism in many fields of biology, including evolutionary and population genetics. Multiple polymorphic inversions segregate at meaningful frequencies in natural populations of D. melanogaster (Lemeunier and Aulard 1992; Aulard et al. 2002; Kapun et al. 2016; Kapun and Flatt 2019), but even in this model system, it is not entirely clear which population genetic processes are most responsible for shaping their frequencies. In this species, achiasmy in males (Morgan 1910; John et al. 2016) and a lack of evidence for female fecundity costs for most inversions (Sturtevant 1921; Sturtevant and Beadle 1936; Coyne et al. 1993; Gong et al. 2005) suggest that most paracentric inversions in D. melanogaster do not generate aneuploid gametes at any appreciable frequency. Further, genome editing tools have allowed for the creation of synthetic inversions, revealing that the tested structural rearrangements are not themselves responsible for the gene regulatory changes that are associated with natural inversions in D. melanogaster (Said et al. 2018). Instead, given that breakpoints of common D. melanogaster inversions occur disproportionately in regions that confer greater recombination suppression (Corbett-Detig 2016), D. melanogaster inversions would appear to be maintained primarily for the sake of maintaining linkage associations between distant variants.

D. melanogaster originated in southern-central Africa, in seasonally dry Miombo and Mopane woodlands, and (potentially after adapting to a human commensal niche) expanded out across much of Africa and into the Middle East around 13,000 years ago, and into Europe around 1,800 years ago (Sprengelmeyer et al. 2020). With an ancestral Ne of around 2 million (Sprengelmeyer et al. 2020), D. melanogaster populations have high levels of standing diversity and high population recombination rates. Many D. melanogaster inversions have geographic clines repeated between several colonized regions (Mettler et al. 1977; Reinhardt et al. 2014; Kapun et al. 2016), and inversions often show geographic differences in frequency that exceed genome-wide average differentiation between the same populations (e.g. Pool et al. 2017). It has therefore been suggested that local ecological adaptation may shape the distribution of D. melanogaster inversions (Kapun et al. 2016; Kapun and Flatt 2019). Indeed, it is difficult to explain their geographic differentiation without some type of spatially-varying selective pressures.

Yet, it is unclear whether ecological adaptation is a sufficient explanation for D. melanogaster inversion frequency patterns, for two reasons. First, most of the common inversions become less frequent as one samples populations farther from the sub-Saharan ancestral range (or similar warm environments) into potentially more challenging, colder high latitude and altitude environments (Aulard et al. 2002; Kapun et al. 2016; Lack et al. 2016a; Pool et al. 2017; Kapun and Flatt 2019) with In(3R)Mo representing an exception to this pattern (Kapun et al. 2014; Kapun et al. 2016). As these inversions are all derived (being absent from closely related species), local adaptation theory would predict that they arose to protect local adaptation against ongoing migration, more likely in the novel environment if it is experiencing high levels of immigration (Kirkpatrick and Barton 2006). It is unclear why inversions would be needed to adapt to the species’ ancestral environments, or why ancestral “standard” arrangements would have an ecological advantage in derived environments that are very different from the ancestral range.

Second, even in the locations where inversions would hypothetically be ecologically advantageous, they rarely surpass intermediate frequencies (Lemeunier and Aulard 1992; Pool et al. 2017; Kapun and Flatt 2019; Sprengelmeyer et al. 2020). Under a local adaptation model, that observation would require a secondary explanation that resists fixation, such as very high levels of long distance gene flow or the fixation of recessive deleterious variants on inverted haplotypes (Ohta and Kimura 1970; Zhao and Charlesworth 2016; Gilbert et al. 2020). While migration may well limit inversion frequencies in some populations, we note that some putatively adaptive Single Nucleotide Polymorphism (SNP) variants do achieve much greater frequency differentials between populations (da Silva Ribeiro et al. 2022), and thus we would have to suppose that inversions are under relatively weaker selection to explain their lesser differentiation. Regarding the hypothesis of high recessive load, we note that D. melanogaster populations do generally contain substantial recessive genetic load (Greenberg and Crow 1960), and it has been speculated that associative overdominance between sampled backgrounds might boost inversion frequencies in D. melanogaster populations during founder events (Pool et al. 2012) or in isolated wilderness environments (Sprengelmeyer et al. 2020). However, it seems likely that in large human-commensal populations, D. melanogaster inversions would escape from linked deleterious variants through gene conversion or double crossover events with standard chromosomes, in light of both the age of most of these inversions (Corbett-Detig and Hartl 2012) and the very high population recombination rates of D. melanogaster. Homozygous lines for each common inversion also appear at appreciable frequency in inbred lab strains, demonstrating a lack of perfect linkage of the inversions with recessive lethal or infertile variants (e.g. Lack et al. 2016a).

While local adaptation, gene flow, and associative overdominance may all contribute to geographic variation in inversion frequency among D. melanogaster populations, it is not clear that these processes are collectively sufficient to explain observed patterns. Alternatively, the allele frequencies and diversity patterns of at least some of these inversions might be primarily shaped by some form of balancing selection acting on inversion-linked variation, in which the balanced frequency is dependent on the environment (Kapun et al. 2023). Balancing selection on inversions has received increasing attention in recent literature (Wellenreuther and Bernatchez 2018; Faria et al. 2019). In Drosophila, inversions have been associated with balancing selection due to seasonal, temporally fluctuating selection (Machado et al. 2021). However, the potential for temporally varying selection to maintain stable balanced polymorphisms on its own is somewhat limited (Gillespie 1988). Further, most common inversions maintain relatively high polymorphic frequencies across broad geographic ranges without obvious shared seasonal selective pressures, from tropical and subtropical dry forests to temperate locations with harsh winter conditions (Lemeunier and Aulard 1992). Therefore, we predict that the frequencies of some of these inversions could be in part explained by a mechanism of balancing selection that would operate even within a temporally and spatially homogenous population. Below, we propose such a model in which alleles that contribute to competitive mating display simultaneously contribute antagonistically to survival to reproductive maturity, a tradeoff that offers a mechanism by which the life-history characteristics of a population could generate epistatic balancing selection.

Sexually Antagonistic Polymorphism in D. melanogaster

Among mechanisms of selection, sexual selection is both prevalent, and provides a simple mechanism for frequency dependence (O’Donald et al. 1997). Alleles that contribute to competitive mate choice systems may often follow a negative frequency dependence within the population, as display success depends on the relative quality of a male among local competitors, and the more common a display-enhancing allele is, the fewer competitors it succeeds over. Combinations of display-enhancing alleles may similarly gain a disproportionate benefit together, as male mating success in competitive display systems can be skewed towards a few successful individuals (Bateman 1948; Jones et al. 2002; Tatarenkov et al. 2008, though see Gowaty et al. 2012; Hoquet et al. 2020) such that the last few alleles to push display into the top quantile contribute a significantly greater fitness increase to the genotype than the first few display alleles. This interaction of alleles is a form of epistasis - a non-independent fitness interaction between loci - that is emergent and non-specific to the genetic identity of the variants. In other words, the alleles contribute independently to the generation of a trait, and the fitness of the individual is a nonlinear function of that trait (Blows and Brooks 2003; Rest et al. 2013; Otwinowski et al. 2018). This epistatic interaction suggests a potential for indirect selection on the recombination-suppressing effects of inversions, in order to maintain combinations of sexually antagonistic variants that have the greatest fitness when carried together. This tendency of selection to favor the reduction of recombination when the haplotypes present are at a fitness optimum is well studied for recombination modifiers including inversions, and has been termed the ‘reduction principle’ (Feldman 1972; Feldman and Balkau 1973; Feldman et al. 1980; Altenberg and Feldman 1987; Altenberg et al. 2017; see Altenberg and Feldman 1987 for a direct treatment of inversions).

While the emergent epistasis generated by a mate choice system may favor linkage of variants, alleles that contribute only to a sexually selected display are not expected to remain as balanced polymorphisms, rather they are expected to fix, albeit perhaps slowly as their fitness benefits diminish at high frequencies. However, if these alleles simultaneously contribute deleteriously to other components of fitness, they are pleiotropic (Stearns 2010), and given the sex-limited benefits of the display they would likely also be sexually antagonistic (Rowe et al. 2018), whereby traits increase the fitness of one sex but decrease fitness if found in the other. Pleiotropy can result from traits that are inherently physiologically coupled, from resource allocation restraints, or from the specific role or behavior of genes. Pleiotropy can further be sexually antagonistic when the traits in the tradeoff are differentially important to different sexes, which may be common given the differing fitness requirements of the sexes in a population (Cox and Calsbeek 2009; Connallon and Chenoweth 2019). It should be noted that a model of two pleiotropically determined traits each with potentially sex-specific fitness can be treated separately from a model where one trait has sexually antagonistic fitness effects.

D. melanogaster mating involves a male display performance and female acceptance or rejection. A number of experimental evolution studies (Rice 1992; Rice 1996; Audet et al. 2024) and genetic studies (Innocenti and Morrow 2010; Cheng and Kirkpatrick 2016; Glaser-Schmitt et al. 2023) have confirmed the prevalence of sexually antagonistic loci in D. melanogaster. Pleiotropic alleles that trade off between survival and male display quality represent one clear potential case of antagonism to consider for a model of balancing selection, and one highly relevant to the biology of D. melanogaster.

Given the relevance of sexual selection and sexual antagonism to Drosophila evolution, we propose that polymorphic D. melanogaster inversion and standard arrangements may help maintain haplotypes of alleles that specify opposite ends of a sexually antagonistic tradeoff. We first explore this model through forward simulation, to establish that it may in theory arise in a population and maintain inversion polymorphism over a significant number of generations. Then we examine the results of an experiment with an outbred laboratory population of D. melanogaster, to test whether common inversions show changes in frequency over the course of a generation in a manner consistent with a trade-off between survival and male reproductive success. Many such antagonistically pleiotropic scenarios are possible under different selective tradeoffs, and indeed some have been proposed for other model inversions with varying mechanisms of maintenance (Betrán et al. 1998; Pearse et al. 2019; Mérot et al. 2020; Pei et al. 2023; see Discussion), but we focus on a specific scenario of sexual antagonism here.

Results

A Model of Sexually Antagonistic Inversions

Here, we consider a population featuring a competitive mate choice system that results in a skewed distribution of reproductive success among males. Within this population, we model a genomic region containing a number of linked loci, each with alternate antagonistic, pleiotropic alleles that trade off between male display success and survival likelihood (of females or both sexes). Critically, we do not directly model any epistatic interactions among loci affecting survival or display traits, nor do we invoke dominance in our diploid model, so any such fitness dynamics emerge from the modeled population processes.

If mate choice is highly competitive, such that each female can choose amongst several or many males, the variance of offspring number among males (Figure 1A) may be much higher than for females - yielding a distribution of male reproductive output that follows a relatively convex trend, in which most males produce few offspring. Conversely, superior display quality can enable males to mate successfully much more often than other display phenotypes they compete against (Figure 1A). An increase in the display of a male with low quality results in a smaller change in number of offspring than an increase in display of a male with high quality (change ii in Figure 1A), due to the reproductive skew towards highly competitive individuals. So, there is an emergent non-additive fitness effect, where multiple variants increasing the reproductive rank of an individual together have a larger fitness benefit than the combined effects of each substitution separately - a positive, synergistic epistatic effect. This effect occurs whether the difference is substituting alleles at two loci, or substituting two homologous alleles at a locus in a diploid or polyploid, which would give an emergent dominance effect. Additionally, fitness conveyed by an allele favoring display quality is also frequency-dependent: since mating success depends on the display qualities of other males, the relative advantage of a display trait will be diminished as more males carry it, for both haploid and diploid populations (Figure 1B, male fitnesses). We refer to this as a negative frequency dependent selection upon alleles favoring display, though the alternate alleles have a positive frequency dependence. We note that some authors instead use “negative frequency dependence” to specifically describe fitness functions in which any and all alternate alleles are of selective benefit when rare (e.g. Charlesworth and Charlesworth 2010). And so, for an antagonistic variant that benefits display and begins under net positive selection, the benefit will decrease as the variant rises in frequency. If a variant only conveys benefit to mate competition success, it is expected to fix given enough time, as even if its evolution is nearly neutral at high population frequency, selection will push the frequency back up if it drifts down (Hazel 1943; Dickerson 1955; Charlesworth and Hughes 2000). However, if the variant is antagonistically pleiotropic for traits not involved in competition between males, such as survival in the face of climate, pathogens, or predators, these fitness costs may typically not be frequency dependent. If the benefit of an antagonistic display-favoring allele overcomes the pleiotropic cost when the variant is rare, its diminished benefit when at higher frequency may reverse its net benefit, while at some intermediate frequency, we predict that these fitness effects should be equal and an equilibrium frequency should be reached and maintained (Figure 1B). In the absence of other forces, this dynamic is expected to maintain the allele under balancing selection indefinitely.

Conceptual representations of the proposed model of inversion-associated balanced sexual antagonism. (A) Hypothetical skewed distribution of male mating success, yielding a greater variance in reproductive success for males than for females in our model. (B) Fitnesses of display-favoring alleles under sexually antagonistic balancing selection, illustrating a pleiotropic variant that should rise in frequency when rare, but decline if frequency exceeds the balanced equilibrium value. (C) Fitnesses of display-favoring alleles at two haploid loci under the same model, illustrating synergistic epistasis between displaying-favoring alleles. (D) A hypothetical trajectory of four such mutations, in which B outcompetes A, an inversion links B and C to create a more strongly display-favoring haplotype, and then the addition of D to that haplotype furthers the accumulation of antagonistic variants, reaching an equilibrium frequency while displacing less extreme display-favoring haplotypes.

We further predict that multiple such balanced pleiotropic loci should generate interesting emergent behavior. Because of the emergent synergistic epistasis from mate competition (see discussion of Figure 1A above), the haplotype with the greatest male display benefit should increase in frequency relative to other haplotypes until reaching an equilibrium at which its relative advantage in mate choice is reduced enough to be balanced by the pleiotropic, frequency-independent, non-epistatic survival cost (Figure 1B). The haplotypes and their linked alleles conveying intermediate phenotypes will have less than proportional benefits during mate choice due to the convex relationship between quality and success described above (Figure 1A), but a larger, proportional cost to survival against non-intraspecific survival challenges, giving them lower average fitness in the equilibrium state than either of the extreme haplotypes (Figure 1C). Therefore, recombination between the two extreme haplotypes will generate less fit haplotypes in the gametes, and a local modifier of recombination that reduces recombination and its associated costs (such as an inversion) will be favored by indirect selection (Altenberg and Feldman 1987; Otto and Lenormand 2002). And so, selection will favor the association of these haplotypes with the alternate standard and inverted karyotypes, which generate fewer recombinant gametes from a heterokaryotypic parent. We further predict that in a population that is close to the balanced equilibrium state for such a pleiotropic trade-off, the population of adults will contain a higher frequency of survival-focused haplotypes than the population of zygotes, but the display-focused haplotypes will increase in frequency in zygotes relative to adults, in line with the life stage-specific fitness differences between these haplotypes.

While Figure 1 illustrates a haploid scenario for the sake of simplicity, this model extends naturally to a diploid case. The main difference is that heterokaryotypic diploids – those heterozygous for the arrangement – would have intermediate trait values and be less fit than either homokaryotype, for the same selective reasons that an intermediate haplotype with two alleles of opposing effects at different loci is least fit in a haploid context. This is therefore, interestingly, an example of a scenario in which the two most common haplotypes in the population can exhibit underdominant fitness, and yet may still be maintained at a stable balanced equilibrium. By comparison, the simplest models of underdominant fitness predict an unstable equilibrium that disfavors polymorphism (Gillespie 2004, Charlesworth and Charlesworth 2010), a characteristic shared with non-frequency-dependent models of disruptive selection. The cost of generating underdominant offspring might be compensated for by a strong co-directional dominance of the variants involved, such that heterokaryotypic individuals exhibit a phenotype similar to one of the homokaryotypes. However, dominance is not a necessary feature of our model, and it is not invoked in our simulations. Whereas, investigating a wider range of genetic architectures would be a potential direction for future study.

Simulation Results

We investigated the predictions of the above model of inversion-associated balanced sexual antagonism using a purpose-built forward simulation program, “Sexually Antagonistic Inversion simulator” (SAIsim; see Materials and Methods). This simulation program incorporates the generation of new nucleotide and inversion polymorphisms by mutation, where each nucleotide mutation independently draws a positive or negative effect on survival and on male display. It also simulates crossing over and gene conversion, and inherently the effects of drift (Materials and Methods). SAIsim includes gene conversion between all homologous sites regardless of karyotype, and models crossing over in heterokaryotypic individuals from even-numbers of crossover events within the inverted region. It explicitly models mate selection, in that for each female randomly selected to reproduce, a set number of ‘encountered’ males are randomly selected, and the observed display value of each male for this competitive event is assigned based on the sum of the male’s genetic effects and an environmental noise effect resampled at every encounter. The male with the highest observed display value then succeeds in this mating (Figure 2).

The layout of a single simulated generation. Simulated individuals each reach the reproductive stage with probability in proportion to the product of the survival effects of their alleles. To generate each offspring in the next generation, a surviving female is first randomly sampled (with replacement) to be the mother. Then a specified number of surviving males are randomly sampled to generate the pool of males encountered by the sampled female - without replacement for each pool and with replacement between pools. For each encounter pool, each male is assigned an observed display quality as the sum of the display effects of its alleles, plus a normally distributed noise effect for each encounter. The highest quality male in the encounter pool is selected as the father, and the offspring genome is generated. The female gamete is generated with crossover and gene conversion events. Odd numbers of crossovers in the interior of an inversion are resampled from the same parent, to model the biology of an organism that eliminates aneuploid gametes during oogenesis, or exhibits reproductive compensation. D. melanogaster males do not cross-over, and so males do not recombine in the simulations presented.

With this simulator, we first confirmed that some individual variants with an antagonistic pleiotropy between survival (initially of both sexes) and male reproduction can generate stable selectively balanced polymorphisms in silico. We simulated a single locus in diploid populations of N = 1000, polymorphic for an allele of a specified survival cost and display benefit at an initial frequency of 0.5, with the alternate allele having no trait effect. One hundred males were sampled for each mating competition. After 20N generations, we calculated the proportion of simulations retaining polymorphism and the average final frequency of the variant across all simulations, revealing that there is a range of paired trait effects for which the mutation remained at intermediate frequency after 20N generations (Figure 3). Alleles with small effect sizes had only a narrow band of balanced selection, outside of which they were either fixed or lost, while large-effect alleles remained balanced across a larger area of simulated parameter space. While this balanced range is somewhat narrow, such variants may persist under balancing selection indefinitely in the absence of other forces, and many small-effect alleles in linkage blocks may confer larger effects in synergy. In a scenario where only females pay the survival cost, the balanced parameter space of antagonistic trait effects is broader and shifted somewhat (Figure S1). In a scenario where only males pay the survival cost, and variants are neutral in females, essentially no simulations retain polymorphism (Figure S1). The number of males sampled to assess mate competition for each offspring also has a significant effect, with larger values giving broader polymorphic ranges, though simulations with costs only in females show robust polymorphism even when only sampling two males for each competition (Figure S1).

Simulations show that a balanced equilibrium frequency exists for certain sexually antagonistic variants. The proportion of simulations retaining polymorphism (top) and the average final frequency (bottom) are plotted for a single locus at which one antagonistic allele with the indicated effects reducing survival proportion (of both sexes) and increasing male reproductive quality score is initiated at equal frequency with an alternative allele that has no such effects. Larger effect sizes are plotted in the left panels, with values incrementing by 0.025, and smaller effect sizes (a subset of the larger) are plotted in greater detail in the right panels, with values incrementing by 0.005. Most parameters result in either all simulations retaining polymorphism (purple), or else all simulations losing polymorphism (white). However, simulations retaining polymorphism have a broad range of equilibrium frequencies (shades of green). For each parameter combination, 500 replicates were simulated for a diploid population of 1,000 individuals evolving for 20,000 generations, with further details as indicated in the Materials and Methods.

We then tested the behavior of two linked loci, both polymorphic for an antagonistic and non-antagonistic allele. We identified two antagonistic variants previously found to undergo balancing selection separately, and performed simulations with varying genetic distances between them (Figure 4). Crossing over and gene conversion were simulated, with gene conversion occurring for each heterozygous variant at a probability of 10−2, with random direction. per female meiosis per site, intended to be a conservative approximation of the number of gene conversion events affecting a given site in D. melanogaster, scaled to simulations of N=1000 (Comeron et al. 2012). The stronger antagonistic variant (modelled at fixed position 0.225M) persisted in all simulations, in line with expectations, and so we focus on whether the weaker variant (at varying positions) also persists with it. As seen in the upper panel of Figure 4, the weaker variant (pale red line) was retained at close linkage, as the haplotype containing both antagonistic variants acts as a single larger effect supergene. However, as the weaker variant was moved farther from the stronger variant, recombination increasingly generated unfit intermediate genotypes compared to the extremes of the reproduction-survival tradeoff, and the stronger variant outcompeted the weaker variant among these haplotypes, leading to the loss of the weaker variant.

Simulated inversions may persist as polymorphisms when linked to sexually antagonistic, pleiotropic variants. In turn, the presence of inversions facilitates the accumulation of antagonistic variation. Rates of long-term persistence are shown for simulations with two defined antagonistic variants alone (left panels) and with the additional presence of a defined inversion (right panels). To focus the potential addition of a second linked variant to an antagonistic haplotype, we begin with the stronger variant already at an intermediate frequency in the population (0.5). The weaker antagonistic variant and the inversion either both start at 0.5 frequency as well (upper row) or both start at a lower frequency of 0.05 (lower row). The 1 Morgan chromosomal segment is diagrammed above the plots, with the inversion breakpoints marked (0.21M, 0.55M), chosen to include space across and beyond the inverted region. The proportion of simulations in which the second, variably-positioned weaker antagonistic variant (survival and reproductive values 0.86, 0.12) maintains polymorphism is shown (light red line), dependent on the recombination distance from a stronger allele (0.75,0.3) at fixed position 0.225M, which always persists (dark red line at 1 across the top of each plot). In the right panels, the proportion of replicates in which the inversion is retained is also plotted (blue line). Values for each parameter combination were averaged over 1000 replicate simulations, for populations of 1000 diploid individuals run for 20N generations. Gene conversion occurs for each heterozygous variant at a probability of 10−2 per female meiosis with random direction. Crossover arrangements generating aneuploidy, where there are an odd number of crossovers within the inverted region of a heterokaryotypic parent, are resampled from the same parent to represent removal to polar bodies or reproductive compensation.

We then examined the same spacings between antagonistic variants across a region containing a polymorphic inversion, with its left breakpoint close to the stronger variant, and the right breakpoint within the range of the positions in which the weaker variant was simulated (Figure 4, lower panel). When the inversion was added, the weaker variant was then retained when the two antagonistic variants are relatively more distant, due to association of the weaker variant with the right-hand inversion breakpoint (pale red line), at which point this association also preserves the inversion itself in a polymorphic state (light blue line). While more stochastic, a weaker antagonistic variant near the middle of the inversion also appeared to maintain polymorphism more frequently in systems starting with an inversion than without, i.e. across approximate positions 0.30-0.45M in Figure 4, in spite of its weaker linkage with the inversion itself due to double cross-over events. The inversion is not maintained when the weaker antagonistic variant is within the inversion but tightly linked to the stronger variant (see positions 0.22-0.25M); here there is little crossing-over for an inversion to suppress, and so the effect of the inversion approaches neutrality. Interestingly, there is a small window of weaker variant position in which the inversion maintains polymorphism despite the already tight linkage between SA variants, where the inversion seemingly does not modify recombination much (see positions 0.2-0.22M). Here, the weaker antagonistic variant is more tightly linked to the breakpoint than to the stronger variant in this range, and presumably benefits more consistently from the crossover reduction. The maintenance of the weaker variant drops more rapidly outside of the inversion than over similar distances in the simulation without inversions (positions 0-0.2M, bottom versus top panel). This seemingly counter-intuitive result may stem from crossover events that result in aneuploid gametes being resampled in our simulations, which effectively increases the crossover rate outside the inversion breakpoints, a behavior which should be biologically reasonable in taxa with aneuploid chromosome elimination or reproductive compensation. So, for a weaker variant outside the left inversion breakpoint, the presence of the inversion can increase the crossover rate between the two antagonistic variants, and thus reduce the probability of the weaker variant surviving by retaining linkage to the stronger one.

In further simulations, we began with a population initially free of variation, in which sexually antagonistic mutations arise stochastically. To assess differences between longer or shorter antagonistic regions, or higher or lower crossover rates, we modeled populations with parameters scaled to either 1 Mb or 10 kb, using estimates from Comeron et al. (2012) and Huang et al. (2016). The mutations had pleiotropic effects increasing reproductive quality score and decreasing survival proportion, each drawn sampled independently (Materials and Methods).

Depending on the effect scores drawn, a mutation could be broadly advantageous (potentially leading to a selective sweep), broadly deleterious, or potentially subject to balancing selection (Figure 3). These mutations arose at a rate equal to one thousandth of the scaled single nucleotide mutation rate given in Huang et al. (2016). Inversions (of random lengths and positions) also arose stochastically, at a hundredth of the rate of antagonistic variants. The frequency trajectories of antagonistic variants and inversions are tracked, and the simulation was run for 20N generations to allow near-equilibrium patterns to be established. These simulations demonstrated the potential of a population without initial balanced polymorphism or genetic variation to establish alternate karyotypes which link sexually antagonistic, pleiotropic alleles and an inversion, to maintain polymorphism in a balanced manner, and then to subsequently accumulate significant antagonistic character (Figure 5A), although the degree of antagonism accumulated depends strongly on the rate of crossing over (relative to the occurrence of antagonistic mutations; Figure 5B).

Simulated populations with stochastic mutation of new inversions and sexually antagonistic variants generate persistent inversion polymorphisms and accumulate linked sexually antagonistic variants. (A) In simulated populations with stochastic mutation of new variants having pleiotropic effects on survival and male display, levels of antagonism increased with time before stabilizing at an intermediate value (left panel). Whereas, simulations which also allowed the mutation of new inversions continued accumulating stronger antagonism throughout the simulated time interval (right panel). (B) In simulations which allowed the mutation of new inversions, a low crossover rate (left panel, 43.6 cM in females after scaling) allows populations to accumulate larger differences between the most common karyotype and all others than when the rate of crossover is particularly high (right panel, 43.6 M in females after scaling). Gene conversion occurs at each heterozygous variant with a scaled probability of 0.1295 per heterozygous variant in both scenarios. Values given are the mean of 1000 replicate simulations of 1000 diploid individuals run for 20N generations.

The simulations that successfully maintained balanced antagonistic chromosomes also had an interesting distribution of allele and genotype frequencies of the balanced haplotypes at equilibrium. Without any selection, the inversion frequency follows a pattern characteristic of the expected neutral site frequency spectrum (Figure S2). But strikingly, in parameter spaces where simulations with shared survival costs between sexes maintained polymorphism under balancing selection, inversion frequency distributions were tightly clustered around 0.25 and 0.75 among the zygote population (Figure S2, 6A). We found that this outcome was related to the genotype frequencies of reproducing adults. These simulations evolve toward strong antagonism, and it is predominantly males homozygous for the display-favoring haplotype (whether that is inverted or standard) who succeed in reproducing, and these males therefore contribute one copy of the display-favoring haplotype to each of their male and female offspring. However, this haplotype’s significant cost to viability and longevity would select for the heterokaryotypic females over the display-favoring homokaryotypes. Given that successful parents would predominantly be heterokaryotypic females and display-favoring homokaryotypic males, around three quarters of transmitted chromosomes have the display-favoring arrangement and one quarter does not, yielding the 0.25 and 0.75 zygote inversion frequencies seen in simulations. Although these are autosomal simulations, selection yields an inheritance pattern is similar to that of a ZW sex chromosome system, where the Z chromosome is homokaryotypic in the fathers, and heterokaryotypic with the W chromosome in the mothers.

The sex-specific fitness dynamics present at equilibrium under this model also result in population-wide genotype frequencies that depart from standard Hardy-Weinberg expectations. If the preponderance of matings are between heterokaryotypic females and homokaryotypic display-favoring males, then homokaryotypes for the survival-focused haplotype will rarely be produced (as confirmed in Figure 6B). We note that in cases where the standard arrangement favors male display, then especially if antagonism is strong, this model could explain an observed deficiency or absence of inversion homokaryotypes, even if the inverted arrangement was not associated with any recessive deleterious load. Further, heterokaryotypes may become disproportionately frequent after viability selection (Figure 6B). And still, this enrichment for heterokaryotypes reflects neither the existence of recessive deleterious load associated with each arrangement, nor overdominance with respect to viability. Hence, our model may offer another alternative explanation for deviations from Hardy-Weinberg genotype frequencies observed for polymorphic inversions in natural populations.

Antagonism-associated karyotypes reach predictable equilibrium frequencies in simulations with antagonistic and inversion mutations, when survival costs are shared between sexes. (A) Histograms of the number of arrangements at a specified frequency across all simulation replicates in females (left) and males (right), colored by how the average survival effect of that arrangement ranks compared to the others in its population. Ranks are normalized to be relative to the median arrangement value, to better account for rare arrangements and ties. These simulations reach an equilibrium with two predominant arrangements, with the more survival-focused arrangement around an overall frequency of 0.25 and more common in females, and the more male-competition-focused arrangement around a frequency of 0.75 and more common in males. (B) Because successful males tend to be homozygotes and successful females heterozygotes, the less frequent of the two major arrangements (which tends to favor survival) has homokaryotype frequencies far below the Hardy-Weinberg expectation, and there is an excess of heterokaryotypes. These genotype frequencies shift somewhat between zygote (left panel) and adult (right panel) populations as the genotype affects survival.

We further ran simulations in which the survival costs were not shared, but instead specific to females, representing a distinct, stronger form of sexual antagonism. In these simulations, we surprisingly recovered three different equilibrium arrangement states (Figure 7). 20.0% of simulations exhibited a state similar to the shared costs simulations presented in Figure 6, where we infer that fathers were homokaryotypic for a display-favoring arrangement, and mothers were heterokaryotypic and carried survival-focused arrangement that is observed in surviving but not reproducing males. 37.0% of simulations also developed two predominant arrangements, but with transmission patterns similar to an XY sex chromosome system rather than a ZW sex chromosome system. In these cases, we infer that the mothers were homokaryotypic for an X-like survival-focused arrangement, but the fathers were heterokaryotypic and carried a Y-like display-focused arrangement that was apparently lethal in females. 40.6% of simulations instead featured three intermediate frequency haplotypes: with both Y-like and W-like arrangements restricted to reproducing fathers and mothers respectively, and a third shared arrangement with intermediate phenotypic effects.

In simulations with female-limited survival costs, three distinct outcomes are observed involving either two or three balanced haplotypes. From simulations with randomly occurring antagonistic and inversion mutations, histograms of inversion frequencies across simulations for surviving females (left) and males (right) are partitioned into three categories based on sex-specific haplotype frequency outcomes. The “Favors Female” category (top) includes simulations in which there are exactly two arrangements with frequencies between 0.1 and 0.9 in both sexes (20.0% of replicates). The “Male-Specific” category (middle) includes simulations in which there are two arrangements with whole-population frequencies between 0.1 and 0.9, but one arrangement is absent in adult females (37.0% of replicates). The “Three Arrangement” category (bottom) includes simulations in which there are exactly three arrangements with whole-population frequencies between 0.1 and 0.9 frequency (40.6% of replicates). 2.3% of simulations did not fit in these categories and had only one predominant arrangement. As in Figure 6, inversions are colored by how the average survival effect of that arrangement ranks compared to the others in its population, and ranks are normalized to be relative to the median arrangement value, to better account for rare arrangements and ties.

Specific properties of our simulations may have influenced the occurrence of these three stable states under the female-limited cost model. To reduce the computational complexity of both crossovers and arrangement-specific trait values, the simulations do not allow two inversions on the same individual chromosome, whether overlapping or not, though two chromosomes sampled from the population may have distinct inversions with overlapping breakpoints. This means that if two arrangements, each with an inversion, become the prevalent haplotypes and thereby remove all standard arrangement chromosomes from the population, there is no mutational path to generate a third arrangement. However, a portion of the two-state simulations still had the ancestral, non-inverted arrangement as one of the two haplotypes present, which could have been mutated to a third arrangement to establish the three-arrangement pattern. We are not sure what selective or mutational barriers may prevent transitioning between each state once established. One clear barrier is that it is difficult to generate a less extreme arrangement from a Y-like arrangement, as these can accumulate an indefinite number of mutations since males experience no costs.

Empirical tests for inversion-associated tradeoffs

A specific motivation for developing and evaluating the model of inversion-associated, balanced sexual antagonism was to potentially better explain the pattern of inversion frequencies observed in D. melanogaster, in which multiple inversions reach some of their highest frequencies in populations from the ancestral range or with similarly tropical/subtropical climates. We therefore set out to test experimentally whether any of four inversions common in a Zambia population of this species show evidence for tradeoffs between male reproductive success and survival (in terms of viability and/or longevity).

We assembled an outbred population with moderate frequencies of each inversion to obtain karyotypically diverse males. We set up four crosses using these males along with females from one of four different inversion-free inbred Zambia lines. This crossing scheme is represented in Figure 8. These parental crosses occurred in a single large population cage, with populations of at least 600 of each sex. We developed a PCR and next-generation sequencing-based assay to estimate inversion-associated SNP frequencies in each pool of males and in each corresponding pool of embryos. These SNPs (given in Table S1) were based on our own analysis of variants that showed perfect association with inversions among 197 Zambia genomes (Table S2; see Materials and Methods). We then implemented a resampling method to test for significant inversion frequency changes between parents and offspring (while accounting for the variance observed in technical replicates; see Materials and Methods), in order to determine whether males carrying each inversion were more or less successful in reproducing than the average male. We also compared frequencies between embryos and aged adults that eclosed either in the first or second half of the collection period, to investigate the relationship between inversions and viability/longevity. We calculated p-values from the resampling tests to assess the significance of each individual change in inversion frequency between relevant samples (Table S3). However, in accord with our primary question regarding fitness tradeoffs, we primarily focus here on joint tests of two opposing frequency changes, as further described below. We could therefore test the hypothesis that a given inversion had opposing influences on survival and male reproduction, which would provide evidence consistent with our model of balanced sexual antagonism, while also examining whether any such tradeoffs depend on maternal genetic background.

The layout and potential expectations of the laboratory evolution experiment. We cross outbred males from a high-inversion population to inbred, non-inverted females from a specific inbred line, and collect DNA samples from the fathers, embryo offspring, and aged adult offspring (left). We hypothesize from our antagonistic pleiotropy model that one or more of the four inversions present on separate chromosome arms in our experimental populations may experience opposing selection between fathers and embryos versus between embryos and aged adults (middle). To test the significance of observed frequency changes, we designed a resampling model to represent the sources of neutral variation expected in the experiment (right).

When we initially examined the joint effects of male reproduction and survival, we found that in a majority of cases (in 13 out of 16 inversion-line trials; binomial P = 0.02; Figure 9), a given inversion increased in frequency in adult offspring compared to the null expectation based on the parental male pool. The remaining three cases were all for In(3L)Ok. Ten of the 13 increases were primarily due to increase in frequency between embryo and adult samples, suggesting that in terms of viability and/or longevity, inversion heterokaryotypes often had higher fitness than standard homokaryotypes, as no offspring homokaryotypic for an inversion were generated by our crossing design. We note that our standard homokaryotypes should not have broad-scale homozygosity within the inverted region, but are instead heterozygous for unique standard-arrangement haplotypes from two independent Zambia strains, and that long genomic tracts of identity-by-descent between Zambia strains are quite rare in this large, ancestral-range population (Lack et al. 2016a). Hence, we do not expect that inversion heterokaryotypy is buffering against deleterious effects of inbreeding. Instead, this fitness advantage suggests either an advantage for heterokaryotypy in particular, or perhaps more likely, that multiple inversions harbor alleles that benefit survival under our experimental conditions (see Discussion).

Some Drosophila inversions show evidence for antagonistic tradeoffs and sex-specific survival. (A) Testing for antagonistic fitness effects on reproduction and survival. Each inversion’s frequency in parents, embryo offspring, and adult offspring are shown for each maternal inbred line. Parental frequency is taken as the paternal frequency divided by two to account for the inversion-free inbred mothers. Adult offspring frequencies are taken from the sum of estimated allele counts across all adult offspring cohorts. * indicates p-value < 0.05 for evidence of significant reversed frequency change from parental to embryo and from embryo to offspring, combined across all four maternal line crosses and corrected for multiple tests, when compared to neutral simulated experiments (details in Materials and Methods). (B) Testing for evidence of sex-specific survival effects. The changes in inversion frequencies between embryos and both female and male adult offspring are illustrated. * indicates p-value < 0.05 for significant differences in female versus male inversion frequency among adult offspring, combined across all four maternal crosses and corrected for multiple tests.

When we separated the effects of male reproduction (based on father to embryo changes) and survival (embryos versus adults), we found that the effects of inversions on survival were generally consistent across maternal genotypes. Three inversions were associated with increased survival: In(3R)K was in all four crosses, while In(2L)t and In(2R)NS each showed elevated survival with all maternal strains except ZI418N (Figure 9A). In contrast, In(3L)Ok was associated with lower survival in each cross (Figure 9A). With regard to male reproduction, the trajectory of inversion frequencies from fathers to embryos to offspring was less consistent between different crosses (Figure 9A), implying that differences in male reproductive success may be dependent on the maternal genotype, which is consistent with past findings in D. melanogaster (Reinhart et al. 2015). Despite such variability, each of the four inversions had multiple frequency changes that departed strongly from null expectations (Figure 9A; Table S3), as might be expected if selection was present, but varied in strength and direction between crosses and life stages.

To test our central hypothesis of a tradeoff between male reproductive success and cumulative survival during development and adult aging, we evaluated the significance of each change in inversion frequency between the paternal and embryo cohorts, alongside the change in inversion frequency between the embryo and combined adult offspring cohorts. Among these results, In(3R)K had particularly consistent changes in frequency across replicate lines, with a decrease between parents and embryos, and then an increase between embryos and aged offspring (Figure 9A). When we assessed the likelihood of observing contrasting frequency shifts between life stages of this magnitude, integrating across the four maternal crosses and accounting for multiple test correction, we found that the In(3R)K pattern was significantly non-random (p = 2.70 × 10−3; Table S4; see Materials and Methods). The other inversion that gave reversed frequency shifts from cross-averaged data was In(3L)Ok, which unlike In(3R)K, was associated with (variably) higher male reproductive success and (consistently) lower survival. The frequency reversal observed for In(3L)Ok was also found to be statistically significant (p = 2.97 × 10−7). Wright-Fisher estimates of selection coefficients for these and all comparisons are given in Table S8.

We also tested for frequency differences between the different sexes of adult offspring and between the early-eclosing and late-eclosing cohorts of adult offspring. All four inversions displayed differences in frequency between female versus male offspring (Figure 9B), consistent with differential effects on sex-specific survival. Strikingly, In(3L)Ok and In(3R)K each displayed sex survival differences in alignment with their antagonistic effects implied above: In(3L)Ok reduced female survival more strongly than male survival (p = 0.0450), whereas In(3R)K favored the survival of females more than it did males (p = 4.61 × 10−3; Table S4; see Materials and Methods). The other two inversions, In(2L)t and In(2R)NS, each consistently favored female over male survival in all four maternal crosses, although their initial unidirectional cross-combined p values (0.0260 and 0.0126) were not significant after multiple test correction (p = 0.282 and 0.137). In contrast, we found no consistent effects of inversions on eclosion time (Figure S3). Estimated selection coefficients for each tested change based on a Wright-Fisher population approximation are available in Table S7; however these may overestimate inversion fitness effects since the survival terms reflect the probability of reaching advanced age. Further, there may be an influence of differential sex-specific mortality leading to greater frequency shifts in females. We collected a total of 1960 females and 3156 males, and if the survivors represent 20% of an original population with an equal sex ratio, then there was 84.7% female mortality and 75.3% male mortality. Overall, we conclude that D. melanogaster inversions have distinct and often context-dependent influences on male reproduction and on survival, with some evidence for antagonistic tradeoffs of the sort proposed in our model.

Discussion

Here, we have introduced a new conceptual model in which (1) sexually antagonistic variation is present and subject to balancing selection that maintains extreme genotypes favoring either male reproduction or survival, and (2) inversions are predicted to facilitate the aggregation of progressively more antagonistically differentiated haplotypes. We have tested this model using simulations, confirming the predicted efficacy of balancing selection and demonstrating that inversions indeed facilitate the construction of haplotypes containing multiple antagonistic variants. Finally, we have deployed Drosophila lab experiments to uncover some evidence for inversion-associated tradeoffs between reproduction and survival, depending on the specific inversion and strains tested.

Inversions in D. melanogaster

We have presented results from a single generation laboratory evolution experiment, in which we crossed outbred male D. melanogaster carrying inversions In(2L)t, In(2R)NS, In(3L)Ok, and In(3R)K to four separate inbred female lines and tracked the frequency of the inversions in the initial males, embryonic offspring, and aged adults. For most inversions, the strength and direction of the tradeoff varied notably between crosses. This might suggest high variance in their phenotypic effects or potentially maternally-dependent variation in male reproductive-success, but is not distinguishable from experimental noise with only one replicate per maternal line. However, greater consistency was observed for In(3R)K, which was associated with lesser male reproductive success but greater survival. A significant tradeoff was also inferred for In(3L)Ok, which favored male reproduction (albeit variably) over survival. These results provide significant but measured evidence for the presence of sexually antagonistic autosomal inversions in this model species, while emphasizing the importance of genetic background.

As indicated in the Introduction, models like ours in which inversion frequencies are influenced by balancing selection on pleiotropic tradeoffs may help explain multiple aspects of observed D. melanogaster inversion frequencies. Adjacent models have been explored in flies, for example an inversion in Drosophila buzzatii that was found to trade off between faster development time and larger adults (Betrán et al. 1998), and an inversion polymorphism found to trade off between larval survival and adult reproduction in the seaweed fly Coelopa frigida (Mérot et al. 2020). Some other groups found to harbor antagonistic inversions include the rainbow trout (Pearse et al. 2019) and the zebra finch (Pei et al. 2023). These related systems are variously hypothesized to involve maintenance by overdominance (Mérot et al. 2020), linked deleterious recessives (Pei et al. 2023), or reversals of dominance (Pearse et al. 2019), but these mechanisms are not mutually exclusive. This class of models inherently accounts for the intermediate maximal frequency of most D. melanogaster inversions in natural populations. To the extent that inversions favor male reproduction, the sexually antagonistic model we propose here could also help account for the predominantly autosomal locations of common inversions in this species, given that male-specific fitness effects have a weaker influence on the frequencies of X-linked variants since the X chromosome spends only one third of its time in males.

This model could also help explain the observed geographic variability of inversion frequencies, as a change in environment may change the fitness value of either pleiotropic effect of the alleles. If a population experiences a novel harsh environment with low survivability, then the population density may decrease, and the number of encountered competing males may decrease as well, reducing the relative fitness advantage of enhanced male display. If there is a genetic tradeoff between mating display and viability, the equilibrium frequency of an antagonistic inversion that favors the genetic extreme of display should be lower. Additionally, in hostile environments, the survival cost of display-favoring variants could be increased, which could further shift their equilibrium frequencies downward or render them not worth carrying at any frequency. These predictions could contribute to the observed negative relationships between some inversion frequencies and latitude or altitude (Kapun et al. 2016; Pool et al. 2017). As a specific example, it is notable that in a high altitude Ethiopian environment with year-round cool weather, which hosts some of this species’ most pronounced phenotypic evolution (Bastide et al. 2014; Lack et al. 2016b), inversions are virtually absent - occurring at the lowest frequencies of any analyzed population (Sprengelmeyer et al. 2020). Consistent with such reasoning, a disadvantage of some inversions in challenging thermal environments (both cold and hot) was supported by a previous experimental evolution analysis that included In(2L)t, In(2R)NS, and In(3R)K (Kapun et al. 2014), though they did find that In(3R)C consistently increasing in frequency in hot conditions and In(3R)Mo in cold. However, our experimental results include evidence that some of these inversions, such as In(3R)K, may have a tradeoff opposite from our prediction - that the inverted arrangement is associated with survival success instead of reproductive success. Conversely, a recent genome-wide association study found In(3R)K heterokaryotypes to be associated with elevated male reproductive success compared to standard homokaryotypes (Yamamoto et al. 2024), underscoring that inversion-related tradeoffs may depend on genetic background and/or experimental context.

More broadly, we observed a trend toward increased inversion frequencies across our full experimental generation. A general benefit for inversions would mirror previous findings in which inversions found at intermediate frequencies in natural D. melanogaster populations (Inoue 1979), as well as inversions from seaweed flies known to be under balancing selection in the wild (Mérot et al. 2020), were both inferred to be under directional selection in the lab. However, one laboratory study did report evidence that a D. melanogaster inversion not tested here, In(3R)P, had frequency-dependent fitness under crowded conditions (Nassar et al. 1973). Furthermore, even an inversion found to be under directional selection in the lab might still be subject to a balanced tradeoff a more challenging or complex natural environment. Other studies in D. melanogaster, including those on focused seasonal evolution, have found evidence for important tradeoffs between reproduction and robustness to environmental challenges (e.g. Behrman et al. 2015). Hence, it would clearly be desirable to study the effects of inversions on a wider range of potential tradeoffs. For example, Said et al. (2018) suggested that inversions In(2L)t and In(3R)K might be maintained by immunity-related tradeoffs, in light of an over-representation of immune-related genes among differentially expressed transcripts between inverted and standard karyotypes.

In addition to a broader array of balanced tradeoffs that could modulate D. melanogaster inversion frequencies, drift and other forms of selection may contribute as well. Although we emphasize above the potential interplay between ecological adaptation and balanced tradeoffs, some inversions could be impacted by simpler forms of directional selection. However, we must then invoke secondary explanations to account for their limited maximal frequencies (see Introduction). Other forms of balancing selection may also come into play as well. It is clear that some variation in this species is subject to temporally-varying selection, and some inversions have displayed seasonal frequency differences (Dobzhansky 1943; Kapun et al. 2016; Machado et al. 2021; Lange et al. 2022). As with spatial clines, temporal shifts could reflect either simple directional selection or else environmentally-modulated tradeoffs in a model such as ours. Spatial selective pressures may vary both between populations and within population scales.

As indicated in the Introduction, balancing selection on inversions could also be achieved by linkage to recessive deleterious variants, which may be observed as an antagonistic pleiotropic phenotype despite potentially involving strictly deleterious variants in linkage (Pei et al. 2023). In extreme form inversions might be under associative overdominance (Zhao and Charlesworth 2016; Gilbert et al. 2020), where two haplotypes are balanced by recessive cost to both homozygotes. Inversions may be particularly susceptible to carrying recessive deleterious alleles if population size and/or inversion frequency are low, since under these circumstances inversion homozygotes are rarely generated and hence the crossover rate among inverted chromosomes is low, plus there are fewer inverted chromosomes to facilitate rare double recombination events with standard chromosomes. However, this dynamic is not expected to establish new inversions easily (Charlesworth 2024) and is expected to decay with increased age and population size of the inversion, as mutation, gene conversion, or double crossover resolve the negative linkage between recessive deleterious alleles. Most common inversions in D. melanogaster are both relatively old and have large effective population sizes. Inversions including In(2L)t, In(2R)NS, and In(3R)K have breakpoints estimated to be several times older than the expansion of the species out of southern-central Africa (Corbett-Detig and Hartl 2012; Sprengelmeyer et al. 2020). Those three inversions, plus In(3L)Ok, for which inversion age has not been estimated, all maintain frequencies greater than 0.15 within very large populations (Kapun and Flatt 2019; Sprengelmeyer et al. 2020), and so it is not clear that the persistence of recessive deleterious or lethal alleles in the inverted regions is expected to be high enough that inversions would be maintained by a shared recessive lethal variant preventing fixation. However, sampling of natural populations has demonstrated a somewhat higher frequency of recessive lethal variants in inverted haplotypes, but these are largely unique and heterokaryotypic individuals remain robust, giving low costs to a heavily outbred population (Mukai and Yamaguchi 1974). Many wild-derived inbred strains of D. melanogaster are homokaryotypic for inversions (Lack et al. 2016a). Furthermore, as noted in the Results, an observed lack of inversion homozygotes could reflect strong sexual antagonism as opposed to recessive deleterious variation. Nevertheless, further simulation and experimental testing regarding the potential influence of associative overdominance and other processes on inversion frequencies in D. melanogaster and other species would be desirable.

Generality of Balanced Pleiotropic Inversions

We have proposed that some polymorphic inversions in D. melanogaster, and in other species with analogous mating dynamics, may be maintained in part through balancing selection acting on a frequency-dependent tradeoff. While we explored the tradeoff between male reproduction and viability, we emphasize that any antagonistic pleiotropy involving at least one frequency-dependent trait has the potential to generate similar inversion polymorphism. Such antagonistic pleiotropy may be common in evolution, and might be useful in explaining the many examples of inversions associated with distinct phenotypes.

For the specific model explored here, of antagonistic pleiotropy between survival and display in a mate choice system, the prevalence of the dynamic across species and populations is unclear. It seems plausible that a similar model of antagonistic pleiotropy and balancing selection under mate choice underlies other systems, such as the balanced inversion system in the European ruff Calidris pugnax (formerly Philomachus pugnax), where three inverted haplotypes on an autosome determine aggressive, cooperative, and female-mimic male lekking morphs, albeit with a dominance relationship and homozygous inviability for an inverted haplotype (Küpper et al. 2016). A recent study on the female-mimic Faeder morph suggests that it may be maintained in part by a balance between its effects lowering female reproductive output while increasing male mating success, and the system is likely more complex as a whole due to the three interacting morphs (Giraldo-Deck et al. 2022). More generally, a number of factors contribute to the potential prevalence of this dynamic. Mate competition and choice during reproduction is rather common among animals, and this may mean sexually antagonistic inversions are plausibly also common. The model depends on the distribution of fitness effects of new mutations and the availability of sexually antagonistic variation, but such variation does seem prevalent (Zajitschek and Connallon 2018; Ruzicka et al. 2019). More likely to limit the prevalence of such inversions are the meiotic costs experienced by inversion heterokaryotypes in some systems (White 1973), whether sexually antagonistic variants are clustered in a region that would allow inversions to increase linkage, and whether other mechanisms like sex-biased expression (Ingleby et al. 2015), gene duplication (Connallon and Clark 2011), and linkage to the sex chromosome or sex-determining locus would resolve the antagonism more frequently (though see following subsection on sex chromosome evolution).

Perhaps more importantly, there are a number of well-studied cases of inversion polymorphism that may follow this dynamic of balancing selection between haplotypes simultaneously demonstrating antagonistic pleiotropy and trait-specific frequency dependence, but which do not involve sexual antagonism. Papilio butterfly populations harbor well-studied mimicry polymorphisms that segregate between inverted segments and may often experience frequency-dependent selection (Clarke and Sheppard 1960; Kunte et al. 2014; Poul et al. 2014). The snail Cepaea nemoralis harbors diverse and distinct shell patterning haplotypes associated with inversions that remain balanced within single sampling locations (Cook 2013; Surmacki et al. 2013). Several inversions in different ant species segregate a single-queen colony strategy from a cooperative multi-queen strategy within the same range (Wang et al. 2013; Libbrecht and Kronauer 2014).

Essentially, any population with antagonistic pleiotropy involving some form of negative frequency-dependent balancing selection on one of the phenotypes might exhibit the same dynamic. The population would experience similar divergent selection towards alternate phenotypic extremes, with the frequency-dependence maintaining the presence of two genotypes. Importantly, this model involves a different mechanism for the maintenance of antagonistically pleiotropic variation than models based on reversals in dominance, where dominance effects generate a net heterozygote advantage at a locus (Connallon and Chenoweth 2019). The model proposed here integrates well with existing thought on the evolution of divergent selection on balanced systems. For example, Kopp and Hermisson (2006) have proposed a specific form of frequency dependent divergent selection (FDDS), involving competitive exclusion between similar genotypes for a trait otherwise under stabilizing selection, and this model was expected to result in single or few loci of large effect. Their model did not consider pleiotropy or linkage modifiers, but still had comparable dynamics in that selection in the FDDS regime favored small numbers of large effect loci, while the inversion haplotypes in the model presented here simplify the recombination architecture to approximate a single locus. This simplification of the genetic architecture follows the reduction principle in recombination modifier theory (Feldman 1972; Feldman and Balkau 1973; Feldman et al. 1980; Altenberg and Feldman 1987; Altenberg et al. 2017). When a population experiences selection towards two (or more) divergent optima, under a selection regime that provides a balancing mechanism to maintain both, indirect selection will favor a reduction of the genetic architectures to prevent the generation of intermediate phenotypes. The likelihood of either model likely rests on the inherent polygenicity of the trait under selection, with simpler traits being more likely to resolve by single locus FDDS dynamics than by inversion association (e.g. Yassin et al. 2016).

Notably, our simulations tended to result in strong sexual antagonism, such that it leads to a substantial fraction of the population failing to survive to reproductive age. It is unclear how common such extreme antagonism may be in natural populations. If it is uncommon in nature, even in taxa that experience high levels of mate competition as modeled here, then it is unclear what prevents our simulation predictions from being realized. It may be that genetic constraints prevent the occurrence of highly antagonistic genotypes, or that the relationship between survival-altering mutations and fitness is different than we have modeled. Or, populations may steadily evolve to mitigate antagonistic effects that arise and persist through processes such as those we modeled. Alternatively, some highly antagonistic haplotypes may become associated with sex determination, as suggested below.

Antagonistic autosomal inversions may precede novel sex chromosomes

If balanced, sexually antagonistic, autosomal inversions occur frequently enough, they may also contribute to the evolution of sex chromosomes. Given an established inversion with significant sexual antagonism, the inverted haplotype that favors one sex still experiences an antagonistic cost of transmission to offspring of the opposite sex. Mutations that link the haplotype with the sex it benefits are therefore favored under selection, and this can occur either by fusing to an existing sex chromosome (Charlesworth and Charlesworth 1980), though this only fully resolves the antagonism for linkage to Y and W chromosomes, or by linkage to a new sex determining mutation (van Doorn and Kirkpatrick 2007).

Here, we discovered that strong inversion-linked antagonism could yield a scenario where essentially all highly-successful males were homokaryotypic for a display-favoring karyotype, whereas reproducing females were overwhelmingly heterokaryotypic (Figure 5). This situation is similar to the ZW sex chromosome system, suggesting that this particular pleiotropic trade-off may favor resolution by linkage to a female sex determination factor. Further, simulations in which only females experienced antagonistic survival cost usually generated populations with a chromosome essentially lethal in females, thus exclusively transmitted in males similar to a Y chromosome (Figure 7). These simulations sometimes had three states that appeared stable at 20N generations, including Y chromosome-like, W chromosome-like, and shared intermediate arrangements. This three-arrangement stable polymorphism seems unusual for a natural population, but might be interesting to explore in populations with polymorphic sex determination.

There are often deleterious effects of chromosome fusion or novel sex determination (Pennell et al. 2015; Saunders et al. 2019), which may represent a general obstacle to sex chromosome evolution. However, balanced sexually antagonistic inversions may concentrate enough antagonism that the built-in benefit of resolving it may sometimes overcome the fitness costs of evolving new sex chromosomes (Blackmon and Brandvain 2017). Hence, antagonistic inversions might be more likely than other loci to be involved in the evolution of new sex chromosomes.

A basic prediction of the above hypothesis is that at least one inversion difference between sex chromosomes should already exist at the time of new sex chromosome formation, and genomic differentiation therefore begins between inversion karyotypes on an autosome even before sex-specific transmission of the region. This scenario contrasts with current models of sex chromosome evolution in which inversion accumulation and genomic differentiation accrue only after sex chromosome formation, due to advantages in suppressing recombination involving sexual antagonistic variants already associated with a sex-determining locus (Wright et al. 2016). In this context, we note that the neo-X and neo-Y homologs of Muller element C in Drosophila miranda, which became sex-linked about 1.5 million years ago, already appear to have inversion differences between them (Chen et al. 2024 – Figure 5A). Further, a recent study in Littorina saxatalis snails has discovered an ecotype-specific sex determination system that may be founded on such tradeoffs, though selection across ecotypes involves discrete geographic boundaries and is only partly comparable (Hearn et al. 2022). Still, additional genomic studies of new sex chromosomes are needed to assess the potential role of sexually antagonistic inversions in sex chromosome evolution.

Summary and Future Prospects

We have advanced a model in which inversion polymorphism is maintained by the frequency-dependent fitness effects of a pleiotropic tradeoff associated with inversion-linked genetic variation. We have used sexual antagonism associated with increased male reproduction versus survival as a specific motivating case, and we explore the predictions of this model using a novel simulation algorithm. We confirmed model predictions that some antagonistic variants can be maintained at an equilibrium frequency, and that inversions can allow multiple antagonistic variants to persist and form more strongly antagonistic haplotypes. We complemented our conceptual and computational modeling with laboratory experiments designed to detect fitness effects of four inversions common in an ancestral range population of D. melanogaster. We indeed found some evidence of tradeoffs between male reproduction and survival for these inversions, with two out of four inversions showing significant evidence for such a tradeoff, although we also observed a strong dependence of male reproductive success on female genetic background.

In light of the persistent mystery of how D. melanogaster inversion frequencies are determined, and how readily we detected tradeoffs involving male reproduction and survival, we suggest that further studies investigating the potential fitness tradeoffs of inversions in this model system are strongly warranted. While some such studies have previously been conducted, the applicability of our amplicon sequencing approach to estimate inversion frequencies in large pools of flies may improve the sensitivity of such experiments. Our understanding of the phenotypic impacts of D. melanogaster inversions would benefit from investigations of a wider range of fitness-related traits, including female fecundity, immunity, and performance in challenging environments such as cool temperatures and high population densities.

Clearly, it will also be important to test the generality of our model by estimating the fitness associations of inversions in a wider range of species. And particularly in the case of species with very recent changes sex chromosomes, it will be of great interest to test whether one or more inversions are already present between these new sex chromosomes, as predicted a model in which sex chromosome evolution involves the resolution of inversion-associated sexual antagonism. Such studies will advance our broader understanding of the roles of balancing selection, pleiotropic tradeoffs, and genome structure in shaping genetic variation and genomic evolution.

Materials and Methods

Simulations

We developed a new individual-level forward simulator (‘SAIsim’ for Sexually Antagonistic Inversion simulator) written in python 3.7 for the simulation of inversions and sexually antagonistic alleles at infinite loci, with uniform rates of mutation, crossing over, and gene conversion (per bp, per generation) over a number of independent chromosome arms. The dynamic we were interested in modeling requires direct accounting of pleiotropy between different components of fitness, crossing over and gene conversion within inversions (as well as the possibility of inversion fixation), and directly modeling male display and female choice, and no existing simulator provided this functionality. These simulations were used to establish the potential for the model to occur in a population generally, and not to compare against specific empirical observations. We have used literature estimates of mutation rate and conversion rate of Zambian D. melanogaster for these simulations (Comeron et al. 2012; Huang et al. 2016) but have no information on the joint distribution of effects of mutations on our model of survival and display in D. melanogaster. So, we sample from relatively broad but arbitrary combinations of mutation effects, only assuming that there is likely some degree of antagonistic pleiotropy naturally present in genomic variation. In SAIsim, a population is instantiated as a python object, and populated with individuals (also python objects) based on provided (or initially empty) genomes. In each generation, as diagrammed in Figure 5, the simulator calculates a survival probability for each individual as the product of the survival values of each variant they carry, which range [0,1]. Here, multiplicative selection for survival prevents biologically unreasonable negative survival probabilities from occurring. Male and female survival costs can each be included in a sex-specific manner, but only globally in the simulation, not per mutation. For each offspring in the next generation, a mother is selected randomly, weighted by individual survival probabilities, with replacement. To model the encountered males that might become the father of an offspring individual from this female, a set number of males (i.e. the uniformly specified encounter number) is randomly sampled from the full population of males, weighted by their survival probabilities. The males each have a genetic display quality given as the sum of the reproductive values of their alleles. A noise parameter, normally distributed around zero with a standard deviation of one, is added to each male display value during each encounter. The male with the highest noise-adjusted display quality among the encountered males is chosen to be the father. As offspring are generated, crossover locations are sampled, and resampled in cases where the resultant gamete would be aneuploid. Resampling aneuploids removes the fecundity cost that may exist in some taxa, but was chosen to reflect the lack of observed fecundity effects for most paracentric inversions of D. melanogaster. Gene conversion is modeled as a constant probability per heterozygous variant, with each event independently affecting just one locus, rather than modeling a conversion tract length (given that gene conversion tracts are generally short and each would typically contain only a single modeled variant in our simulations). This probability reflects the likelihood of any gene conversion event affecting a single nucleotide site, rather than the per-bp rate of initiation of gene conversion tracts. The dynamics of gene conversion are expected to have separate effects on the assembly and the maintenance of arrangements. The assembly of an arrangement is positively dependent on the population-scaled rate of recombination of alleles as new polymorphisms start at low frequency and often not on their optimal haplotype. However, the indirect fitness effect of the inversion is dependent on its effect on recombinant offspring, and therefore negatively depends on the per-individual rate of conversion. The simulation is intended to represent D. melanogaster life history and allows removal of recombination in males, which also removes gene conversion in males by default. Finally, inversions and new mutations are potentially added to the resultant gamete, with inversions resampled when overlapping the coordinates of an inversion already present on the same individual chromosome. SAIsim also allows the specification of mutation rates and the distribution of effect sizes of new mutations.

The simulations presented in Figures 5, 6, and 7 rescale parameter values by a factor of 2,000 to model an estimated ancestral Ne of around 2 × 106 (Sprengelmeyer et al. 2020) while using a smaller simulated population of 1,000 individuals. The low crossover rate simulations model a 0.0218 cM region and the high crossover rate simulations model 2.18 cM, which scale to 43.6 cM and 43.6 M female meiotic maps in the simulated populations. Based on an average autosomal recombination rate in females of 2.18 × 10−8 for D. melanogaster (Comeron et al. 2012), these represent 10kb and 1Mb regions. Gene conversion occurs at each heterozygous variant with a probability of 0.1295, based on scaling γ=1.25 × 10−7 events/bp/female meiosis with a conversion track length of 518 bp (Comeron et al. 2012) to a per-single-nucleotide-variant estimate of 6.475 × 10−5 conversions/bp/female meiosis. Sexually antagonistic mutations were modeled at a rate corresponding to one per thousand mutations, and so based on an autosomal estimate of 5.21 × 10−9 mutations/bp/gen on autosomes (Huang et al. 2016), and modeling 10kb and 1Mb regions, these were scaled to 1.042 × 10−4 and 1.042 × 10−2 events per gamete respectively. Inversions were approximated as arising once per hundred sexually antagonistic mutations, for 1.042 × 10−6 and 1.042 × 10−4 events per gamete. Both of these rates are empirically underdetermined. The simulations use mutations with uniformly distributed quality values in the range [0,1]; although distributions skewed toward small effects may be more realistic, our uniform approach avoids the need to invoke a specific but empirically unknown distribution. SAIsim’s object-oriented design allows specification of a population model by a new user familiar with the basics of python scripting. A population object can be parameterized and populated in a few lines of python script, and then parameters can be changed between periods of simulation, allowing bottlenecks and demographic changes as well as changes in mutational parameters between generations. Replicate SAIsim simulations were performed via the UW-Madison Center For High Throughput Computing (CHTC), which manages cycle servers using the HTCondor distributed computing project (Thain et al. 2005; Erlandson and Theisen 2018).

Experimental Populations and Sequencing

To act as a high-inversion source population for the potential fathers in our experiment, we combined 15 males and 15 females from each of 10 inbred strains derived from wild flies collected in Zambia. These 10 strains were chosen to collectively carry the following common inversions: In(2L)t, In(2R)NS, In(3L)Ok, In(3R)K. We then collected at least 500 F2 males from this high-inversion population to use in each of our independent maternal line experiments, and these males were expected to carry a wide range of inversion genotypes. Each batch of high-inversion males was crossed to an equal number of virgin females from a single inbred Zambian strain (Figure 6). Four such experiments were performed, each using one inversion-free maternal strain. Since no inversions could be inherited through the mothers, inversion frequencies of successfully reproducing F2 males could be inferred from their pooled offspring. Unless otherwise noted, all flies were kept in a lab space of 23°C with around a degree of temperature fluctuation and without a controlled day/night cycle. All fly food used was prepared in batches of 4.5 L water, 500 mL cornmeal, 500 mL molasses, 200 mL yeast, 54 g agar, 20 mL propionic acid, and 45 mL tegosept 10% (in 95% ethanol). The four parental populations were each allowed to mate freely in clear plexiglass cages of dimension 19.3 by 19.3 by 30.0 cm, with a loose mesh covering on one end for access by hand, twisted and clipped shut with a binder clip. The parents were given either fly food bottles or grape agar in 7.9 cm petri dishes as food or laying substrate on alternating days over two weeks, after which all surviving male parents were counted and frozen. The grape agar was prepared in batches of 500 mL from FlyStuff.com grape agar powder, and wet Red Star yeast was brushed over the agar plates before use to encourage laying. Embryos were collected from the agar, counted, and frozen for sequencing. Embryos laid in the bottles were allowed to hatch and develop into adults. These adult offspring were separated by sex and by the time after laying at which they eclosed. In our husbandry conditions, the earliest eclosion was 11 days after laying, but this was rare and a first eclosion at 12 days was more common among the bottles. Most flies took several days longer to develop, and we considered flies collected in the first 6 days of a 10 day collection window to be the early-eclosing pool, in order to split the pool sample sizes closer to evenly. The collected adult flies were then aged by sequentially transferring them between bottles weekly for approximately 2.5 months, and then the adult offspring were sexed, counted, and collected. A 10-week aging window was chosen based on the time taken for a control set of four bottles of flies from the paternal outbred population to reach 80% mortality. All adult and embryo samples were frozen at −80°C before DNA extraction. This sampling scheme was chosen to identify inversion frequencies in the set of potential parents, in embryonic offspring, and in older adults, and so to assess potentially selected changes in inversion frequency across male reproduction or in egg to adult (encompassing viability plus longevity). We chose to age the populations to include longevity as long term survival is relevant both in the estimated reproductive age of the species (Pool 2015), and the need to survive during periods of unsuitable environments, often while under reproductive diapause (Saunders et al. 1990; Ragland et al. 2019). In seasonally varying climates, either temperate or with long dry season, survival through challenging conditions is expected to require several months. In many such cases females are in reproductive diapause, and so longevity is the main selective pressure.

DNA for each replicate age group was extracted by a phenol-chloroform extraction protocol used previously in preparing DNA for the Drosophila Genome Nexus and Drosophila Population Genomics Project. The protocol is similar to protocol ‘BPC’ used in DPGP phase 1 (Langley et al. 2012), except with the addition of proteinase-K during homogenization of adult flies, and generally based on the common quick preparation protocol presented in Drosophila Protocols (Huang et al. 2009). Inversion genotyping was performed by amplicon sequencing, where amplicons associated with each inversion were designed to target at least one single nucleotide polymorphism in the inversion interior that is fixed for alternate alleles between the inversion karyotypes, while still sharing conserved primer regions across karyotypes to prevent primer bias. Genetic variation in these inversions was assessed using previously sequenced and inversion-called haploid genomes from the Zambian D. melanogaster population, available from the Drosophila Genome Nexus (Lack et al. 2016a). The amplicon region and primer selection was performed with a custom script. The script calls Primer3 version 2.4.0 for primer candidate generation (Koressaar and Remm 2007; Untergasser et al. 2012; Koressaar et al. 2018). Primers thus selected were ordered from IDT as oligos ligated to Illumina stubby adaptors [Table S1]. Illumina sequencing libraries were then prepared for each sample following the Illumina metagenomic amplicon sequencing protocol (Illumina 2013), with 50 ng/uL diluted DNA samples. and sequenced 2×150 on an Illumina NovaSeq to at least a depth of 20000 reads. Depth was selected to ensure high likelihood of detecting a 0.025 change in inversion frequency from an initial frequency of 0.10, based on simulated resampling of the collected flies, DNA extraction, and sequencing.

Sequence Analysis and Statistics

Sequence preparation and analysis were performed using custom pipeline scripts. Read trimming, alignment, and QC were performed using bwa version 0.7.17-r1188 (Li 2013), GATK version 3.4-46 (Van der Auwera and O’Connor 2020), and Picard version 1.79 (Broad Institute 2019). Reads were trimmed at the ends to remove segments with base quality less than 20, and read pairs were filtered to retain only those pairs where both reads aligned with mapping quality >20. SNP identity at the sites of interest was called from those remaining reads with base quality >20 at the SNP of interest. Due to the presence of chimeric reads in the amplicon library sequencing, we chose to use a single discriminant SNP to determine each inversion’s frequency in the read pool, instead of incorporating haplotype or additional SNP information.

To assess the significance of the difference in inversion frequency between just two sampled life stages in the same experiment, we first calculated a chi-square test statistic from the two-by-two table of inverted and non-inverted allele counts at two life stages. Assessment of the significance of frequency change reversals across three cohorts was done with concordant methods discussed subsequently. Allele counts were taken by multiplying the inversion-specific SNP frequency in the sequencing reads by the allele sample size. We then generated an approximate distribution of the chi-square statistic when selection is absent (our null hypothesis), using a resampling approach (Figure 6). Comparing the observed statistic to our modeled neutral distribution of test-statistic values allowed us to identify significance thresholds. Note that we are not conducting chi-square tests, but simply using the associated statistic as a convenient metric for comparing empirical and simulated data.

To generate the neutral distribution of the chi-square statistic, we took the observed paternal inversion frequencies and then sampled from them to obtain new paternal and offspring read counts to represent a single neutral data set as follows. First, we resampled the collected allelic counts (given in Tables S6 and S7) to account for sampling variance between replicates. For the paternal samples, we needed to account for pre-sequencing mortality as up to 15% of individuals died before collection. So, in generating a new paternal read set, we first modeled down-sampling of the paternal allele sample from the initial census allelic count to the collected allelic count as a hypergeometric sampling. For all other cohorts we used a binomial resampling from the collected allelic count. Second, we used a binomial sampling step to account for the experimental variance introduced by the extraction and library preparation. The extraction and library preparation appeared to introduce variation independent of allele sample size, possibly due to variation introduced during PCR or sample homogenization. This inference was based on analysis of data from control library preparations (Table S5) and from sequencing done in parallel with our experiment from sample fly pools with known inversion frequencies. These control samples included an average of 6 technical replicates (separate library prep and sequencing replicates from the same flies) from four different expected frequencies for In(2L)t, as well as one expected frequency for each of In(2R)NS and In(3R)K. Due to genotyping and husbandry difficulties we did not have a control set for In(3L)Ok. Based on the inversion frequency differences observed among these technical replicates, we estimated that taking a binomial sampling of size 316 from the sample allele frequency reflected a maximum likelihood model for the sample-size-independent variation introduced by the library prep and sequencing process. Third, from the preliminary resampled paternal inversion frequency, we then took a binomial sample of size equal to the read count to account for the random sample of sequencing reads, thus yielding resampled paternal read counts (given in Tables S6 and S7) that accounted for the number of sampled and non-sampled individuals, experimental variance, and depth of coverage. Fourth, to generate resampled offspring read sets, we instead used a binomial sampling of half the offspring alleles from the fully resampled paternal allele frequency for each separate offspring cohort, since the homozygous mothers all contributed non-inverted alleles. Finally, each offspring allele sample was resampled using the same size 316 binomial sampling to account for the variation introduced during this library’s preparation, and binomial sampled again with the read count size to account for sampling reads from the library, thus yielding resampled offspring read counts that accounted for the same three factors as for paternal counts.

For a pairwise comparison of inversion frequency between two samples, we calculated the same allele count chi-square statistic as described above for the empirical data for each resampled set of allele counts, generating the full distribution of neutral chi-square test statistics from which we could identify extreme value cutoffs. We then obtained a p-value from the proportion of resampled replicates in which the chi-square statistic was greater than or equal to the empirical value. If the test was directional, we took the proportion of resampled replicates with both a chi-square statistic greater than or equal to the empirical value and a frequency difference in the expected direction.

In order to identify a frequency change reversal between life history stages that could reflect a fitness tradeoff, we tested for the joint significance of two opposing frequency changes using three cohorts. A tradeoff would require a frequency change in one direction between paternal and embryo samples, followed by a frequency change in the opposite direction between embryo and adult offspring samples. These changes both depend on the shared embryonic cohort in the test, and so are not independent. Therefore, we evaluated the two inversion frequency changes jointly from a set of resampled neutral data sets in which the three cohorts arose from the same resampling process. Here we considered how likely it was that a sample of the three cohorts from the associated estimated neutral model had chi-square values for both the comparison of the fathers to embryos and the embryos to combined adults that were equivalent or more extreme in the test direction than the observed data. For example, when testing the significance of an observed paternal-embryo increase and then embryo-adult decrease in frequency, our p-value would be defined by the proportion of resampled replicates in which the frequency changes were in the tested directions of increase then decrease, and their chi-square values were both more extreme than the empirical values.

To assess the significance of inversion frequency changes considered across all four maternal line crosses at once, we used Fisher’s combined probability test. We applied this to both particular three-cohort frequency change reversals and to the set of two-cohort frequency changes. The inversion trajectory is independent between maternal crosses, but the differences between each cross in maternal line and initial paternal genetic variation make the replicates difficult to compare in a more structured way, so we felt this was an appropriate test for combined significance. In assessing our tradeoff hypothesis, we applied this test only for inversions where it was well-motivated; i.e. where the averaged frequency change across maternal crosses was in opposite directions between paternal/embryo and embryo/adult comparisons.

To account for multiple hypothesis testing in our cohort comparisons, we used a Benjamini-Yekutieli false discovery rate correction for significance. We applied this correction across the set of Fisher’s method combined p-values we generated to assess the relevant frequency change reversal hypotheses. So, this corrected for the multiple hypotheses of different tradeoff directions and different inversions involved, from p-values that already combined evidence from different maternal line crosses. Benjamini-Yekutieli was chosen due to its robustness to the unclear dependence relationships between some of the tests involved. For example, multiple tests of frequency changes from a single cohort (e.g. an embryo sample compared against both paternal and aged adult samples) are likely to be positively correlated.

Supporting information

eLife Assessment

This study attempts to solve long-standing puzzles about inversion polymorphisms in Drosophila melanogaster by invoking sexually antagonism and negative frequency dependent selection. While the idea developed here is a valuable contribution to the field, the description of the empirical work and the simulations remain incomplete, as they do not provide a full picture of what exactly has been done.

Reviewer #1 (Public Review):

The hypothesis is based on the idea that inversions capture genetic variants that have antagonistic effects on male sexual success (via some display traits) and survival of females (or both sexes) until reproduction. Furthermore, a sufficiently skewed distribution of male sexual success will tend to generate synergistic epistasis for male fitness even if the individual loci contribute to sexually selected traits in an additive way. This should favor inversions that keep these male-beneficial alleles at different loci together at a cis-LD. A series of simulations are presented and show that the scenario works at least under some conditions. While a polymorphism at a single locus with large antagonistic effects can be maintained for a certain range of parameters, a second such variant with somewhat smaller effects tends to be lost unless closely linked. It becomes much more likely for genomically distant variants that add to the antagonism to spread if they get trapped in an inversion; the model predicts this should drive accumulation of sexually antagonistic variants on the inversion versus standard haplotype, leading to the evolution of haplotypes with very strong cumulative antagonistic pleiotropic effects. This idea has some analogies with one of predominant hypotheses for the evolution of sex chromosomes, and the authors discuss these similarities. The model is quite specific, but the basic idea is intuitive and thus should be robust to the details of the model assumption. It makes perfect sense in the context of the geographic pattern of inversion frequencies.

To provide empirical support for this idea, the authors study the dynamics of inversions in population cages over one generation, tracking their frequencies through amplicon sequencing at three time points: (young adults), embryos and very old adult offspring of either sex (>2 months from adult emergence). Out of four inversions included in the experiment, two show patterns consistent with antagonistic effects on male sexual success (competitive paternity) and the survival of offspring, especially females, until an old age, which the authors interpret as consistent with their theory.

There are several reasons why the support from these data for the proposed theory is not waterproof.

(1) As I have already pointed out in my previous review, survival until 2 months (in fact, it is 10 weeks and so 2.3 months) of age is of little direct relevance to fitness, whether under natural conditions or under typical lab conditions.

The authors argue this objection away with two arguments

First, citing Pool (2015) they claim that the average generation time (i.e. the average age at which flies reproduce) in nature is 24 days. That paper made an estimate of 14.7 generations per year under the North Carolina climate. As also stated in Pool (2015), the conditions in that locality for Drosophila reproduction and development are not suitable during three months of the year. This yields an average generation length of about 19.5 days during the 9 months during which the flies can reproduce. On the highly nutritional food used in the lab and at the optimal temperature of 25 C, Drosophila need about 11-12 days to develop from egg to adult. Even assuming these perfect conditions, the average age (counted from adult eclosion) would be about 8 days. In practice, larval development in nature is likely longer for nutritional and temperature reasons, and thus the genomic data analyzed by Pool imply that the average adult age of reproducing flies in nature would be about 5 days, and not 24 days, and even less 10 weeks. This corresponds neatly to the 2-6 days median life expectancy of Drosophila adults in the field based on capture-recapture (e.g., Rosewell and Shorrocks 1987).

Second, the authors also claim that survival over a period of 2 month is highly relevant because flies have to survive long periods where reproduction is not possible. However, to survive the winter flies enter a reproductive diapause, which involves profound physiological changes that indeed allow them to survive for months, remaining mostly inactive, stress resistant and hidden from predators. Flies in the authors' experiment were not diapausing, given that they were given plentiful food and kept warm. It is still possible that survival to the ripe old age of 10 weeks under these conditions still correlates well with surviving diapause under harsh conditions, but if so, the authors should cite relevant data. Even then, I do not think this allows the authors to conclude that longevity is "the main selective pressure" on Drosophila (l. 936).

(2) It appears that the "parental" (in fact, paternal) inversion frequency was estimated by sequencing sires that survived until the end of the two-week mating period. No information is provided on male mortality during the mating period, but substantial mortality is likely given constant courtship and mating opportunities. If so, the difference between the parental and embryo inversion frequency could reflect the differential survival of males until the point of sampling rather than / in addition to sexual selection.

(3) Finally, irrespective of the above caveats, the experimental data only address one of the elements of the theoretical hypothesis, namely antagonistic effects of inversions on reproduction and survival, notably that of females. It does not test for two other key elements of the proposed theory: the assumption of frequency-dependence of selection on male sexual success, and the prediction of synergistic epistasis for male fitness among genetic variants in the inversion. To be fair, particularly testing the latter prediction would be exceedingly difficult. Nonetheless, these limitations of the experiment mean that the paper is much stronger theoretical than empirical contribution.

Reviewer #2 (Public Review):

Summary:

In their manuscript the authors address the question of whether the inversion polymorphism in D. melanogaster can be explained by sexually antagonistic selection. They designed a new simulation tool to perform computer simulations, which confirmed their hypothesis. They also show a tradeoff between male reproduction and survival. Furthermore, some inversions display sex-specific survival.

Strengths:

It is an interesting idea on how chromosomal inversions may be maintained

Weaknesses:

General points:

The manuscript lacks clarity of writing. It is impossible to fully grasp what the authors did in this study and how they reached their conclusions. Therefore, I cannot guarantee that I spotted all the problems in the study and in my review I will only highlight some cases that I found problematic.

Although this is an interesting idea, but it clearly cannot explain the apparent influence of seasonal and clinal variation on inversion frequencies.

Comments on the latest version:

I would like to give an example of the confusing terminology of the authors:

"Additionally, fitness conveyed by an allele favoring display quality is also frequency-dependent: since mating success depends on the display qualities of other males, the relative advantage of a display trait will be diminished as more males carry it..."

I do not understand the difference to an advantageous allele, as it increases in frequency the frequency increase of this allele decreases, but this has nothing to do with frequency dependent selection. In my opinion, the authors re-define frequency dependent selection, as for frequency dependent selection needs to change with frequency, but from their verbal description this is not clear.

One example of how challenging the style of the manuscript is comes from their description of the DNA extraction procedure. In principle a straightforward method, but even here the authors provide a convoluted uninformative description of the procedure.

It is not apparent to the reviewer why the authors have not invested more effort to make their manuscript digestible.

Reviewer #3 (Public Review):

Summary:

In this study, McAllester and Pool develop a new simulation model to explain the maintenance of balanced inversion polymorphism, based on (sexually) antagonistic alleles and a trade-off between male reproduction and survival (in females or both sexes). In support of the plausibility of this model, the authors use laboratory experiments on four naturally occurring inversion polymorphisms in Drosophila melanogaster, finding evidence for the existence of the above-mentioned trade-off in two out of the four cases.

Strengths:

(1) The study develops and analyzes a new (Drosophila melanogaster-inspired) model for the maintenance of balanced inversion polymorphism, combining elements of (sexually) antagonistically (pleiotropic) alleles, negative frequency-dependent selection and synergistic epistasis. To this end, the authors develop and use a new simulator.

(2) The above-mentioned model assumes, as a specific example, a trade-off between male reproductive display and survival; in the second part of their study, the authors perform laboratory experiments on four common D. melanogaster inversions to study whether these polymorphisms may be subject to such a trade-off. The authors find that two of the four inversions show suggestive evidence that is consistent with a trade-off between male reproduction and survival. The new amplicon sequencing approach to track inversion frequencies used by the authors seems promising in terms of studying fitness effects / trade-offs associated with polymorphic inversions and how such effects play out dynamically.

Weaknesses:

A gap in the current modeling is that, while a diploid situation is being studied, the model does not investigate the effects of varying degrees of dominance. It would be important and interesting to fill this gap in future work.

Comments on the latest version:

Most of the comments which I have made in my public review have been adequately addressed.

Some of the writing still seems somewhat verbose and perhaps not yet maximally succinct; some additional line-by-line polishing might still be helpful at this stage in terms of further improving clarity and flow (for the authors to consider and decide).

Author response:

The following is the authors’ response to the current reviews.

Public Reviews:

Reviewer #1 (Public review):

The hypothesis is based on the idea that inversions capture genetic variants that have antagonistic effects on male sexual success (via some display traits) and survival of females (or both sexes) until reproduction. Furthermore, a sufficiently skewed distribution of male sexual success will tend to generate synergistic epistasis for male fitness even if the individual loci contribute to sexually selected traits in an additive way. This should favor inversions that keep these male-beneficial alleles at different loci together at a cis-LD. A series of simulations are presented and show that the scenario works at least under some conditions. While a polymorphism at a single locus with large antagonistic effects can be maintained for a certain range of parameters, a second such variant with somewhat smaller effects tends to be lost unless closely linked. It becomes much more likely for genomically distant variants that add to the antagonism to spread if they get trapped in an inversion; the model predicts this should drive accumulation of sexually antagonistic variants on the inversion versus standard haplotype, leading to the evolution of haplotypes with very strong cumulative antagonistic pleiotropic effects. This idea has some analogies with one of predominant hypotheses for the evolution of sex chromosomes, and the authors discuss these similarities. The model is quite specific, but the basic idea is intuitive and thus should be robust to the details of model assumption. It makes perfect sense in the context of the geographic pattern of inversion frequencies. One prediction of the models (notably that leads to the evolution of nearly homozygously lethal haplotypes) does not seem to reflect the reality of chromosomal inversions in Drosophila, as the authors carefully discuss, but it is the case of some other "supergenes", notably in ants. So the theoretical part is a strong novel contribution.

We appreciate the detailed and accurate summary of our main theoretic results.

To provide empirical support for this idea, the authors study the dynamics of inversions in population cages over one generation, tracking their frequencies through amplicon sequencing at three time points: (young adults), embryos and very old adult offspring of either sex (>2 months from adult emergence). Out of four inversions included in the experiment, two show patterns consistent with antagonistic effects on male sexual success (competitive paternity) and the survival of offspring, especially females, until an old age, which the authors interpret as consistent with their theory.

As I have argued in my comments on previous versions, the experiment only addresses one of the elements of the theoretical hypothesis, namely antagonistic effects of inversions on male reproductive success and other fitness components, in particular of females. Furthermore, the design of this experiment is not ideal from the viewpoint of the biological hypothesis it is aiming to test. This is in part because, rather than testing for the effects of inversion on male reproductive success versus the key fitness components of survival to maturity and female reproductive output, it looks at the effects on male reproductive success versus survival to a rather old age of 2 months. The relevance of survival until old age to fitness under natural conditions is unclear, as the authors now acknowledge. Furthermore, up to 15% of males that may have contributed to the next generation did not survive until genotyping, and thus the difference between these males' inversion frequency and that in their offspring may be confounded by this potential survival-based sampling bias. The experiment does not test for two other key elements of the proposed theory: the assumption of frequency-dependence of selection on male sexual success, and the prediction of synergistic epistasis for male fitness among genetic variants in the inversion. To be fair, particularly testing for synergistic epistasis would be exceedingly difficult, and the authors have now included a discussion of the above caveats and limitations, making their conclusions more tentative. This is good but of course does not make these limitations of the experiment go away. These limitations mean that the paper is stronger as a theoretical than as an empirical contribution.

We discuss the choice to focus on exploring the potential antagonistic effects of the inversion karyotype on male reproductive success and survival in our general response above. Primarily, this prediction seemed to be the most specific to the proposed model as compared to other alternate models. Still, further studies are clearly needed to elucidate the potential frequency dependence and genetic architecture of the inversions.

Regarding the choice of age at collection, it is unknown to what degree our selected collection age of 10 weeks correlates with survival in the wild, but we feel confident that there will be some positive correlation.

We now further clarify that across our experiments, a minimum of 5% and a mean of 9% of the males used in the parental generation died before collection. These proportions do not appear sufficient to explain the differences between paternal and embryo inversion frequencies shown in Figure 9.

Reviewer #2 (Public review):

Summary:

In their manuscript the authors address the question whether the inversion polymorphism in D. melanogaster can be explained by sexually antagonistic selection. They designed a new simulation tool to perform computer simulations, which confirmed their hypothesis. They also show a tradeoff between male reproduction and survival. Furthermore, some inversions display sex-specific survival.

Strengths:

It is an interesting idea on how chromosomal inversions may be maintained

Weaknesses:

The authors motivate their study by the observation that inversions are maintained in D. melanogaster and because inversions are more frequent closer to the equator, the authors conclude that it is unlikely that the inversion contributes to adaptation in more stressful environments. Rather the inversion seems to be more common in habitats that are closer to the native environment of ancestral Drosophila populations.

While I do agree with the authors that this observation is interesting, I do not think that it rules out a role in local adaptation. After all, the inversion is common in Africa, so it is perfectly conceivable that the non-inverted chromosome may have acquired a mutation contributing to the novel environment.

Based on their hypothesis, the authors propose an alternative strategy, which could maintain the inversion in a population. They perform some computer simulations, which are in line with the predicted behavior. Finally, the authors perform experiments and interpret the results as empirical evidence for their hypothesis. While the reviewer is not fully convinced about the empirical support, the key problem is that the proposed model does not explain the patterns of clinal variation observed for inversions in D. melanogaster. According to the proposed model, the inversions should have a similar frequency along latitudinal clines. So in essence, the authors develop a complicated theory because they felt that the current models do not explain the patterns of clinal variation, but this model also fails to explain the pattern of clinal variation.

To the contrary – in the Discussion paragraph beginning on Line 671, we explain why we would predict that a tradeoff between survival and reproduction should lead to clinal inversion frequencies. We suggest that a karyotype associated with a survival penalty should be increasingly disadvantageous in more challenging environments (such as high altitudes and latitudes for this species). Furthermore, an advantage in male reproductive competition conferred by that same haplotype may be reduced by the lower population densities that we would expect in more challenging environments (meaning that each female should encounter fewer males). Individually or jointly, these two factors predict that the equilibrium frequency of a balanced inversion frequency polymorphism should depend on a local population’s environmental harshness and population density, with the ensuing prediction that inversion frequency should correlate with certain environmental variables.

Reviewer #3 (Public review):

Summary:

In this study, McAllester and Pool develop a new model to explain the maintenance of balanced inversion polymorphism, based on (sexually) antagonistic alleles and a trade-off between male reproduction and survival (in females or both sexes). Simulations of this model support the plausibility of this mechanism. In addition, the authors use experiments on four naturally occurring inversion polymorphisms in D. melanogaster and find tentative evidence for one aspect of their theoretical model, namely the existence of the above-mentioned trade-off in two out of the four inversions.

Strengths:

(1) The study develops and analyzes a new (Drosophila melanogaster-inspired) model for the maintenance of balanced inversion polymorphism, combining elements of (sexually) antagonistically (pleiotropic) alleles, negative frequency-dependent selection and synergistic epistasis. Simulations of the model suggest that the hypothesized mechanism might be plausible.

(2) The above-mentioned model assumes, as a specific example, a trade-off between male reproductive display and survival; in the second part of their study, the authors perform laboratory experiments on four common D. melanogaster inversions to study whether these polymorphisms may be subject to such a trade-off. The authors observe that two of the four inversions show suggestive evidence that is consistent with a trade-off between male reproduction and survival.

Open issues:

(1) A gap in the current modeling is that, while a diploid situation is being studied, the model does not investigate the effects of varying degrees of dominance. It would thus be important and interesting, as the authors mention, to fill this gap in future work.

(2) It will also be important to further explore and corroborate the potential importance and generality of trade-offs between different fitness components in maintaining inversion polymorphisms in future work.

We appreciate the work put in to evaluating, improving, and summarizing our study. We agree that further work studying the effects of dominance and of the fitness components of the inversions is important.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

l. 354 : I don't understand what the authors mean by "an antagonistic and non-antagonistic allele". If there is a antagonistic polymorphism at a locus, then both alleles have antagonistic effects; i.e., allele B increases trait 1 and reduced trait 2 relative to allele A and vice versa.

Edited, agreed that the terminology used here was sub-optimal.

Reviewer #2 (Recommendations for the authors):

The motivation for their model is their claim that the clinal inversion frequencies are not compatible with local adaptation. The reviewer doubts this strong statement. Furthermore, the proposed model also fails to explain the inversion frequencies in natural populations.

Hence, rather than building a straw man, it would be better if the authors first show their experiments and then present their model as an explanation for the empirical results. Nevertheless, it is also clear that the empirical data are not very strong and cannot be fully explained by the proposed model.

This claim that we reject any role of local adaptation in clinal variation and selection upon inversion polymorphism does not hold up in a reading of our manuscript. We even suggest that locally varying selective pressures must be playing some role, although that does not imply that local adaptation is the ultimate driver of inversion frequencies. Indeed, we suggest that local adaptation alone is an insufficient explanation for inversion frequency clines in D. melanogaster, including because (1) these frequency clines do not approach the alternate fixed genotypes predicted by local directional selection, (2) these derived inversions tend to be more frequent in more ancestral environments (l.113-158).

In our public review response above, and in the Discussion section of our paper, we explain why our model can predict both the clinal frequencies of many Drosophila inversions and their intermediate maximal frequencies. Of course, we do not predict that most inversions in this species should follow the specific tradeoff investigated here. In fact, we were surprised to find even two inversions that experimentally supported our predicted tradeoff. Still, it remains possible that other inversions in this species are subject to other balanced tradeoffs not investigated here, which could help explain why they rarely reach high local frequencies.

Reviewer #3 (Recommendations for the authors):

My previous comments have been adequately addressed.

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

[…]

To provide empirical support for this idea, the authors study the dynamics of inversions in population cages over one generation, tracking their frequencies through amplicon sequencing at three time points: (young adults), embryos and very old adult offspring of either sex (>2 months from adult emergence). Out of four inversions included in the experiment, two show patterns consistent with antagonistic effects on male sexual success (competitive paternity) and the survival of offspring, especially females, until an old age, which the authors interpret as consistent with their theory.

There are several reasons why the support from these data for the proposed theory is not waterproof.

(1) As I have already pointed out in my previous review, survival until 2 months (in fact, it is 10 weeks and so 2.3 months) of age is of little direct relevance to fitness, whether under natural conditions or under typical lab conditions.

The authors argue this objection away with two arguments

First, citing Pool (2015) they claim that the average generation time (i.e. the average age at which flies reproduce) in nature is 24 days. That paper made an estimate of 14.7 generations per year under the North Carolina climate. As also stated in Pool (2015), the conditions in that locality for Drosophila reproduction and development are not suitable during three months of the year. This yields an average generation length of about 19.5 days during the 9 months during which the flies can reproduce. On the highly nutritional food used in the lab and at the optimal temperature of 25 C, Drosophila need about 11-12 days to develop from egg to adult. Even assuming these perfect conditions, the average age (counted from adult eclosion) would be about 8 days. In practice, larval development in nature is likely longer for nutritional and temperature reasons, and thus the genomic data analyzed by Pool imply that the average adult age of reproducing flies in nature would be about 5 days, and not 24 days, and even less 10 weeks. This corresponds neatly to the 2-6 days median life expectancy of Drosophila adults in the field based on capture-recapture (e.g., Rosewell and Shorrocks 1987).

Second, the authors also claim that survival over a period of 2 month is highly relevant because flies have to survive long periods where reproduction is not possible. However, to survive the winter flies enter a reproductive diapause, which involves profound physiological changes that indeed allow them to survive for months, remaining mostly inactive, stress resistant and hidden from predators. Flies in the authors' experiment were not diapausing, given that they were given plentiful food and kept warm. It is still possible that survival to the ripe old age of 10 weeks under these conditions still correlates well with surviving diapause under harsh conditions, but if so, the authors should cite relevant data. Even then, I do not think this allows the authors to conclude that longevity is "the main selective pressure" on Drosophila (l. 936).

This is overall a thoughtfully presented critique and we have endeavored to improve our discussion of Pool (2015) and to clarify some of the language used about survival elsewhere. While we agree that challenges other than survival to 10 weeks are very relevant to Drosophila melanogaster, collection at 10 weeks does encompass some of these other challenges. Egg to adult viability still contributes to the frequencies of the inversions at collection and is not separable from longevity in this data. Collection at longevity was chosen in part to encompass all lifetime fitness challenges that might influence the inversion frequency at collection, albeit still within permissive laboratory conditions. Future experiments exploring specific stressors independently and beyond permissive lab conditions would generate a clearer picture.

In addition to general edits, the specific phrase mentioned at 1. 936 [now line 1003] has been revised from “In many such cases females are in reproductive diapause, and so longevity is the main selective pressure.” to “While longevity is a key selective pressure underlying overwintering, the relationship between longevity in permissive lab conditions without diapause and in natural conditions under diapause is unclear (Schmidt et al. 2005; Flatt 2020), and our experiment represents just one of many possible ways to examine tradeoffs involving survival.”

(2) It appears that the "parental" (in fact, paternal) inversion frequency was estimated by sequencing sires that survived until the end of the two-week mating period. No information is provided on male mortality during the mating period, but substantial mortality is likely given constant courtship and mating opportunities. If so, the difference between the parental and embryo inversion frequency could reflect the differential survival of males until the point of sampling rather than / in addition to sexual selection.

We have further clarified that when referenced as parental frequency, the frequency presented is ½ the paternal frequency as the mothers were homokaryotypic for the standard arrangement. We chose to present both due to considerations in representing the frequency change from paternal to embryo frequencies, where a hypothetical change from 0.20 frequency in fathers to 0.15 frequency in embryos represents a selective benefit (a frequency increase in the population), despite the reality that this is a decrease in allele frequency between paternal and embryo cohorts.

We mentioned a maximum 15% paternal mortality at line 827 [now l.1056], but have now added complete data on the counts of flies in the experiment as a supplemental table (Table S1) and have added or corrected further references to this in the results and methods [lines 555, 638, 975]. It is true that this may influence the observed frequency changes to some degree, and while we adjusted our sampling method to account for the effects of this mortality on statistical power [l.1056ff], we have now edited the manuscript to better highlight potential effects of this phenomenon on the recorded frequency changes.

It is also worth noting that, if mortality among fathers over the mating period is codirectional with mortality among aged offspring, this would bias the results against detecting an opposing antagonistic selective effect of the inversions on paternity share. This is now also mentioned in the manuscript, l.639ff.

(3) Finally, irrespective of the above caveats, the experimental data only address one of the elements of the theoretical hypothesis, namely antagonistic effects of inversions on reproduction and survival, notably that of females. It does not test for two other key elements of the proposed theory: the assumption of frequency-dependence of selection on male sexual success, and the prediction of synergistic epistasis for male fitness among genetic variants in the inversion. To be fair, particularly testing the latter prediction would be exceedingly difficult. Nonetheless, these limitations of the experiment mean that the paper is much stronger theoretical than empirical contribution.

This is a fair criticism of the limitations of our results, and we now summarize such caveats more directly in the discussion summary, lines 876ff.

Reviewer #2 (Public Review):

[…]

Comments on the latest version:

I would like to give an example of the confusing terminology of the authors:

"Additionally, fitness conveyed by an allele favoring display quality is also frequency-dependent: since mating success depends on the display qualities of other males, the relative advantage of a display trait will be diminished as more males carry it..."

I do not understand the difference to an advantageous allele, as it increases in frequency the frequency increase of this allele decreases, but this has nothing to do with frequency dependent selection. In my opinion, the authors re-define frequency dependent selection, as for frequency dependent selection needs to change with frequency, but from their verbal description this is not clear.

We have edited this text for greater clarity, now line 232ff. We did not seek to redefine frequency dependence, and did mean by “the relative advantage of a display trait will be diminished” that an equivalent s would diminish with frequency. We have now remedied terminological issues introduced in the prior revision with regard to frequency dependent selection.

One example of how challenging the style of the manuscript is comes from their description of the DNA extraction procedure. In principle a straightforward method, but even here the authors provide a convoluted uninformative description of the procedure.

We have edited for clarity the text on lines 1016-1020. Citing a published protocol and mentioning our modifications seems an appropriate trade-off between representing what was done accurately, citing the sources we relied on in doing it, and limiting the volume of information in the main text for such a straightforward and common method.

It is not apparent to the reviewer why the authors have not invested more effort to make their manuscript digestible.

We have invested a great deal of effort in making this manuscript as clear as we are able to. We regret that our writing has not been to this reviewer’s liking. We believe we have been highly responsive to all specific criticisms, including revising all passages cited as unclear. In this round, we have again scrutinized the entire manuscript for any opportunity to clarify it, and we have made further changes throughout. Although our subject matter is conceptually nuanced, we nevertheless remain optimistic that a careful, fresh reading of our revised manuscript would yield a more favorable impression.

Reviewer #3 (Public Review):

[…]

Weaknesses:

A gap in the current modeling is that, while a diploid situation is being studied, the model does not investigate the effects of varying degrees of dominance. It would be important and interesting to fill this gap in future work.

Agreed, and now reinforced at lines 892ff.

Comments on the latest version:

Most of the comments which I have made in my public review have been adequately addressed.

Some of the writing still seems somewhat verbose and perhaps not yet maximally succinct; some additional line-by-line polishing might still be helpful at this stage in terms of further improving clarity and flow (for the authors to consider and decide).

We have made further changes and some polishing in this draft, and greatly appreciate the guidance provided in improving the draft so far.

Reviewer #1 (Recommendations For The Authors):

(1) While the model results are convincing, some of the verbal interpretation is confusing. In particular, the authors state that in their model the allele favoring male display quality shows a negative frequency dependence whereas the alternative allele has a positive frequency dependence. This does not make sense to me in the context of population genetics theory. For a one-locus, two-allele model the change of allele frequency under selection depends on the fitness of the genotypes concerned relative to each other. Thus, at least under no dominance assumed in this model, if the relative fitness of AA decreases with the frequency of allele A, the relative fitness of aa must decrease with the frequency of allele a. I.e., if selection is negatively frequency dependent, then it is so for both alleles.

This phrasing was wrong, and we have edited the relevant section.

(2) I am still not entirely sure that the synergistic epistasis assumed in the verbal model is actually generated in the simulations; this would be easy enough to check by extracting the mating success of males with different genotypes from the simulation output should be reported, e.g., as a figure supplement.

Our new Figure S2, which depicts haplotype frequencies for a set of the simulations presented in Figure 4, should demonstrate a necessary presence of synergistic epistasis. These results further clarify that the weaker allele B is only kept when linked to A. The same fitness classes of genotype are present in the simulations with and without the inversion, so the only mechanical difference is the rate of recombination, and the only way this might change selection on the alleles is if a variant has a different fitness in one haplotype background than another – i.e. epistasis. The maintenance of haplotypes AB and ab to the exclusion of Ab and aB relies on the lesser relative fitness of Ab and aB. And since survival values are multiplicative, this additional contribution must come from the mate success of AB being disproportionately larger than Ab or aB, indicating the emergent synergistic epistasis posited by our model. We have clarified this point in the text at line 363ff.

(3) l. 318ff: What was this set number of males? I could not find this information anywhere. Also, this model of the mating system is commonly referred to as "best of N", so the authors may want to include this label in the description.

We indicate this detail just after the referenced line, now reworded and on l. 338-340 as “For each female’s mating competition, 100 males were sampled, though see Figure S1 for plots with varying encounter number.” Among these edits, “one hundred” has been changed to a numeral for easier skimming, and Figure S1 is now referenced here earlier in the text. Several edits have also been made in the caption of Figures 2 and 3, and in the relevant methods section to clarify the number of encountered males simulated, mention best of N terminology, and clarify how the quality score is used in the mate competition.

(4) The description of the experiment is still confusing. The number of individuals of each sex entered in each mating cage is missing from the Methods (l. 914); although I did finally find it in the Results. These flies were laying over 2 weeks - does this mean that offspring from the entire period were used to obtain the embryo and aged offspring frequencies, or only from a particular egg collection? If the former, does this mean that the offspring obtained from different egg batches were aged separately? Were the offspring aged in cages or bottles, at what density? Given that only those males that survived until the end of the two-week mating period were sequenced, it is important to know what % of the initial number of males these survivors were. A substantial mortality of the parental males could bias the estimate of parental frequencies. How many parental males, embryos and aged offspring were sequenced? Were all individuals of a given cage and stage extracted and sequenced as a single pool or were there multiple pools? The description could also be structured better. For example, the food and grape agar recipes and cage construction are inserted at random points of the description of the crossing design, which does not help.

We have now reorganized and edited these portions of the Methods text. Portions of this comment overlap with edits responding to (2) of the Public Review and below for l. 921 in Details. Offspring from different laying periods were aged in different bottles, further separated by the time at which they eclosed. They were then pooled for DNA extraction and library preparation by sex and a binary early or late eclosion time. This data was present in the “D. mel. Sample Size” column of supplemental tables S6 and S7 (now S7 and S8), but we have added and referenced a new table to specifically collate the sample sizes of different experimental stages, table S1. Now referenced at lines 555, 638, 975, 1057.

(5) The caption of figure 9 and the discussion of its results should be clear and explicit about the fact that "adult offspring" in Fig 9A and "female" and "male" refers to adults surviving to old age (whereas "parental" in Fig 9A refers to young adults in their reproductive prime. This has consequences for the interpretation of the difference between "parental" and "adult offspring", as it combines one generation of usual selection as it occurs under the conditions of the lab culture (young adult at generation t -> young adult in generation t+1) with an additional step of selection for longevity. Thus, a marked change in allele frequency does not imply that the "parental" frequency does not represent an equilibrium frequency of the inversions under the lab culture conditions. Furthermore, it would be useful to state explicitly that Figure 9B represents the same results as figure 9A, but with the aged offspring split by sex.

Figure caption edited to provide further clarity on the age of cohorts and presented data, along with the relevant results section (2.3) referencing this figure.

We avoid making any statements about the equilibrium frequencies of inversions under lab conditions, and whether or not any step of our experiment reflects such equilibria, because our investigation does not rely upon or test for such conditions. Instead, our analysis focuses on whether inversions have contrasting effects (as indicated by frequency changes that are incompatible with neutral sampling) between different life history components. Under our model, such frequency reversals might be detectable both at equilibrium balanced inversion frequencies and also at frequencies some distance away from equilibria. We have now clarified this point at l. 970-972.

Details:

l. 211: this should be modified as male-only costs are now included.

Edited. “survival likelihood (of either or both sexes).”

l. 343: misplaced period

Edited.

l. 814: "We confirmed model predictions...": This sounds like it refers to an empirical confirmation of a theory prediction, but I think the authors just want to say that their simulations predicted antagonistic variants can be maintained at an intermediate equilibrium frequency. So the wording should be changed to avoid ambiguity.

Edited. Now line 869.

l. 853: How can a genome be "empty"? Do the authors mean an absence of any polymorphism?

Edited to: “In SAIsim, a population is instantiated as a python object, and populated with individuals which are also represented by python objects. These individuals may be instantiated using genomes specified by the user, or by default carry no genomic variation.” Lines 913ff.

l. 853: I do not see this diagramed in Figure 5

Apologies, fixed to Fig. 2

l. 864: is crossing-over in the model limited to female gametogenesis (reflecting the Drosophila case) or does it occur in both sexes?

There is a variable in the simulator to make crossover female-specific. All simulations were performed with female-only crossover. Edited for clarity. “While the simulator can allow recombination in both sexes, all simulations presented only generate crossovers and gene conversion events for female gametes, in accordance with the biology of D. melanogaster.” Lines 928-929.

l. 906: "F2" is ambiguous; does this mean that the mix of lines was allowed to breed for two generations? Also, in other places in the manuscript these flies appear to be referred to are "parental". So do not use F2.

Edited, F2 language removed and replaced with being allowed to breed for two generations. Now lines 967ff.

l. 910: this is incorrect/imprecise; what can be inferred is the frequency of the inversions in male gametes that contributed to fertilization. This would correspond to the frequency in successful males only if each successful male genotype had the same paternity share.

Edited, now “Since no inversions could be inherited through the mothers, inversion frequencies among successful male gametes could be inferred from their pooled offspring.” Now line 994.

l. 912: "without a controlled day/night cycle" meaning what? Constant light? Constant darkness? Daylight falling through the windows?

Edited to “Unless otherwise noted, all flies were kept in a lab space of 23°C with around a degree of temperature fluctuation and without a controlled day/night cycle. Light exposure was dependent on the varying use of the space by laboratory workers but amounted to near constant exposure to at least a minimal level of lighting, with some variable light due to indirect lighting from adjacent rooms with exterior windows.” Now lines 1007-1010.

l. 921: I cannot parse this sentence. Were the offspring isolated as virgins?

No, the logistics of collecting virgins would have been prohibitive, and it did not seem essential for our experiment. Hopefully the edits to this section are clearer, now lines 978ff.

Version 3

Abstract

Abstract

Chromosomal inversion polymorphisms can be common, but the causes of their persistence are often unclear. We propose a model for the maintenance of inversion polymorphism, which requires that some variants contribute antagonistically to two phenotypes, one of which has negative frequency-dependent fitness. These conditions yield a form of frequency-dependent disruptive selection, favoring two predominant haplotypes segregating alleles that favor opposing antagonistic phenotypes. An inversion associated with one haplotype can reduce the fitness load incurred by generating recombinant offspring, reinforcing its linkage to the haplotype and enabling both haplotypes to accumulate more antagonistic variants than expected otherwise. We develop and apply a forward simulator to examine these dynamics under a tradeoff between survival and male display. These simulations indeed generate inversion-associated haplotypes with opposing sex-specific fitness effects. Antagonism strengthens with time, and can ultimately yield karyotypes at surprisingly predictable frequencies, with striking genotype frequency differences between sexes and between developmental stages. To test whether this model may contribute to well-studied yet enigmatic inversion polymorphisms in Drosophila melanogaster, we track inversion frequencies in laboratory crosses to test whether they influence male reproductive success or survival. We find that two of the four tested inversions show significant evidence for the tradeoff examined, with In(3R)K favoring survival and In(3L)Ok favoring male reproduction. In line with the apparent sex-specific fitness effects implied for both of those inversions, In(3L)Ok was also found to be less costly to the viability and/or longevity of males than females, whereas In(3R)K was more beneficial to female survival. Based on this work, we expect that balancing selection on antagonistically pleiotropic traits may provide a significant and underappreciated contribution to the maintenance of natural inversion polymorphism.

Introduction

The Enigmatic Prevalence of Inversion Polymorphism

Inversions, DNA segments in reversed sequence order relative to the ancestor, have been a focus of population genetic study for their direct or indirect effects on recombination in natural populations for more than a century. In their most dramatic effect on meiosis, inversions generate aneuploid gametes when a crossover occurs between the interior of an inverted region and its non-inverted homolog (White 1973). As aneuploidy is often fatal in embryos or severely deleterious, this may translate to a fecundity cost for a heterokaryotypic parent: a parent heterozygous for an inversion arrangement. However, depending on a species’ reproductive biology, such reproductive fitness effects may be mitigated if the zygotes terminate early or before significant parental resource investment (Charlesworth 1994, Munasinghe and Brandvain 2023). Alternatively, crossovers may be inhibited directly in heterokaryotypes around the inversion by mechanisms similar to crossover interference (Koury 2023), avoided by achiasmy in one sex or all, or aneuploid gametes may be segregated to polar bodies in female meiosis (Sturtevant 1921; Sturtevant and Beadle 1936; Krimbas and Powell 1992; Coyne et al. 1993; Gong et al. 2005). Regardless of the differences between these scenarios, effectively fewer recombinant offspring are produced in heterokaryotypes.

Inversions may also have direct effects on fitness when disrupting a gene, a regulatory element, or other genomic structure like a topologically associated domain (Lupiáñez et al. 2015; McBroome et al. 2020). Given that another potential effect of inversions in heterokaryotypic individuals is the deleterious generation of aneuploid gametes, it seems reasonable to expect inversion rearrangements to confer either deleterious or neutral fitness consequences in the absence of association with other variants. However, when a new inversion mutates, it samples a single haplotype from a population, which is likely to have some non-neutral fitness effect, even if mild. Greater natural variation in fitness within a population increases the likelihood of sampling haplotypes with exceptionally deleterious or beneficial associations of alleles, even when there are no other effects on fitness (Berdan et al. 2023).

In contrast to the above predictions, a considerable number of well-studied inversion variants are maintained at an intermediate frequency within a population, presumably due to the indirect effect inversions have on reducing recombination and maintaining fit haplotypes that are themselves under selection (reviewed in Berdan et al. 2023). The evolutionary forces maintaining both haplotypes are thought to be quite varied between populations and arrangements, but a common first hypothesis for a frequently sampled inversion is that local adaptation in one region or environment is balanced with migration from another (Kirkpatrick and Barton 2006; Charlesworth and Barton 2018). This association may be maintained more strongly when the alleles have beneficial epistatic interactions, and assumed beneficial epistasis between loci is often paired with another mechanism of balancing selection operating at a locus to explain inversion (Dobzhansky 1949, 1950, 1970; Schaeffer et al. 2003; Hoffmann and Rieseberg 2008). However, unless migration rates are very high, local adaptation models predict inversions to be at high frequency in their favored geographic range but at low frequency elsewhere, and many inversions do not have sampled locations at which they reach very high frequency.

Alternately, chance associations of both an inversion and its standard arrangement counterpart with different recessive deleterious alleles may mimic the fitness profile of a single overdominant locus (sometimes termed associative overdominance), where heterozygotes carrying both main haplotypes experience none of the deleterious effects of a homozygous recessive across the region (Sturtevant and Mather 1938; Frydenberg 1963; Zhao and Charlesworth 2016; Gilbert et al. 2020). However, recent theory and simulation suggests that this dynamic is uncommon unless an arrangement polymorphism is maintained at intermediate frequency under some other selective force, and the population scaled recombination rate between arrangements is low (Charlesworth 2024).

Balanced polymorphic loci may be maintained in other ways, under various models and selective pressures. For example, multiple overdominant loci that are each most beneficial as heterozygotes may benefit from linkage maintaining complementary haplotypes (Sved 1968; Charlesworth and Charlesworth 1973; Charlesworth 1974). Several recent studies have proposed that functionally overdominant loci may be more common than previously considered. Specifically, the dominance relation between alleles at a locus may change between different sexes, seasons, life history stages, or other trait or fitness challenge contexts, generally termed ‘reversal of dominance’, which in some cases can lead heterozygotes to have greatest average fitness (Rose 1982; Wittman et al. 2017; Grieshop and Arnqvist 2018; Connallon and Chenoweth 2019). Other specific biological dynamics may predict the maintenance of inversion polymorphism in particular ways. For example, inversions may also link drivers and enhancers in meiotic drive systems (Babcock and Anderson 1996; Mroczek et al. 2006; Courret et al. 2019), and these drive-associated inversions may persist at equilibrium frequencies in the presence of repressors (e.g. Bastide et al. 2022).

In natural populations, it can be difficult to distinguish among these alternate hypotheses. They are not always mutually exclusive, and the selective pressures acting upon an inversion may vary across space and time. The frequency patterns of many inversions, including a number in Drosophila melanogaster, remain poorly explained.

Inversions in Drosophila melanogaster

D. melanogaster is a widely studied organism in many fields of biology, including evolutionary and population genetics. Multiple polymorphic inversions segregate at meaningful frequencies in natural populations of D. melanogaster (Lemeunier and Aulard 1992; Aulard et al. 2002; Kapun et al. 2016; Kapun and Flatt 2019), but even in this model system, it is not entirely clear which population genetic processes are most responsible for shaping their frequencies. In this species, achiasmy in males (Morgan 1910; John et al. 2016) and a lack of evidence for female fecundity costs for most inversions (Sturtevant 1921; Sturtevant and Beadle 1936; Coyne et al. 1993; Gong et al. 2005) suggest that most paracentric inversions in D. melanogaster do not generate aneuploid gametes at any appreciable frequency. Further, genome editing tools have allowed for the creation of synthetic inversions, revealing that the tested structural rearrangements are not themselves responsible for the gene regulatory changes that are associated with natural inversions in D. melanogaster (Said et al. 2018). Instead, given that breakpoints of common D. melanogaster inversions occur disproportionately in regions that confer greater recombination suppression (Corbett-Detig 2016), D. melanogaster inversions would appear to be maintained primarily for the sake of maintaining linkage associations between distant variants.

D. melanogaster originated in southern-central Africa, in seasonally dry Miombo and Mopane woodlands, and (potentially after adapting to a human commensal niche) expanded out across much of Africa and into the Middle East around 13,000 years ago, and into Europe around 1,800 years ago (Sprengelmeyer et al. 2020). With an ancestral Ne of around 2 million (Sprengelmeyer et al. 2020), D. melanogaster populations have high levels of standing diversity and high population recombination rates. Many D. melanogaster inversions have geographic clines repeated between several colonized regions (Mettler et al. 1977; Reinhardt et al. 2014; Kapun et al. 2016), and inversions often show geographic differences in frequency that exceed genome-wide average differentiation between the same populations (e.g. Pool et al. 2017). It has therefore been suggested that local ecological adaptation may shape the distribution of D. melanogaster inversions (Kapun et al. 2016; Kapun and Flatt 2019). Indeed, it is difficult to explain their geographic differentiation without some type of spatially-varying selective pressures.

Yet, it is unclear whether ecological adaptation is a sufficient explanation for D. melanogaster inversion frequency patterns, for two reasons. First, most of the common inversions become less frequent as one samples populations farther from the sub-Saharan ancestral range (or similar warm environments) into potentially more challenging, colder high latitude and altitude environments (Aulard et al. 2002; Kapun et al. 2016; Lack et al. 2016a; Pool et al. 2017; Kapun and Flatt 2019) with In(3R)Mo representing an exception to this pattern (Kapun et al. 2014; Kapun et al. 2016). As these inversions are all derived (being absent from closely related species), local adaptation theory would predict that they arose to protect local adaptation against ongoing migration, more likely in the novel environment if it is experiencing high levels of immigration (Kirkpatrick and Barton 2006). It is unclear why inversions would be needed to adapt to the species’ ancestral environments, or why ancestral “standard” arrangements would have an ecological advantage in derived environments that are very different from the ancestral range.

Second, even in the locations where inversions would hypothetically be ecologically advantageous, they rarely surpass intermediate frequencies (Lemeunier and Aulard 1992; Pool et al. 2017; Kapun and Flatt 2019; Sprengelmeyer et al. 2020). Under a local adaptation model, that observation would require a secondary explanation that resists fixation, such as very high levels of long distance gene flow or the fixation of recessive deleterious variants on inverted haplotypes (Ohta and Kimura 1970; Zhao and Charlesworth 2016; Gilbert et al. 2020). While migration may well limit inversion frequencies in some populations, we note that some putatively adaptive Single Nucleotide Polymorphism (SNP) variants do achieve much greater frequency differentials between populations (da Silva Ribeiro et al. 2022), and thus we would have to suppose that inversions are under relatively weaker selection to explain their lesser differentiation. Regarding the hypothesis of high recessive load, we note that D. melanogaster populations do generally contain substantial recessive genetic load (Greenberg and Crow 1960), and it has been speculated that associative overdominance between sampled backgrounds might boost inversion frequencies in D. melanogaster populations during founder events (Pool et al. 2012) or in isolated wilderness environments (Sprengelmeyer et al. 2020). However, it seems likely that in large human-commensal populations, D. melanogaster inversions would escape from linked deleterious variants through gene conversion or double crossover events with standard chromosomes, in light of both the age of most of these inversions (Corbett-Detig and Hartl 2012) and the very high population recombination rates of D. melanogaster. Homozygous lines for each common inversion also appear at appreciable frequency in inbred lab strains, demonstrating a lack of perfect linkage of the inversions with recessive lethal or infertile variants (e.g. Lack et al. 2016a).

While local adaptation, gene flow, and associative overdominance may all contribute to geographic variation in inversion frequency among D. melanogaster populations, it is not clear that these processes are collectively sufficient to explain observed patterns. Alternatively, the allele frequencies and diversity patterns of at least some of these inversions might be primarily shaped by some form of balancing selection acting on inversion-linked variation, in which the balanced frequency is dependent on the environment (Kapun et al. 2023). Balancing selection on inversions has received increasing attention in recent literature (Wellenreuther and Bernatchez 2018; Faria et al. 2019). In Drosophila, inversions have been associated with balancing selection due to seasonal, temporally fluctuating selection (Machado et al. 2021). However, the potential for temporally varying selection to maintain stable balanced polymorphisms on its own is somewhat limited (Gillespie 1988). Further, most common inversions maintain relatively high polymorphic frequencies across broad geographic ranges without obvious shared seasonal selective pressures, from tropical and subtropical dry forests to temperate locations with harsh winter conditions (Lemeunier and Aulard 1992). Therefore, we predict that the frequencies of some of these inversions could be in part explained by a mechanism of balancing selection that would operate even within a temporally and spatially homogenous population. Below, we propose such a model in which alleles that contribute to competitive mating display simultaneously contribute antagonistically to survival to reproductive maturity, a tradeoff that offers a mechanism by which the life-history characteristics of a population could generate epistatic balancing selection.

Sexually Antagonistic Polymorphism in D. melanogaster

Among mechanisms of selection, sexual selection is both prevalent, and provides a simple mechanism for frequency dependence (O’Donald et al. 1997). Alleles that benefit an individual in competitive mate choice systems often decrease in fitness as the beneficial allele increases in frequency, as display success depends on the relative quality of a male among local competitors, and the more common a display-enhancing allele is, the fewer competitors it succeeds over. Albeit, the direction and magnitude of the frequency dependent effect can depend on the way mate choice is parameterized even among “best of n” models, as well as the strength of selective versus stochastic effects (O’Donald 1973; O’Donald et al. 1997). Combinations of display-enhancing alleles at multiple loci may similarly gain a disproportionate benefit together, as male mating success in competitive display systems can be skewed towards a few successful individuals (Bateman 1948; Jones et al. 2002; Tatarenkov et al. 2008, though see Gowaty et al. 2012; Hoquet et al. 2020). In a mate choice system such as a best of n model in which a female chooses the highest quality male encountered, a marginally higher quality male similarly outcompetes either a marginally lower or an exceptionally worse male. So, the last few alleles to push display into the top quantile should contribute a significantly greater fitness increase to the genotype than the first few display alleles. This interaction of alleles is a form of epistasis - a non-independent fitness interaction between loci - that is emergent and non-specific to the genetic identity of the variants. In other words, the alleles contribute independently to the generation of a trait, and the fitness of the individual is a nonlinear function of that trait (Blows and Brooks 2003; Rest et al. 2013; Otwinowski et al. 2018). This epistatic interaction suggests a potential for indirect selection on the recombination-suppressing effects of inversions, in order to maintain combinations of sexually antagonistic variants that have the greatest fitness when carried together. This tendency of selection to favor the reduction of recombination when the haplotypes present are at a fitness optimum is well studied for recombination modifiers including inversions, and has been termed the ‘reduction principle’ (Feldman 1972; Feldman and Balkau 1973; Feldman et al. 1980; Altenberg and Feldman 1987; Altenberg et al. 2017; see Altenberg and Feldman 1987 for a direct treatment of inversions).

While the emergent epistasis generated by a mate choice system may favor linkage of variants, alleles that contribute only to a sexually selected display are not expected to remain as balanced polymorphisms, rather they are expected to fix, albeit perhaps slowly as their fitness benefits diminish at high frequencies (Hazel 1943; Dickerson 1955; Charlesworth and Hughes 2000). However, if these alleles simultaneously contribute deleteriously to other components of fitness, they are pleiotropic (Stearns 2010), and given the sex-limited benefits of the display they would likely also be sexually antagonistic (Rowe et al. 2018), whereby traits increase the fitness of one sex but decrease fitness if found in the other. Pleiotropy can result from traits that are inherently physiologically coupled, from resource allocation restraints, or from the specific role or behavior of genes. Pleiotropy can further be sexually antagonistic when the traits in the tradeoff are differentially important to different sexes, which may be common given the differing fitness requirements of the sexes in a population (Cox and Calsbeek 2009; Connallon and Clark 2014; Connallon and Chenoweth 2019). It should be noted that a model of two pleiotropically determined traits each with potentially sex-specific fitness can be treated separately from a model where one trait has sexually antagonistic fitness effects.

D. melanogaster mating involves a male display performance and female acceptance or rejection. A number of experimental evolution studies (Rice 1992; Rice 1996; Audet et al. 2024) and genetic studies (Innocenti and Morrow 2010; Cheng and Kirkpatrick 2016; Glaser-Schmitt et al. 2023) have confirmed the prevalence of sexually antagonistic loci in D. melanogaster. Pleiotropic alleles that trade off between survival and male display quality represent one clear potential case of antagonism to consider for a model of balancing selection, and one highly relevant to the biology of D. melanogaster. Many such antagonistically pleiotropic scenarios are possible under different selective tradeoffs, and indeed some have been proposed for other model inversions with varying mechanisms of maintenance (Betrán et al. 1998; Pearse et al. 2019; Mérot et al. 2020; Pei et al. 2023; see Discussion), but we focus on a specific scenario of sexual antagonism here.

Given the relevance of sexual selection and sexual antagonism to Drosophila evolution, we propose that polymorphic D. melanogaster inversion and standard arrangements may help maintain haplotypes of alleles that specify opposite ends of a sexually antagonistic tradeoff that is subject to balancing selection and for which intermediate states are inherently disfavored. We first explore this model through forward simulation, to establish that it may in theory arise in a population and maintain inversion polymorphism over a significant number of generations. Then we examine the results of an experiment with an outbred laboratory population of D. melanogaster, to test whether common inversions show changes in frequency over the course of a generation in a manner consistent with a trade-off between survival and male reproductive success.

Results

A Model of Sexually Antagonistic Inversions

Here, we consider a population featuring a competitive mate choice system that results in a skewed distribution of reproductive success among males. Within this population, we model a genomic region containing a number of linked loci, each with alternate antagonistic, pleiotropic alleles that trade off between male display success and survival likelihood (of either or both sexes). Critically, we do not directly model any epistatic interactions among loci affecting survival or display traits, nor do we invoke dominance in our diploid model, so any such fitness dynamics emerge from the modeled population processes.

If mate choice is highly competitive, such that each female can choose amongst several or many males, the variance of offspring number among males (Figure 1A) may be much higher than for females - yielding a distribution of male reproductive output that follows a relatively convex trend, in which most males produce few offspring. Conversely, superior display quality can enable males to mate successfully much more often than other display phenotypes they compete against (Figure 1A). An increase in the display of a male with low quality results in a smaller change in number of offspring than an increase in display of a male with high quality (change ii in Figure 1A), due to the reproductive skew towards highly competitive individuals. So, there is an emergent non-additive fitness effect, where multiple variants increasing the reproductive rank of an individual together have a larger fitness benefit than the combined effects of each substitution separately - a positive, synergistic epistatic effect. This effect occurs whether the difference is from substituting alleles at two loci, or from substituting two homologous alleles at a single locus in a diploid or polyploid organism, which would give an emergent dominance effect. Additionally, fitness conveyed by an allele favoring display quality is also frequency-dependent: since mating success depends on the frequency of comparable display qualities of other males in a best of n model, the relative advantage of a display trait can diminish as more males carry it, for both haploid and diploid populations (Figure 1B, male fitnesses; O’Donald 1973; O’Donald et al. 1997). If a variant only conveys benefit to mate competition success, it is expected to fix given enough time, as even if its evolution is nearly neutral at high population frequency, selection will push the frequency back up if it drifts down. Whereas, if the variant is antagonistically pleiotropic for a trait not involved in competition between males, such as survival in the face of climate, pathogens, or predators, these fitness costs may typically not be frequency dependent. If the benefit of an antagonistic display-favoring allele overcomes the pleiotropic cost when the variant is rare, its diminished benefit when at higher frequency may reverse its net benefit, while at some intermediate frequency, we predict that these fitness effects should be equal and an equilibrium frequency should be reached and maintained (Figure 1B). In the absence of other forces, this dynamic is expected to maintain the allele under balancing selection indefinitely. We note that this dynamic can be described as a “negative frequency dependence” of the fitness component for mate choice alone, as the selection coefficient of the favored allele may decrease towards zero while remaining positive, though some authors instead use negative frequency dependence to specifically describe fitness functions in which all alternate alleles are of selective benefit when rare (e.g. Charlesworth and Charlesworth 2010). In addition to the relationship between sex averaged fitness and frequency, the contrasting sex-specific fitnesses of antagonistic alleles under this model could further enhance the potential for balanced polymorphism.

Conceptual representations of the proposed model of inversion-associated balanced sexual antagonism.

(A) Hypothetical skewed distribution of male mating success, yielding a greater variance in reproductive success for males than for females, such as expected under a “best of n” mate choice model. (B) Fitnesses of display-favoring alleles under sexually antagonistic balancing selection, illustrating a pleiotropic variant that should rise in frequency when rare, but decline if frequency exceeds the balanced equilibrium value. (C) Fitnesses of display-favoring alleles at two haploid loci under the same model, illustrating synergistic epistasis between displaying-favoring alleles. (D) A hypothetical trajectory of four such mutations, in which B outcompetes A, an inversion links B and C to create a more strongly display-favoring haplotype, and then the addition of D to that haplotype furthers the accumulation of antagonistic variants, reaching an equilibrium frequency while displacing less extreme display-favoring haplotypes.

We further predict that multiple such balanced pleiotropic loci should generate interesting emergent behavior. Because of the emergent synergistic epistasis from mate competition (see discussion of Figure 1A above), the haplotype with the greatest male display benefit should increase in frequency relative to other haplotypes until reaching an equilibrium at which its relative advantage in mate choice is reduced enough to be balanced by the pleiotropic, frequency-independent, non-epistatic survival cost (Figure 1B). The haplotypes and their linked alleles conveying intermediate phenotypes will have less than proportional benefits during mate choice due to the convex relationship between quality and success described above (Figure 1A), but a larger, proportional cost to survival against non-intraspecific survival challenges, giving them lower average fitness in the equilibrium state than either of the extreme haplotypes (Figure 1C). Therefore, recombination between the two extreme haplotypes will generate less fit haplotypes in the gametes, and a local modifier of recombination that reduces recombination and its associated costs (such as an inversion) will be favored by indirect selection (Altenberg and Feldman 1987; Otto and Lenormand 2002). And so, selection will favor the association of these haplotypes with the alternate standard and inverted karyotypes, which generate fewer recombinant gametes from a heterokaryotypic parent. We further predict that in a population that is close to the balanced equilibrium state for such a pleiotropic trade-off, the population of adults will contain a higher frequency of survival-focused haplotypes than the population of zygotes, but the display-focused haplotypes will increase in frequency in zygotes relative to adults, in line with the life stage-specific fitness differences between these haplotypes.

While Figure 1 illustrates a haploid scenario for the sake of simplicity, this model extends naturally to a diploid case. The main difference is that heterokaryotypic diploids – those heterozygous for the arrangement – would have intermediate trait values and be less fit than either homokaryotype, for the same selective reasons that an intermediate haplotype with two alleles of opposing effects at different loci is least fit in a haploid context. This is therefore, interestingly, an example of a scenario in which the two most common haplotypes in the population can exhibit underdominant fitness, and yet may still be maintained at a stable balanced equilibrium. By comparison, the simplest models of underdominant fitness predict an unstable equilibrium that disfavors polymorphism (Gillespie 2004, Charlesworth and Charlesworth 2010), a characteristic shared with non-frequency-dependent models of disruptive selection. The cost of generating underdominant offspring might be compensated for by a strong co-directional dominance of the variants involved, such that heterokaryotypic individuals exhibit a phenotype similar to one of the homokaryotypes. However, dominance is not a necessary feature of our model, and it is not invoked in our simulations. Whereas, investigating a wider range of genetic architectures would be a potential direction for future study.

Simulation Results

We investigated the predictions of the above model of inversion-associated balanced sexual antagonism using a purpose-built forward simulation program, “Sexually Antagonistic Inversion simulator” (SAIsim; see Materials and Methods). This simulation program incorporates the generation of new nucleotide and inversion polymorphisms by mutation, where each nucleotide mutation independently draws a positive or negative effect on survival and on male display. It also simulates crossing over and gene conversion, and inherently the effects of drift (Materials and Methods). SAIsim includes gene conversion between all homologous sites regardless of karyotype, and models crossing over in heterokaryotypic individuals from even numbers of crossover events within the inverted region. It explicitly models mate selection, in that for each female selected to reproduce, a set number of ‘encountered’ males are randomly selected, and the observed display value of each male for this competitive event is assigned based on the sum of the male’s genetic effects and a noise effect resampled at every encounter. The male with the highest observed display value then succeeds in this mating (Figure 2).

The layout of a single simulated generation.

Simulated individuals each reach the reproductive stage with probability in proportion to the product of the survival effects of their alleles. To generate each offspring in the next generation, a surviving female is first randomly sampled (with replacement) to be the mother. Then a specified number of surviving males are randomly sampled to generate the pool of males encountered by the sampled female for the best of n mate choice - without replacement for each pool and with replacement between pools. Unless otherwise noted, we defaulted to 100 encountered males in simulations presented here. For each encounter pool, each male is assigned an observed display quality as the sum of the display effects of its alleles, plus a normally distributed noise effect for each encounter. The highest quality male in the encounter pool is selected as the father, and the offspring genome is generated. The female gamete is generated with crossover and gene conversion events. Odd numbers of crossovers in the interior of an inversion are resampled from the same parent, to model the biology of an organism that eliminates aneuploid gametes during oogenesis or exhibits reproductive compensation. D. melanogaster males do not cross-over, and so males do not recombine in the simulations presented.

With this simulator, we first confirmed that some individual variants with an antagonistic pleiotropy between survival (initially of both sexes) and male reproduction can generate stable selectively balanced polymorphisms in silico. We simulated a single locus in diploid populations of N = 1000, polymorphic for an allele of a specified survival cost and display benefit at an initial frequency of 0.5, with the alternate allele having no trait effect. For each female’s mating competition, 100 males were sampled, though see Figure S1 for plots with varying encounter number. After 20N generations, we calculated the proportion of simulations retaining polymorphism and the average final frequency of the variant across all simulations, revealing that there is a range of paired trait effects for which the mutation consistently remained polymorphic, with intermediate mean frequency, after 20N generations (Figure 3) – implying the action of balancing selection. Alleles with small effect sizes had only a narrow band of selective maintenance, outside of which they were either fixed or lost, while large-effect alleles remained balanced across a larger area of simulated parameter space. While the overall balanced range is somewhat narrow, such variants may persist under balancing selection indefinitely in the absence of other forces, and many small-effect alleles in linkage blocks may confer larger effects in synergy. In a scenario where only females pay the survival cost, the balanced parameter space of antagonistic trait effects is broader and shifted somewhat (Figure S1). In a scenario where only males pay the survival cost, and variants are neutral in females, essentially no simulations retain polymorphism (Figure S1). The number of males sampled to assess mate competition for each offspring also has a significant effect, with larger values giving broader polymorphic ranges, though simulations with costs only in females show robust polymorphism even when only sampling two males for each competition (Figure S1).

Simulations show that a balanced equilibrium frequency exists for certain sexually antagonistic variants.

The proportion of simulations retaining polymorphism (top) and the average final frequency (bottom) are plotted for a single locus at which there is one antagonistic allele with the indicated effects reducing survival proportion (of both sexes) and increasing male reproductive quality score, initiated at equal frequency with an alternative allele that has no such effects. Larger values of both effect sizes are plotted in the left panels, with values incrementing by 0.025, and smaller effect sizes (a subset of the larger) are plotted in greater detail in the right panels, with values incrementing by 0.005. Reproductive quality contributes to male success in a best of 100 mate competition with an added normally distributed noise effect of standard deviation 1 (Figure 2; Materials and Methods). See Figure S1 for plots with sex-specific costs and differing encounter numbers in the best of n mate competition. Most parameters result in either all simulations retaining polymorphism (purple), or else all simulations losing polymorphism (white). However, simulations retaining polymorphism have a broad range of equilibrium frequencies (shades of green). For each parameter combination, 500 replicates were simulated for a diploid population of 1,000 individuals evolving for 20,000 generations, with further details as indicated in the Materials and Methods.

We then tested the behavior of two linked loci, both polymorphic for an antagonistic and non-antagonistic allele. We identified two antagonistic variants previously found to undergo balancing selection separately, and performed simulations with varying genetic distances between them (Figure 4). Crossing over and gene conversion were simulated, with gene conversion occurring for each heterozygous variant at a probability of 10-2 per female meiosis per site, with random direction. This behavior was intended to be a conservative approximation of the number of gene conversion events affecting a given site in D. melanogaster, scaled to simulations of N=1000 (Comeron et al. 2012). The stronger antagonistic variant (modelled at fixed position 0.225M) persisted in all simulations, in line with expectations, and so we focus on whether the weaker variant (at varying positions) also persisted with it. As seen in the upper panel of Figure 4, the weaker variant (pale red line) was retained at close linkage, as the haplotype containing both antagonistic variants acts as a single larger effect supergene. The haplotypes with intermediate trait values were less fit, leaving less offspring per haplotype across simulations (Figure S2A), whereas the extreme haplotypes were maintained at much higher frequency (Figure S2B), generating linkage disequilibrium between display-favoring variants (r2=0.2139 across simulations in which the smaller effect variant was positioned at 0.250M). This demonstrates the synergistic epistasis proposed in our verbal model. The maintenance of haplotypes AB and ab under selection while Ab and aB are selectively removed relies on the lesser relative fitness of Ab and aB. And since survival values are multiplicative, this additional contribution must come from the mate success of AB being disproportionately larger than Ab or aB. As the weaker variant was then moved farther from the stronger variant, recombination increasingly generated unfit intermediate genotypes compared to the extremes of the reproduction-survival tradeoff, and the stronger variant outcompeted the weaker variant among these haplotypes despite the epistatic benefit from associating the two variants together, leading to the loss of the weaker variant. In the scenario examined by Figure S2, the differences among haplotypes in sex-averaged fitness are modest, whereas the sex-specific fitness effects are more dramatic. These results could indicate that in addition to the frequency-dependent dynamics proposed above, the contrasting fitness effects in females and males may also play an important role in maintaining balanced polymorphism, particularly when near frequency-dependent balance (Connallon and Clark 2014; Kaufmann et al. 2023).

Simulated inversions may persist as polymorphisms when linked to sexually antagonistic, pleiotropic variants.

In turn, the presence of inversions facilitates the accumulation of antagonistic variation, demonstrating synergistic epistasis. Rates of long-term persistence are shown for simulations with two defined antagonistic variants alone (left panels) and with the additional presence of a defined inversion (right panels). To focus the potential addition of a second linked variant to an antagonistic haplotype, we begin with the stronger variant already at an intermediate frequency in the population (0.5). The weaker antagonistic variant and the inversion either both start at 0.5 frequency as well (upper row) or both start at a lower frequency of 0.05 (lower row), with all variants initially in linkage equilibrium. The 1 Morgan chromosomal segment is diagrammed above the plots, with the inversion breakpoints marked (0.21M, 0.55M), chosen to include space across and beyond the inverted region. The proportion of simulations in which the second, variably positioned weaker antagonistic variant (survival and reproductive values 0.86, 0.12) maintains polymorphism is shown (light red line), dependent on the recombination distance from a stronger allele (0.75, 0.3) at fixed position 0.225M, which always persists (dark red line at 1 across the top of each plot). In the right panels, the proportion of replicates in which the inversion is retained is also plotted (blue line). Values for each parameter combination were averaged over 1000 replicate simulations, for populations of 1000 diploid individuals run for 20N generations. Gene conversion occurs for each heterozygous variant at a probability of 10-2 per female meiosis with random direction. Crossover arrangements generating aneuploidy, where there are an odd number of crossovers within the inverted region of a heterokaryotypic parent, are resampled from the same parent to represent removal to polar bodies or reproductive compensation.

We then examined the same spacings between antagonistic variants across a region containing a polymorphic inversion, with its left breakpoint close to the stronger variant, and the right breakpoint within the range of the positions in which the weaker variant was simulated (Figure 4, lower panel). When the inversion was added, the weaker variant was then retained when the two antagonistic variants were relatively more distant, due to association of the weaker variant with the right-hand inversion breakpoint (pale red line), at which point this association also preserved the inversion itself in a polymorphic state (light blue line). While less predictable, a weaker antagonistic variant near the middle of the inversion also appeared to maintain polymorphism more frequently in systems starting with an inversion than without, i.e. across approximate positions 0.30-0.45M in Figure 4, in spite of its weaker linkage with the inversion itself due to double cross-over events. The inversion was not maintained when the weaker antagonistic variant was within the inversion but tightly linked to the stronger variant (see positions 0.22-0.25M); here there is little crossing-over for an inversion to suppress, and so the effect of the inversion approaches neutrality. Interestingly, there was a small window of weaker variant position in which the inversion remained polymorphic despite the already tight linkage between SA variants, where the inversion seemingly does not modify recombination much (see positions 0.2-0.22M). Here, the weaker antagonistic variant is more tightly linked to the breakpoint than to the stronger variant in this range, and presumably benefits more consistently from the crossover reduction. The maintenance of the weaker variant drops more rapidly outside of the inversion than over similar distances in the simulation without inversions (positions 0-0.2M, bottom versus top panel). This seemingly counter-intuitive result may stem from crossover events that result in aneuploid gametes being resampled in our simulations, which effectively increases the crossover rate outside the inversion breakpoints, a behavior which should be biologically reasonable in taxa with aneuploid chromosome elimination or reproductive compensation. So, for a weaker variant outside the left inversion breakpoint, the presence of the inversion can increase the crossover rate between the two antagonistic variants, and thus reduce the probability of the weaker variant surviving by retaining linkage to the stronger one.

In further simulations, we began with a population initially free of variation, in which sexually antagonistic mutations arise stochastically. To assess differences between longer or shorter antagonistic regions, or higher or lower crossover rates, we modeled populations with parameters scaled to either 1 Mb or 10 kb, using estimates from Comeron et al. (2012) and Huang et al. (2016). The mutations had pleiotropic effects increasing reproductive quality score and decreasing survival proportion, each drawn sampled independently (Materials and Methods).

Depending on the effect scores drawn, a mutation could be broadly advantageous (potentially leading to a selective sweep), broadly deleterious, or potentially subject to balancing selection (Figure 3, Figure S1). These mutations arose at a rate equal to one thousandth of the scaled single nucleotide mutation rate given in Huang et al. (2016). Inversions (of random lengths and positions) also arose stochastically, at a hundredth of the rate of antagonistic variants. The frequency trajectories of antagonistic variants and inversions were tracked, and the simulation was run for 20N generations to allow near-equilibrium patterns to be established. These simulations demonstrated the potential of a population without initial genetic variation to establish alternate karyotypes which link sexually antagonistic, pleiotropic alleles and an inversion, to maintain polymorphism in a balanced manner, and then to subsequently accumulate significant antagonistic character (Figure 5A), although the degree of antagonism accumulated depends strongly on the rate of crossing over (relative to the occurrence of antagonistic mutations; Figure 5B).

Simulated populations with stochastic mutation of new inversions and sexually antagonistic variants generate persistent inversion polymorphisms and accumulate linked sexually antagonistic variants.

(A) In simulated populations with stochastic mutation of new variants having pleiotropic effects on survival and male display, levels of antagonism increased with time before stabilizing at an intermediate value (left panel). Whereas, simulations which also allowed the mutation of new inversions continued accumulating stronger antagonism throughout the simulated time interval (right panel). (B) In simulations which allowed the mutation of new inversions, a low crossover rate (left panel, 43.6 cM in females after scaling) allowed populations to accumulate larger differences between the most common karyotype and all others than when the rate of crossover was particularly high (right panel, 43.6 M in females after scaling). Gene conversion occurs at each heterozygous variant with a scaled probability of 0.1295 per heterozygous variant in both scenarios. Values given are the mean of 1000 replicate simulations of 1000 diploid individuals run for 20N generations.

The simulations that successfully maintained balanced antagonistic chromosomes also had an interesting distribution of allele and genotype frequencies of the balanced haplotypes at equilibrium. Without any selection, the inversion frequencies followed a pattern characteristic of the expected neutral site frequency spectrum (Figure S3). But strikingly, in parameter spaces where simulations with shared survival costs between sexes maintained polymorphism under balancing selection, inversion frequency distributions were tightly clustered around 0.25 and 0.75 among the zygote population (Figure S3, 6A). We found that this outcome was related to the genotype frequencies of reproducing adults. These simulations evolve toward strong antagonism, and it is predominantly males homozygous for the display-favoring haplotype (whether that is inverted or standard) who succeed in reproducing, and these males therefore contribute one copy of the display-favoring haplotype to each of their male and female offspring. However, this haplotype’s significant cost to viability and longevity would select for the heterokaryotypic females over the display-favoring homokaryotypes. Given that successful parents would predominantly be heterokaryotypic females and display-favoring homokaryotypic males, around three quarters of transmitted chromosomes have the display-favoring arrangement and one quarter does not, yielding the 0.25 and 0.75 zygote inversion frequencies seen in simulations. Although these are autosomal simulations, selection yields an inheritance pattern similar to that of a ZW sex chromosome system, where the Z chromosome is homokaryotypic in the fathers, and heterokaryotypic with the W chromosome in the mothers.

The sex-specific fitness dynamics present at equilibrium under this model also resulted in population-wide genotype frequencies that depart from standard Hardy-Weinberg expectations. If the preponderance of matings are between heterokaryotypic females and homokaryotypic display-favoring males, then homokaryotypes for the survival-focused haplotype will rarely be produced (as confirmed in Figure 6B). We note that in cases where the standard arrangement favors male display, then especially if antagonism is strong, this model could explain an observed deficiency or absence of inversion homokaryotypes, even if the inverted arrangement was not associated with any recessive deleterious load. Further, heterokaryotypes may become disproportionately frequent after viability selection (Figure 6B). And still, this enrichment for heterokaryotypes reflects neither the existence of recessive deleterious load associated with each arrangement, nor overdominance with respect to viability. Hence, our model may offer an alternative explanation for deviations from Hardy-Weinberg genotype frequencies observed for polymorphic inversions in natural populations.

Antagonism-associated karyotypes reach predictable equilibrium frequencies in simulations with antagonistic and inversion mutations, when survival costs are shared between sexes.

(A) Histograms of the number of karyotypic arrangements at a specified frequency across all simulation replicates in females (left) and males (right), colored by how the average survival effect of that arrangement ranks compared to the others in its population. Ranks are normalized to be relative to the median arrangement value, to better account for rare arrangements and ties. These simulations reach an equilibrium with two predominant arrangements, with the more survival-focused arrangement around an overall frequency of 0.25 and more common in females, and the more male-competition-focused arrangement around a frequency of 0.75 and more common in males. (B) Because successful males tend to be homozygotes and successful females heterozygotes, the less frequent of the two major arrangements (which tends to favor survival) has homokaryotype frequencies far below the Hardy-Weinberg expectation, and there is an excess of heterokaryotypes. These genotype frequencies shift between zygote (left panel) and adult (right panel) populations as the genotype affects survival.

We further ran simulations in which the survival costs were not shared, but instead specific to females, representing a distinct, stronger form of sexual antagonism. In these simulations, we surprisingly recovered three different equilibrium arrangement states (Figure 7). In 20.0% of simulations, we observed a state similar to the shared costs simulations presented in Figure 6, where we infer that fathers were homokaryotypic for a display-favoring arrangement, and mothers were heterokaryotypic and carried survival-focused arrangement that is observed in surviving but not reproducing males. For 37.0% of simulations, there were again two predominant arrangements, but with transmission patterns similar to an XY sex chromosome system rather than a ZW sex chromosome system. In these cases, we infer that the mothers were homokaryotypic for an X-like survival-focused arrangement, but the fathers were heterokaryotypic and carried a Y-like display-focused arrangement that was apparently lethal in females. In 40.6% of simulations, we instead detected three intermediate frequency haplotypes: with both Y-like and W-like arrangements restricted to reproducing fathers and mothers respectively, and a third shared arrangement with intermediate phenotypic effects (Figure 7).

In simulations with female-limited survival costs, three distinct outcomes are observed involving either two or three balanced haplotypes.

From simulations with randomly occurring antagonistic and inversion mutations, histograms of inversion frequencies across simulations for surviving females (left) and males (right) are partitioned into three categories based on sex-specific haplotype frequency outcomes. The “Favors Female” category (top) includes simulations in which there are exactly two arrangements with frequencies between 0.1 and 0.9 in both sexes (20.0% of replicates). The “Male-Specific” category (middle) includes simulations in which there are two arrangements with whole-population frequencies between 0.1 and 0.9, but one arrangement is absent in adult females (37.0% of replicates). The “Three Arrangement” category (bottom) includes simulations in which there are exactly three arrangements with whole-population frequencies between 0.1 and 0.9 frequency (40.6% of replicates). 2.3% of simulations did not fit in these categories and had only one predominant arrangement. As in Figure 6, inversions are colored by how the average survival effect of that arrangement ranks compared to the others in its population, and ranks are normalized to be relative to the median arrangement value, to better account for rare arrangements and ties.

Specific properties of our simulations may have influenced the occurrence of these three stable states under the female-limited cost model. To reduce the computational complexity of both crossovers and arrangement-specific trait values, the simulations do not allow two inversions on the same individual chromosome, whether overlapping or not, though two chromosomes sampled from the population may have distinct inversions with overlapping breakpoints. This means that if two arrangements, each with an inversion, become the prevalent haplotypes and thereby remove all standard arrangement chromosomes from the population, there is no mutational path to generate a third arrangement. However, a portion of the two-state simulations still had the ancestral, non-inverted arrangement as one of the two haplotypes present, which could have been mutated to a third arrangement to establish the three-arrangement pattern. We are not sure which selective or mutational barriers may prevent transitioning between each state once established. One clear barrier is that it is difficult to generate a less extreme arrangement from a Y-like arrangement, as these can accumulate an indefinite number of mutations since males experience no costs in this set of simulations.

Empirical tests for inversion-associated tradeoffs

A specific motivation for developing and evaluating the model of inversion-associated, balanced sexual antagonism was to potentially better explain the pattern of inversion frequencies observed in D. melanogaster, in which multiple inversions reach some of their highest frequencies in populations from the ancestral range or with similarly tropical/subtropical climates. We therefore set out to test experimentally whether any of four inversions common in a Zambia population of this species show evidence for tradeoffs between male reproductive success and survival (in terms of viability and/or longevity).

We assembled an outbred population with moderate frequencies of each inversion to obtain karyotypically diverse males. We set up four crosses using these males along with females from one of four different inversion-free inbred Zambia lines (Figure 8). Each of these parental crosses occurred in a single large population cage, with populations of at least 600 and equal numbers of each sex (Table S1). We developed a PCR and next-generation sequencing-based assay to estimate inversion-associated SNP frequencies in each pool of males and in each corresponding pool of embryos. These SNPs (given in Table S2) were based on our own analysis of variants that showed perfect association with inversions among 197 Zambia genomes (Table S3; see Materials and Methods). We then implemented a resampling method to test for significant inversion frequency changes between parents and offspring (while accounting for the variance observed in technical replicates; see Materials and Methods), in order to determine whether males carrying each inversion were more or less successful in reproducing than the average male. We also compared frequencies between embryos and aged adults collected 10 weeks after eclosion, to investigate the relationship between inversions and viability/longevity. These aged adults eclosed either in the first or second half of the collection period from the same batch of eggs, thus allowing us to test for inversion associations with development time. We calculated p-values from the resampling tests to assess the significance of each individual change in inversion frequency between relevant samples (Table S4). However, in accord with our primary question regarding fitness tradeoffs, we primarily focus here on joint tests of two opposing frequency changes, as further described below. We could therefore test the hypothesis that a given inversion had opposing influences on survival and male reproduction, which would provide evidence consistent with our model of balanced sexual antagonism, while also examining whether any such tradeoffs depend on maternal genetic background.

The layout and potential expectations of the laboratory evolution experiment.

We cross outbred males from a high-inversion population to inbred, non-inverted females from a specific inbred line, and collect DNA samples from the fathers, embryo offspring, and aged adult offspring for inversion frequency estimation via sequencing (left). We hypothesize from our antagonistic pleiotropy model that one or more of the four inversions present on separate chromosome arms in our experimental populations may experience opposing selection between fathers and embryos versus between embryos and aged adults (middle). To test the significance of observed frequency changes, we designed a resampling model to represent the sources of neutral variation expected in the experiment (right).

When we initially examined the joint effects of male reproduction and survival, we found that in a majority of cases (in 13 out of 16 inversion-line trials; binomial P = 0.02; Figure 9), a given inversion increased in frequency in aged adult offspring compared to the null expectation based on the frequency in fathers. The remaining three cases were all for In(3L)Ok. Ten of the 13 increases were primarily due to increase in frequency between embryo and aged adult samples, suggesting that in terms of viability and/or longevity, inversion heterokaryotypes often had higher fitness than standard homokaryotypes (no offspring were homokaryotypic for an inversion under our crossing design). We note that even standard homokaryotypes among the fathers and offspring should not have broad-scale homozygosity within the inverted region, but should instead be heterozygous for unique standard-arrangement haplotypes from two independent Zambia strains, and that long genomic tracts of identity-by-descent between Zambia strains are quite rare in this large, ancestral-range population (Lack et al. 2016a). Hence, we do not expect that inversion heterokaryotypy is buffering against deleterious effects of inbreeding. Instead, this fitness advantage suggests either an advantage for heterokaryotypy in particular, or perhaps more likely, that multiple inversions harbor alleles that benefit survival or longevity under our experimental conditions (see Discussion).

Some Drosophila inversions show evidence for antagonistic tradeoffs and sex-specific survival.

(A) Testing for antagonistic fitness effects on reproduction and survival (encompassing viability and longevity). Each inversion’s frequency in parents after collection (2 to 3 weeks from eclosion), embryo offspring, and aged adult offspring (10 weeks from eclosion) are shown for each maternal inbred line. Parental frequency is represented instead of paternal frequency for easier visual comparison, and is taken as the paternal frequency divided by two to account for the inversion-free inbred mothers. Adult offspring frequencies are taken from the sum of estimated allele counts across all adult offspring cohorts. * indicates p-value < 0.05 for evidence of significant reversed frequency change from parental to embryo and from embryo to offspring, combined across all four maternal line crosses and corrected for multiple tests, when compared to neutral simulated experiments (details in Materials and Methods). (B) Testing for evidence of sex-specific survival effects. The changes in inversion frequencies between embryos and the aged adult offspring are again illustrated, now separated by their sex. * indicates p-value < 0.05 for significant differences in female versus male inversion frequency among adult offspring, combined across all four maternal crosses and corrected for multiple tests.

When we separated the effects of male reproduction (based on father to embryo changes) and survival (embryos versus adults), we found that the effects of inversions on survival were generally consistent across maternal genotypes. Three inversions were associated with increased survival to 10 weeks: In(3R)K was in all four crosses, while In(2L)t and In(2R)NS each showed elevated survival with all maternal strains except ZI418N (Figure 9A). In contrast, In(3L)Ok was associated with lower survival in each cross (Figure 9A). With regard to male reproduction, the trajectory of inversion frequencies from fathers to embryos to offspring was less consistent between different crosses (Figure 9A), implying that differences in male reproductive success may be dependent on the maternal genotype, which is consistent with past findings in D. melanogaster (Reinhart et al. 2015). Despite such variability, each of the four inversions had multiple frequency changes that departed strongly from null expectations (Figure 9A; Table S4), as might be expected if selection was present but varied in strength and direction between crosses and life stages.

To test our central hypothesis of a tradeoff between male reproductive success and cumulative survival during development and adult aging, we used neutral resampling to test the likelihood of observing the magnitudes of opposing inversion frequency changes (between fathers and embryos, and between embryos and aged adult offspring) under the null hypothesis, integrating across the four maternal crosses and accounting for multiple test correction (Materials and Methods). For In(3R)K, which had particularly consistent changes in frequency across replicate lines (decreasing between parents and embryos, and then increasing between embryos and aged offspring; Figure 9A), this frequency reversal was significantly non-random (p = 2.70 × 10-3; Table S5). In contrast to In(3R)K, In(3L)Ok was associated with (variably) higher male reproductive success and (consistently) lower survival; this reversal was also statistically significant (p = 2.97 × 10-7).

We also tested for frequency differences between aged adult offspring of different sexes and from early-eclosing versus late-eclosing cohorts. All four inversions displayed differences in frequency between female versus male offspring (Figure 9B), consistent with differential effects on sex-specific survival. Strikingly, In(3L)Ok and In(3R)K each displayed sex survival differences in alignment with their antagonistic effects implied above: In(3L)Ok (which favored male reproduction) reduced female survival more strongly than male survival (p = 0.0450), whereas In(3R)K (associated with lower male reproduction) favored the survival of females more than it did males (p = 4.61 × 10-3; Table S5; see Materials and Methods). The other two inversions, In(2L)t and In(2R)NS, each consistently favored female over male survival in all four maternal crosses, although their raw unidirectional cross-combined p values (0.0260 and 0.0126) were not significant after multiple test correction (p = 0.282 and 0.137). In contrast, we found no consistent differences in inversion frequency between early- and late-eclosing offspring (Figure S4).

Estimated selection coefficients for each tested change based on a Wright-Fisher population approximation are available in Table S6; however these may overestimate inversion fitness effects since the survival terms reflect the probability of reaching advanced age. Further, there may be an influence of differential sex-specific mortality during aging leading to greater frequency shifts in females. We collected a total of 1960 females and 3156 males, and if the survivors represent 20% of an original population with an equal sex ratio, then there was 84.7% female mortality and 75.3% male mortality. Additionally, there was some mortality among the parents over the laying period, so some of the frequency change observed in embryos may be attributable to paternal mortality. Paternal mortality was at most 15% among the crosses (Table S1), and the fathers were at most three weeks old, suggesting mild effects on inversion frequency estimation. Further, any positive correlation between the paternal and offspring survival effects of an inversion would bias against detecting the tradeoffs implied above. Overall, we conclude that D. melanogaster inversions have distinct and often context-dependent influences on male reproduction and on survival, with some evidence for antagonistic tradeoffs of the sort proposed in our model.

Discussion

Here, we have introduced a new conceptual model in which (1) sexually antagonistic variation is present and subject to balancing selection that maintains extreme genotypes favoring either male reproduction or survival, and (2) inversions are predicted to facilitate the aggregation of progressively more antagonistically differentiated haplotypes. We have tested this model using simulations, confirming the predicted efficacy of balancing selection and demonstrating that inversions indeed facilitate the construction of haplotypes containing multiple antagonistic variants. Finally, we have deployed Drosophila lab experiments to uncover some evidence for inversion-associated tradeoffs between reproduction and survival, depending on the specific inversion and strains tested.

Inversions in D. melanogaster

We have presented results from a single generation laboratory evolution experiment, in which we crossed outbred male D. melanogaster carrying inversions In(2L)t, In(2R)NS, In(3L)Ok, and In(3R)K to four separate inbred female lines and tracked the frequency of the inversions in the initial males, embryonic offspring, and aged adults. For most inversions, the strength and direction of the tradeoff varied notably between crosses. This might suggest high variance in their phenotypic effects or potentially maternally-dependent variation in male reproductive-success, but is not distinguishable from experimental noise with only one replicate per maternal line. However, greater consistency was observed for In(3R)K, which was associated with lesser male reproductive success but greater survival. A significant tradeoff was also inferred for In(3L)Ok, which favored male reproduction (albeit variably) over survival. These results provide significant but measured evidence for the presence of sexually antagonistic autosomal inversions in this model species, while emphasizing the importance of genetic background.

As indicated in the Introduction, models like ours in which inversion frequencies are influenced by balancing selection on pleiotropic tradeoffs may help explain multiple aspects of observed D. melanogaster inversion frequencies. This class of models inherently accounts for the intermediate maximal frequency of most D. melanogaster inversions in natural populations. To the extent that inversions favor male reproduction, the sexually antagonistic model we propose here could also help account for the predominantly autosomal locations of common inversions in this species, given that male-specific fitness effects have a weaker influence on the frequencies of X-linked variants since the X chromosome spends only one third of its time in males.

This model could also help explain the observed geographic variability of inversion frequencies, as a change in environment may change the fitness value of either pleiotropic effect of the alleles. If a population experiences a novel harsh environment with low survivability, then the population density may decrease, and the number of encountered competing males may decrease as well, reducing the relative fitness advantage of enhanced male display. If there is a genetic tradeoff between mating display and viability, the equilibrium frequency of an antagonistic inversion that favors the genetic extreme of display should be lower. Additionally, in hostile environments, the survival cost of display-favoring variants could be increased, which could further shift their equilibrium frequencies downward or render them not worth carrying at any frequency. These predictions could contribute to the observed negative relationships between some inversion frequencies and latitude or altitude (Kapun et al. 2016; Pool et al. 2017). As a specific example, it is notable that in a high altitude Ethiopian environment with year-round cool weather, which hosts some of this species’ most pronounced phenotypic evolution (Bastide et al. 2014; Lack et al. 2016b), inversions are virtually absent - occurring at the lowest frequencies of any analyzed population (Sprengelmeyer et al. 2020). Concordantly, geographically marginal populations of other Drosophila species are often found to host less inversion polymorphism than more central populations (Krimbas and Powell 1992). Also consistent with tradeoffs involving inversion-associated survival costs, a disadvantage of some D. melanogaster inversions in challenging thermal environments (both cold and hot) was supported by a previous experimental evolution analysis that included In(2L)t, In(2R)NS, and In(3R)K (Kapun et al. 2014), though they did find that In(3R)C consistently increased in frequency in hot conditions and In(3R)Mo in cold.

In contrast, our experiments yielded evidence that In(3R)K may have a tradeoff opposite from our prediction - with the inverted arrangement associated with greater survival but lower male reproductive success. In line with our findings, Said et al. (2018) found that among Zambia strains, In(3R)K was associated with strongly reduced expression of desat2, involved in pheromone production. Whereas, higher desat2 expression – as reported by Said et al. (2018) for standard Zambia karyotypes – has been linked to the “Z-type” males preferred by females of some strains from this region (e.g. Fang et al. 2002; but see Grillet et al. 2012). Conversely, a recent genome-wide association study found In(3R)K heterokaryotypes to be associated with elevated male reproductive success compared to standard homokaryotypes, when males from various North Carolina strains were paired with a Z-type Zimbabwe strain known to exhibit preference for southern African male phenotypes (Yamamoto et al. 2024), underscoring that inversion-related tradeoffs may depend on genetic background and/or experimental context.

Given that our model relies on the ability of either an inverted or a standard karyotype to place males carrying it among the top reproductive competitors, it may seem surprising that we found two different inversions associated with a tradeoff between survival and male reproduction – in that an individual’s karyotype at one inversion could undercut the advantage of their karyotype the other. In our case, it might seem unfavorable for a male carrying In(3L)Ok (associated with increased male reproductive performance but lower survival) to also carry In(3R)K (which showed the opposite patterns). Strikingly then, among 197 sequenced haploid embryo genomes from this Zambia population (each reflecting one female gamete from independent isofemale lines; Lack et al. 2016a), not a single genome carried both of these inversions (even though In(3L)Ok and In(3R)K were carried individually by 36 and 35 of these genomes respectively), representing a significant lack of this hypothetically disadvantageous doubly-inverted karyotype (p = 0.0031, two-tailed Fisher’s Exact Test). These inversions are 20.4 cM apart in terms of meiosis in standard karyotype females (Comeron et al. 2012), but the contributions of natural selection versus potentially altered meiotic dynamics in individuals heterozygous for two inversions in generating this pattern will require further investigation.

More broadly, we observed a trend toward increased inversion frequencies across our full experimental generation. A general benefit for inversions would mirror previous findings in which inversions found at intermediate frequencies in natural D. melanogaster populations (Inoue 1979), as well as inversions from seaweed flies known to be under balancing selection in the wild (Mérot et al. 2020), were both inferred to be under directional selection in the lab. However, one laboratory study did report evidence that a D. melanogaster inversion not tested here, In(3R)P, had frequency-dependent fitness under crowded conditions (Nassar et al. 1973). Furthermore, even an inversion found to be under directional selection in the lab might still be subject to a balanced tradeoff under a more challenging or complex natural environment. Other studies in D. melanogaster, including those on focused seasonal evolution, have found evidence for important tradeoffs between reproduction and robustness to environmental challenges (e.g. Behrman et al. 2015). Hence, it would clearly be desirable to study the effects of inversions on a wider range of potential tradeoffs. For example, Said et al. (2018) suggested that inversions In(2L)t and In(3R)K might be maintained by immunity-related tradeoffs, in light of an over-representation of immune-related genes among differentially expressed transcripts between inverted and standard karyotypes.

In addition to a broader array of balanced tradeoffs that could modulate D. melanogaster inversion frequencies, drift and other forms of selection may contribute as well. Although we emphasize above the potential interplay between ecological adaptation and balanced tradeoffs, some inversions could be impacted by simpler forms of directional selection. However, we must then invoke secondary explanations to account for their limited maximal frequencies (see Introduction). Other forms of balancing selection may also come into play as well. It is clear that some variation in this species is subject to temporally-varying selection, and some inversions have displayed seasonal frequency differences (Dobzhansky 1943; Kapun et al. 2016; Machado et al. 2021; Lange et al. 2022). As with spatial clines, temporal shifts could reflect either simple directional selection or else environmentally-modulated tradeoffs in a model such as ours. Spatial selective pressures may vary both between populations and within population scales.

As indicated in the Introduction, balancing selection on inversions could also be achieved by linkage to recessive deleterious variants, which may be observed as an antagonistic pleiotropic phenotype despite potentially involving strictly deleterious variants in linkage (Pei et al. 2023). In extreme form inversions might be under associative overdominance (Zhao and Charlesworth 2016; Gilbert et al. 2020), where two haplotypes are balanced by recessive cost to both homozygotes. Inversions may be particularly susceptible to carrying recessive deleterious alleles if population size and/or inversion frequency are low, since under these circumstances inversion homozygotes are rarely generated and hence the crossover rate among inverted chromosomes is low, plus there are fewer inverted chromosomes to facilitate rare double recombination events with standard chromosomes.

However, this dynamic is not expected to establish new inversions easily (Charlesworth 2024) and is expected to decay with increased age and population size of the inversion, as mutation, gene conversion, or double crossover resolve the negative linkage between recessive deleterious alleles. Most common inversions in D. melanogaster are both relatively old and have large effective population sizes. Inversions including In(2L)t, In(2R)NS, and In(3R)K have breakpoints estimated to be several times older than the expansion of the species out of southern-central Africa (Corbett-Detig and Hartl 2012; Sprengelmeyer et al. 2020). Those three inversions, plus In(3L)Ok, for which inversion age has not been estimated, all maintain frequencies greater than 0.15 within very large populations (Kapun and Flatt 2019; Sprengelmeyer et al. 2020), and so it is not clear that the persistence of recessive deleterious or lethal alleles in the inverted regions is expected to be high enough that inversions would be maintained by a shared recessive lethal variant preventing fixation. However, sampling of natural populations has demonstrated a somewhat higher frequency of recessive lethal variants in inverted haplotypes, but these are largely unique and heterokaryotypic individuals remain robust, giving low costs to a heavily outbred population (Mukai and Yamaguchi 1974). Many wild-derived inbred strains of D. melanogaster are homokaryotypic for inversions (Lack et al. 2016a). Furthermore, as noted in the Results, an observed lack of inversion homozygotes could reflect strong sexual antagonism as opposed to recessive deleterious variation. Nevertheless, further simulation and experimental testing regarding the potential influence of associative overdominance and other processes on inversion frequencies in D. melanogaster and other species would be desirable.

Generality of Balanced Pleiotropic Inversions

We have proposed that some polymorphic inversions in D. melanogaster, and in other species with analogous mating dynamics, may be maintained in part through balancing selection acting on a frequency-dependent tradeoff. While we explored the tradeoff between male reproduction and viability, we emphasize that any antagonistic pleiotropy involving at least one frequency-dependent trait has the potential to generate similar inversion polymorphism. Such antagonistic pleiotropy may be common in evolution, and might be useful in explaining the many examples of inversions associated with distinct phenotypes.

For the specific model explored here, of antagonistic pleiotropy between survival and display in a mate choice system, the prevalence of the dynamic across species and populations is unclear. It seems plausible that a similar model of antagonistic pleiotropy and balancing selection under mate choice underlies other systems, such as the balanced inversion system in the European ruff Calidris pugnax (formerly Philomachus pugnax), where three inverted haplotypes on an autosome determine aggressive, cooperative, and female-mimic male lekking morphs, albeit with a dominance relationship and homozygous inviability for an inverted haplotype (Küpper et al. 2016). A recent study on the female-mimic Faeder morph suggests that it may be maintained in part by a balance between its effects lowering female reproductive output while increasing male mating success, and the system is likely more complex as a whole due to the three interacting morphs (Giraldo-Deck et al. 2022). More generally, a number of factors contribute to the potential prevalence of this dynamic. Mate competition and choice during reproduction is rather common among animals, and this may mean sexually antagonistic inversions are plausibly also common.

The model depends on the distribution of fitness effects of new mutations and the availability of sexually antagonistic variation, but such variation does seem prevalent (Zajitschek and Connallon 2018; Ruzicka et al. 2019). More likely to limit the prevalence of such inversions are the meiotic costs experienced by inversion heterokaryotypes in some systems (White 1973), whether sexually antagonistic variants are clustered in a region that would allow inversions to increase linkage, and whether other mechanisms like sex-biased expression (Ingleby et al. 2015), gene duplication (Connallon and Clark 2011), and linkage to the sex chromosome or sex-determining locus would resolve the antagonism more frequently (though see following subsection on sex chromosome evolution).

Perhaps more importantly, there are a number of well-studied cases of inversion polymorphism that might potentially follow this dynamic of balancing selection between haplotypes simultaneously demonstrating antagonistic pleiotropy and trait-specific frequency dependence, but which do not involve sexual antagonism. An inversion polymorphism in the seaweed fly Coelopa frigida has been found to trade off between larval survival and adult reproduction (Mérot et al. 2020). Papilio butterfly populations harbor well-studied mimicry polymorphisms that segregate between inverted segments and may often experience frequency-dependent selection (Clarke and Sheppard 1960; Kunte et al. 2014; Poul et al. 2014). The snail Cepaea nemoralis harbors diverse and distinct shell patterning haplotypes associated with inversions that remain balanced within single sampling locations (Cook 2013; Surmacki et al. 2013). Several inversions in different ant species segregate a single-queen colony strategy from a cooperative multi-queen strategy within the same range (Wang et al. 2013; Libbrecht and Kronauer 2014). Some vertebrates found to harbor antagonistically pleiotropic inversions include the rainbow trout (Pearse et al. 2019) and the zebra finch (Pei et al. 2023). These balanced polymorphisms are variously hypothesized to involve maintenance by overdominance (Mérot et al. 2020), linked deleterious recessives (Pei et al. 2023), or reversals of dominance (Pearse et al. 2019), but these mechanisms are not mutually exclusive with each other, or with the model presented here.

Essentially, any population with antagonistic pleiotropy involving some form of negative frequency-dependent balancing selection on one of the phenotypes might exhibit the same dynamic. The population would experience similar divergent selection towards alternate phenotypic extremes, with the frequency-dependence maintaining the presence of two genotypes. Importantly, this model involves a different mechanism for the maintenance of antagonistically pleiotropic variation than models based on reversals in dominance, where dominance effects generate a net heterozygote advantage at a locus (Connallon and Chenoweth 2019). The model proposed here integrates well with existing thought on the evolution of divergent selection on balanced systems. For example, Kopp and Hermisson (2006) have proposed a specific form of frequency dependent divergent selection (FDDS), involving competitive exclusion between similar genotypes for a trait otherwise under stabilizing selection, and this model was expected to result in single or few loci of large effect. Their model did not consider pleiotropy or linkage modifiers, but still had comparable dynamics in that selection in the FDDS regime favored small numbers of large effect loci, while the inversion haplotypes in the model presented here simplify the recombination architecture to approximate a single locus. This simplification of the genetic architecture follows the reduction principle in recombination modifier theory (Feldman 1972; Feldman and Balkau 1973; Feldman et al. 1980; Altenberg and Feldman 1987; Altenberg et al. 2017). When a population experiences selection towards two (or more) divergent optima, under a selection regime that provides a balancing mechanism to maintain both, indirect selection will favor a reduction of the genetic architectures to prevent the generation of intermediate phenotypes. The likelihood of either model likely rests in part upon the inherent polygenicity of the trait under selection, with simpler traits being more likely to resolve by single locus FDDS dynamics than by inversion association (e.g. Yassin et al. 2016).

Notably, our simulations tended to result in strong sexual antagonism, such that it leads to a substantial fraction of the population failing to survive to reproductive age. It is unclear how common such extreme antagonism may be in natural populations. If it is uncommon in nature, even in taxa that experience high levels of mate competition as modeled here, then it is unclear what prevents our simulation predictions from being realized. It may be that genetic constraints prevent the occurrence of highly antagonistic genotypes, or that the relationship between survival-altering mutations and fitness is different than we have modeled. Or, populations may steadily evolve to mitigate antagonistic effects that arise and persist through processes such as those we modeled. Alternatively, some highly antagonistic haplotypes may become associated with sex determination, as suggested below.

Antagonistic autosomal inversions may precede novel sex chromosomes

If balanced, sexually antagonistic, autosomal inversions occur frequently enough, they may also contribute to the evolution of sex chromosomes. Given an established inversion with significant sexual antagonism, the inverted haplotype that favors one sex still experiences an antagonistic cost of transmission to offspring of the opposite sex.

Mutations that link the haplotype with the sex it benefits are therefore favored under selection, and this can occur either by fusing to an existing sex chromosome (Charlesworth and Charlesworth 1980), though this only fully resolves the antagonism for linkage to Y and W chromosomes, or by linkage to a new sex determining mutation (van Doorn and Kirkpatrick 2007).

Here, we discovered that strong inversion-linked antagonism could yield a scenario where essentially all highly-successful males were homokaryotypic for a display-favoring karyotype, whereas reproducing females were overwhelmingly heterokaryotypic (Figure 5). This situation is similar to the ZW sex chromosome system, suggesting that this particular pleiotropic trade-off may favor resolution by linkage to a female sex determination factor. Further, simulations in which only females experienced antagonistic survival cost usually generated populations with a chromosome essentially lethal in females, thus exclusively transmitted in males similar to a Y chromosome (Figure 7). These simulations sometimes had three states that appeared stable at 20N generations, including Y chromosome-like, W chromosome-like, and shared intermediate arrangements. This three-arrangement stable polymorphism seems unusual for a natural population, but might be interesting to explore in populations with polymorphic sex determination.

There are often deleterious effects of chromosome fusion or novel sex determination (Pennell et al. 2015; Saunders et al. 2019), which may represent a general obstacle to sex chromosome evolution. However, balanced sexually antagonistic inversions may concentrate enough antagonism that the built-in benefit of resolving it may sometimes overcome the fitness costs of evolving new sex chromosomes (Blackmon and Brandvain 2017). Hence, antagonistic inversions might be more likely than other loci to be involved in the evolution of new sex chromosomes.

A basic prediction of the above hypothesis is that at least one inversion difference between sex chromosomes should already exist at the time of new sex chromosome formation, and genomic differentiation therefore begins between inversion karyotypes on an autosome even before sex-specific transmission of the region. This scenario contrasts with current models of sex chromosome evolution in which inversion accumulation and genomic differentiation accrue only after sex chromosome formation, due to advantages in suppressing recombination involving sexual antagonistic variants already associated with a sex-determining locus (Wright et al. 2016). In this context, we note that the neo-X and neo-Y homologs of Muller element C in Drosophila miranda, which became sex-linked only about 1.5 million years ago, already appear to have inversion differences between them (Chen et al. 2024 – Figure 5A). Further, a recent study in Littorina saxatalis snails has discovered an ecotype-specific sex determination system that may be founded on such tradeoffs, though selection across ecotypes involves discrete geographic boundaries and is only partly comparable (Hearn et al. 2022). Still, additional genomic studies of new sex chromosomes are needed to assess the potential role of sexually antagonistic inversions in sex chromosome evolution.

Summary and Future Prospects

We have advanced a model in which inversion polymorphism is maintained by the frequency-dependent fitness effects of a pleiotropic tradeoff associated with inversion-linked genetic variation. We have used sexual antagonism associated with increased male reproduction versus survival as a specific motivating case, and we explore the predictions of this model using a novel simulation algorithm. In our simulations, antagonistic variants could be maintained at an equilibrium frequency, and inversions enabled multiple antagonistic variants to persist and form more strongly antagonistic haplotypes. We complemented our conceptual and computational modeling with laboratory experiments designed to detect fitness effects of four inversions common in an ancestral range population of D. melanogaster. We indeed found some evidence of tradeoffs between male reproduction and survival for these inversions, with two out of four inversions showing significant evidence for such a tradeoff, although we also observed a strong dependence of male reproductive success on female genetic background.

In light of the persistent mystery of how D. melanogaster inversion frequencies are determined, and how readily we detected tradeoffs involving male reproduction and survival, we suggest that further studies investigating the potential fitness tradeoffs of inversions in this model system are strongly warranted. While some such studies have previously been conducted, the applicability of our amplicon sequencing approach to estimate inversion frequencies in large pools of flies may improve the sensitivity of such experiments. Our understanding of the phenotypic impacts of D. melanogaster inversions would benefit from investigations of a wider range of fitness-related traits that might be involved in balanced tradeoffs, including female fecundity and immunity. It will also be important to verify the potential frequency dependence and phenotypic consequences of D. melanogaster inversions under a wider range of conditions, including challenging environments such as cool temperatures and high population densities. While inherently challenging, a further goal would be to identify loci within an inversion such as In(3L)Ok or In(3R)P that contribute to observed tradeoffs, and in the case of an arrangement favoring male reproduction, to test for the predicted synergistic epistasis (for male reproductive fitness) among linked variants.

Clearly, it will also be important to test the generality of our model by estimating the fitness associations of inversions in a wider range of species. And particularly in the case of species with very recent changes sex chromosomes, it will be of great interest to test whether one or more inversions are already present between these new sex chromosomes, as predicted a model in which sex chromosome evolution involves the resolution of inversion-associated sexual antagonism. Further analytic and simulation work will also be important in clarifying the range of potential scenarios in which balancing selection and antagonistic pleiotropy may lead to the maintenance of polymorphic haplotypes involving inversions (or parts of the genome that rarely recombine), including studies that consider the effects of dominance and that consider tradeoffs other than the one examined here. Collectively, such studies will advance our broader understanding of the roles of balancing selection, pleiotropic tradeoffs, and genome structure in shaping genetic variation and genomic evolution.

Materials and Methods

Simulations

We developed a new individual-level forward simulator (‘SAIsim’ for Sexually Antagonistic Inversion simulator) written in python 3.7 for the simulation of inversions and sexually antagonistic alleles at infinite loci, with uniform rates of mutation, crossing over, and gene conversion (per bp, per generation) over a number of independent chromosome arms. The dynamic we were interested in modeling requires direct accounting of pleiotropy between different components of fitness, crossing over and gene conversion within inversions (as well as the possibility of inversion fixation), and directly modeling male display and female choice, and no existing simulator provided this functionality. These simulations were used to establish the potential for the model to occur in a population generally, and not to compare against specific empirical observations. We have used literature estimates of mutation rate and conversion rate of Zambian D. melanogaster for these simulations (Comeron et al. 2012; Huang et al. 2016) but have no information on the joint distribution of effects of mutations on our model of survival and display in D. melanogaster. So, we sampled from relatively broad but arbitrary combinations of mutation effects, only assuming that there is likely some degree of antagonistic pleiotropy naturally present in genomic variation.

In SAIsim, a population is instantiated as a python object, and populated with individuals which are also represented by python objects. These individuals may be instantiated using genomes specified by the user, or by default carry no genomic variation. In each generation, as diagrammed in Figure 2, the simulator calculates a survival probability for each individual as the product of the survival values of each variant they carry, which range [0,1]. Here, multiplicative selection for survival prevents biologically unreasonable negative survival probabilities from occurring. Male and female survival costs can each be included in a sex-specific manner, but only globally in the simulation, not per mutation. For each offspring in the next generation, a mother is selected randomly, weighted by individual survival probabilities, with replacement. To model the encountered males that might become the father of an offspring individual from this female, a set number of males (i.e. the uniformly specified encounter number, here 100 males unless otherwise specified) is randomly sampled from the full population of males, weighted by their survival probabilities. The males each have a genetic display quality given as the sum of the reproductive values of their alleles, each of which range [0,1]. A noise parameter, normally distributed around zero with a standard deviation of one, is added to each male display value during each encounter. The male with the highest noise-adjusted display quality among the encountered males is chosen to be the father. As offspring are generated, crossover locations are sampled, and resampled in cases where the resultant gamete would be aneuploid. While the simulator can allow recombination in both sexes, all simulations presented only generate crossovers and gene conversion events for female gametes, in accordance with the biology of D. melanogaster.

Resampling aneuploids removes the fecundity cost that may exist in some taxa, but was chosen to reflect the lack of observed fecundity effects for most paracentric inversions of D. melanogaster. Gene conversion is modeled as a constant probability per heterozygous variant, with each event independently affecting just one locus, rather than explicitly modeling a per-bp conversion rate and tract length (given that gene conversion tracts are generally short and each would typically contain only a single modeled variant in our simulations). The dynamics of gene conversion are expected to have separate effects on the assembly and the maintenance of arrangements. The assembly of an arrangement is positively dependent on the population-scaled rate of recombination of alleles as new polymorphisms start at low frequency and often not on their optimal haplotype. However, the indirect fitness effect of the inversion is dependent on its effect on recombinant offspring, and therefore negatively depends on the per-individual rate of conversion. Finally, inversions and new mutations are potentially added to the resultant gamete, with inversions resampled when overlapping the coordinates of an inversion already present on the same individual chromosome. SAIsim also allows the specification of mutation rates and the distribution of effect sizes of new mutations.

The simulations presented in Figures 5, 6, and 7 rescale parameter values by a factor of 2,000 to emulate an estimated ancestral Ne of roughly 2 × 106 (Sprengelmeyer et al. 2020) while using a smaller simulated population of 1,000 individuals. The low crossover rate simulations model a 0.0218 cM region and the high crossover rate simulations model 2.18 cM, which scale to 43.6 cM and 43.6 M female meiotic maps in the simulated populations. Based on an average autosomal recombination rate in females of 2.18 × 10-8 for D. melanogaster (Comeron et al. 2012), these represent 10kb and 1Mb regions. Gene conversion occurs at each heterozygous variant with a probability of 0.1295, based on scaling γ=1.25 × 10-7 events/bp/female meiosis with a conversion track length of 518 bp (Comeron et al. 2012) to a per-single-nucleotide-variant estimate of 6.475 × 10-5 conversions/bp/female meiosis.

Rates of sexually antagonistic mutations and inversion mutations are empirically undetermined. Here, sexually antagonistic mutations were modeled at a rate corresponding to one per thousand mutations, and so based on an estimate of 5.21 × 10-9 mutations/bp/gen on autosomes (Huang et al. 2016), and modeling 10kb and 1Mb regions, these were scaled to 1.042 × 10-4 and 1.042 × 10-2 events per gamete respectively. Inversions were approximated as arising once per hundred sexually antagonistic mutations, i.e. 1.042 × 10-6 and 1.042 × 10-4 events per gamete. The simulations use mutations with uniformly distributed quality values in the range [0,1]; although distributions skewed toward small effects may be more realistic, our uniform approach avoids the need to invoke a specific but empirically unknown distribution.

SAIsim’s object-oriented design allows specification of a population model by a new user familiar with the basics of python scripting. A population object can be parameterized and populated in a few lines of python script, and then parameters can be changed between periods of simulation, allowing bottlenecks and demographic changes as well as changes in mutational parameters between generations. Replicate SAIsim simulations were performed via the UW-Madison Center For High Throughput Computing (CHTC), which manages cycle servers using the HTCondor distributed computing project (Thain et al. 2005; Erlandson and Theisen 2018).

Experimental Populations and Sequencing

To act as a high-inversion source population for the potential fathers in our experiment, we combined 15 males and 15 females from each of 10 inbred strains derived from wild flies collected in Zambia, and allowed the population to breed for two generations. These 10 strains were chosen to collectively carry the following common inversions: In(2L)t, In(2R)NS, In(3L)Ok, In(3R)K. This high-inversion population may not and need not have been at an equilibrium frequency for any particular inversion; instead it was simply designed to transmit each inversion to offspring at meaningful frequency. We collected males from this second generation to use as the paternal pool in each of our independent maternal line experiments, and these males were expected to carry a wide range of inversion genotypes. Each batch of high-inversion males was crossed to an equal number of virgin females from a single inbred Zambian strain (Figure 8; Table S1). Four such experiments were performed, each using one inversion-free maternal strain, in clear plexiglass cages (19.3 by 19.3 by 30.0 cm) with a loose mesh covering on one end for access by hand.

Embryos were collected by placing four 7.9 cM petri dishes of grape agar in a population cage for one day, then counting and freezing the collected eggs. To obtain adult offspring, six bottles of standard Drosophila medium (see below) were placed in a population cage for one day. Egg-laying for embryo and adult collections occurred on alternating days over two weeks, after which all surviving parents were counted and frozen. From the first day in which flies eclosed in a bottle, adults were collected at the end of each period of a day to two days and were separated by sex. These cohorts were aged by weekly transfer to new bottles. The adult offspring were therefore initially separated by sex and by the time after laying at which they eclosed, as well as the time in which they were laid, with different early- or late-eclosing batches from a given cross later pooled before DNA extraction. In our husbandry conditions, the earliest eclosion was 11 days after laying, but this was rare and a first eclosion at 12 days was more common among the bottles. Most flies took several days longer to develop, and we considered flies collected in the first 6 days of a 10 day collection window to be the early-eclosing pool, in order to split the pool sample sizes closer to evenly. The collected adult flies were then aged by sequentially transferring them between bottles weekly for approximately 2.5 months, and then the adult offspring were sexed, counted, and collected. A 10-week aging window was chosen based on the time taken for a control set of four bottles of flies from the paternal outbred population to reach 80% mortality, to reflect the survival and longevity stressors of the laboratory bottle environment. All adult and embryo samples were frozen at −80°C before DNA extraction.

Since no inversions could be inherited through the mothers, inversion frequencies among successful male gametes could be inferred from their pooled offspring. Therefore, the above sampling scheme allowed us to identify inversion frequencies in the set of potential parents, in embryonic offspring, and in older adults, and so to assess potentially selected changes in inversion frequency across male reproduction or between eggs and aged adults (encompassing viability plus longevity). We chose to age the populations to include longevity as long term survival, which both encompasses fitness challenges that generate mortality at earlier time points, and is relevant to the need to survive during periods of unsuitable environments. Although a natural population from North Carolina was estimated to have an average generation time of about 24 days (including developmental time; Pool 2015), in many climates fly populations cycle seasonally between periods of rapid reproduction and survival for months as adults under reproductive diapause (Saunders et al. 1990; Ragland et al. 2019). While longevity is a key selective pressure underlying overwintering, the relationship between longevity in permissive lab conditions without diapause and in natural conditions under diapause is unclear (Schmidt et al. 2005; Flatt 2020), and our experiment represents just one of many possible ways to examine tradeoffs involving survival.

Unless otherwise noted, all flies in the above experiments were kept in a lab space of 23°C with around a degree of temperature fluctuation and without an artificially controlled day/night cycle. Light exposure was dependent on the varying use of the space by laboratory workers but amounted to near constant exposure to at least a minimal level of lighting, with some variable light due to indirect lighting from adjacent rooms with exterior windows. All fly food used was prepared in batches of 4.5 L water, 500 mL cornmeal, 500 mL molasses, 200 mL yeast, 54 g agar, 20 mL propionic acid, and 45 mL tegosept 10% (in 95% ethanol). The grape agar was prepared in batches of 500 mL from FlyStuff.com grape agar powder, and wet Red Star yeast was brushed over the agar plates before use to encourage laying.

Sequencing and Data Preparation

DNA for each replicate age group was extracted by a phenol-chloroform extraction protocol used previously in preparing DNA for the Drosophila Genome Nexus and Drosophila Population Genomics Project. We followed protocol ‘BPC’ used in DPGP phase 1 (presented in Langley et al. 2012), except with the addition of proteinase-K during homogenization of adult flies. These protocols are ultimately based on the common quick preparation protocol presented in Drosophila Protocols (Huang et al. 2009). Inversion genotyping was performed by amplicon sequencing, where amplicons associated with each inversion were designed to contain at least one single nucleotide polymorphism in the inversion interior that is fixed for alternate alleles between the inversion karyotypes, while still sharing conserved primer regions across karyotypes to prevent primer bias. Genetic variation in these inversions was assessed using 197 previously sequenced and inversion-called haploid genomes from the same Zambian D. melanogaster population studied here, available from the Drosophila Genome Nexus (Lack et al. 2016a). The amplicon region identification and primer selection was performed with a custom script, which calls Primer3 version 2.4.0 for primer candidate generation (Koressaar and Remm 2007; Untergasser et al. 2012; Koressaar et al. 2018). Primers thus selected were ordered from IDT as oligos ligated to Illumina stubby adaptors (Table S2). Illumina sequencing libraries were then prepared for each sample following the Illumina metagenomic amplicon sequencing protocol (Illumina 2013), with 50 ng/uL diluted DNA samples. and sequenced using 150bp paired end reads on an Illumina NovaSeq to at least a depth of 20,000 reads. Depth was selected to ensure high likelihood of detecting a 0.025 change in inversion frequency from an initial frequency of 0.10, based on simulated resampling of the collected flies, DNA extraction, and sequencing.

Sequence preparation and analysis were performed using custom pipeline scripts. Read trimming, alignment, and QC were performed using bwa version 0.7.17-r1188 (Li 2013), GATK version 3.4-46 (Van der Auwera and O’Connor 2020), and Picard version 1.79 (Broad Institute 2019). Reads were trimmed at the ends to remove segments with base quality less than 20, and read pairs were filtered to retain only those pairs where both reads aligned with mapping quality >20. SNP identity at the sites of interest was called from those remaining reads with base quality >20 at the SNP of interest. Due to the presence of chimeric reads in the amplicon library sequencing, we chose to use a single discriminant SNP to determine each inversion’s frequency in the read pool, instead of incorporating haplotype or additional SNP information.

Statistical Analysis of Inversion Frequencies

Before analyzing the significance of frequency change reversals across three cohorts (which is described below), we assessed the significance of the difference in inversion frequency between just two sampled life stages in the same experiment. For each such comparison, we first calculated a chi-square test statistic (as a convenient metric to use when comparing empirical data with simulated experimental replicates, rather than as a formal test) from the two-by-two table of inverted and non-inverted allele counts at two life stages. Allele counts were estimated by multiplying the inversion-specific SNP allele frequency in the sequencing reads by the allelic sample size based on the number of flies present in the sample. We then generated an approximate distribution of the chi-square statistic when selection is absent (our null hypothesis), using a resampling approach (Figure 8), as described in the following paragraph. Comparing the observed statistic to our modeled neutral distribution of test-statistic values allowed us to calculate a p-value for this inversion’s frequency change between these two life stages.

To generate the neutral distribution of the chi-square statistic, we took the observed paternal inversion frequencies and then sampled from them to obtain new paternal and offspring read counts to represent a single neutral data set as follows. First, we resampled the collected allelic counts (given in Tables S7 and S8) to account for sampling variance between replicates. For the paternal samples, we needed to account for pre-sequencing mortality as up to 15% of males died before collection (Table S1). So, in generating a new paternal read set, we first modeled down-sampling of the paternal allele sample from the initial census allelic count to the collected allelic count as a hypergeometric sampling. For all other cohorts we used a binomial resampling from the collected allelic count. Second, we used a binomial sampling step to account for the experimental variance introduced by the extraction and library preparation. The extraction and library preparation appeared to introduce variation independent of allele sample size, possibly due to variation introduced during PCR or sample homogenization. This inference was based on analysis of data from control library preparations (Table S9) and from sequencing done in parallel with our experiment from sample fly pools with known inversion frequencies. These control samples included an average of 6 technical replicates (separate library prep and sequencing replicates from the same flies) from four different expected frequencies for In(2L)t, as well as one expected frequency for each of In(2R)NS and In(3R)K. Due to genotyping and husbandry difficulties we did not have a control set for In(3L)Ok. Based on the inversion frequency differences observed among these technical replicates, we estimated that taking a binomial sampling of size 316 from the sample allele frequency reflected a maximum likelihood model for the sample-size-independent variation introduced by the library prep and sequencing process. Third, from the preliminary resampled paternal inversion frequency, we then took a binomial sample of size equal to the read count to account for the random sample of sequencing reads, thus yielding resampled paternal read counts (given in Tables S7 and S8) that accounted for the number of sampled and non-sampled individuals, experimental variance, and depth of coverage. Fourth, to generate resampled offspring read sets, we instead used a binomial sampling of half the offspring alleles from the fully resampled paternal allele frequency for each separate offspring cohort, since the homozygous mothers all contributed non-inverted alleles. Finally, each offspring allele sample was resampled using the same size 316 binomial sampling to account for the variation introduced during this library’s preparation, and binomial sampled again with the read count size to account for sampling reads from the library, thus yielding resampled offspring read counts that accounted for the same three factors as for paternal counts.

For a pairwise comparison of inversion frequency between two samples, we calculated the same allele count chi-square statistic as described above for the empirical data for each resampled set of allele counts, generating the full distribution of neutral chi-square test statistics from which we could identify extreme value cutoffs. We then obtained a p-value from the proportion of resampled replicates in which the chi-square statistic was greater than or equal to the empirical value. If the test was directional, we took the proportion of resampled replicates with both a chi-square statistic greater than or equal to the empirical value and a frequency difference in the expected direction.

In order to identify a frequency change reversal between life history stages that could reflect a fitness tradeoff, we tested for the joint significance of two opposing frequency changes using three cohorts: specifically a frequency change in one direction between paternal and embryo samples, followed by a frequency change in the opposite direction between embryo and adult offspring samples. These changes both depend on the shared embryonic cohort in the test, and so are not independent. Therefore, we evaluated the two inversion frequency changes jointly from a set of resampled neutral data sets in which the three cohorts arose from the same resampling process. Here we considered how likely it was that a sample of the three cohorts from the associated estimated neutral model had chi-square values for both the comparison of the fathers to embryos and the embryos to adult offspring that were each equivalent or more extreme in the test direction than the observed data. For example, when testing the significance of an observed paternal-embryo increase and then embryo-adult decrease in frequency, our p-value would be defined by the proportion of resampled replicates in which the frequency changes were in the tested directions of increase then decrease, and their chi-square values were both more extreme than the empirical values. The potential for empirical frequency change reversals to occur in either direction was accounted for at the multiple test correction step as indicated below.

To assess the significance of inversion frequency changes considered across all four maternal line crosses at once, we used Fisher’s combined probability test. We applied this to both particular three-cohort frequency change reversals and to the set of two-cohort frequency changes. The inversion trajectory is independent between maternal crosses, but the differences between each cross in maternal line and initial paternal genetic variation make the replicates difficult to compare in a more structured way, so we felt this was an appropriate test for combined significance. In assessing our tradeoff hypothesis, we applied this test only for inversions where it was well-motivated; i.e. where the averaged frequency change across maternal crosses was in opposite directions between paternal/embryo and embryo/adult comparisons.

To account for multiple hypothesis testing in our cohort comparisons, we used a Benjamini-Yekutieli false discovery rate correction for significance. We applied this correction across the set of Fisher’s method combined p-values we generated to assess the relevant frequency change reversal hypotheses. So, this corrected for the multiple hypotheses of two potential tradeoff directions and the four inversions studied, from p-values that already combined evidence from different maternal line crosses. Benjamini-Yekutieli was chosen due to its robustness to the unclear dependence relationships between some of the tests involved. For example, multiple tests of frequency changes from a single cohort (e.g. an embryo sample compared against both paternal and aged adult samples) are likely to be positively correlated.